---
title: "PBMC DEG and GSEA Summary"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(readr)
library(glue)
library(DT)
```

# Overview

This report summarizes DEG and GSEA results from two different analyses:

* **paired**: before vs after treatment
* **pooled**: Severe vs Healthy vs Mild

These analyses answer different biological questions, so they are presented separately.

The combined files contain metadata columns including:

* `cluster`
* `cluster_source`
* `source_file`

These are used to organize tables and plots.

# Setup

```{r}
base_dir <- "PBMC_report_outputs"
cluster_source_name <- "Monaco_fine"

deg_file <- glue("{cluster_source_name}__ALL_clusters_limma.csv")
gsea_go_file <- glue("{cluster_source_name}__ALL_clusters_GSEA_GO.csv")
gsea_reactome_file <- glue("{cluster_source_name}__ALL_clusters_GSEA_REACTOME.csv")
```

```{r}
dt_opts_10 <- list(
  pageLength = 10,
  scrollX = TRUE,
  autoWidth = TRUE,
  dom = "Bfrtip",
  buttons = c("copy", "csv", "excel")
)

dt_opts_25 <- list(
  pageLength = 25,
  scrollX = TRUE,
  autoWidth = TRUE,
  dom = "Bfrtip",
  buttons = c("copy", "csv", "excel")
)

make_dt <- function(df, caption_text, page_len = 10) {
  opts <- if (page_len == 25) dt_opts_25 else dt_opts_10
  
  datatable(
    df,
    caption = caption_text,
    filter = "top",
    rownames = FALSE,
    extensions = "Buttons",
    options = opts
  )
}
```

```{r}
read_deg <- function(type) {
  f <- file.path(base_dir, type, cluster_source_name, "deg_results", deg_file)
  if (!file.exists(f)) return(NULL)

  read_csv(f, show_col_types = FALSE) %>%
    mutate(
      analysis = type,
      sig = !is.na(adj.P.Val) & adj.P.Val < 0.05,
      direction = case_when(
        logFC > 0 ~ "Up",
        logFC < 0 ~ "Down",
        TRUE ~ "Zero"
      )
    )
}

read_gsea <- function(type, geneset_name, file_name) {
  f <- file.path(base_dir, type, cluster_source_name, "gsea_results", file_name)
  if (!file.exists(f)) return(NULL)

  read_csv(f, show_col_types = FALSE) %>%
    mutate(
      analysis = type,
      sig = !is.na(p.adjust) & p.adjust < 0.05,
      direction = case_when(
        NES > 0 ~ "Positive",
        NES < 0 ~ "Negative",
        TRUE ~ "Zero"
      ),
      geneset = geneset_name
    )
}

deg_paired <- read_deg("paired_cross_pool")
deg_pooled <- read_deg("pooled")

gsea_paired <- bind_rows(
  read_gsea("paired_cross_pool", "GO", gsea_go_file),
  read_gsea("paired_cross_pool", "REACTOME", gsea_reactome_file)
)

gsea_pooled <- bind_rows(
  read_gsea("pooled", "GO", gsea_go_file),
  read_gsea("pooled", "REACTOME", gsea_reactome_file)
)
```

# Paired Analysis

## DEG summary

```{r}
paired_deg_summary <- deg_paired %>%
  group_by(cluster_source, cluster, source_file) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    up = sum(sig & logFC > 0, na.rm = TRUE),
    down = sum(sig & logFC < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cluster, source_file)

make_dt(
  paired_deg_summary,
  "Paired DEG summary by cluster and source_file",
  page_len = 10
)
```

## DEG counts by cluster

```{r}
paired_deg_cluster <- deg_paired %>%
  group_by(cluster_source, cluster) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    up = sum(sig & logFC > 0, na.rm = TRUE),
    down = sum(sig & logFC < 0, na.rm = TRUE),
    n_source_files = n_distinct(source_file),
    .groups = "drop"
  ) %>%
  arrange(desc(significant))

make_dt(
  paired_deg_cluster,
  "Paired DEG counts by cluster",
  page_len = 10
)
```

## Plot: paired DEG by cluster

```{r}
paired_deg_cluster %>%
  ggplot(aes(x = reorder(cluster, significant), y = significant)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Paired analysis: significant DEG by cluster",
    x = "Cluster",
    y = "Significant genes"
  )
```

## Plot: paired DEG by source_file

```{r}
paired_deg_summary %>%
  ggplot(aes(x = reorder(source_file, significant), y = significant, fill = cluster)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Paired analysis: significant DEG by source file",
    x = "Source file",
    y = "Significant genes"
  )
```

## Top paired DEG hits

```{r}
top_paired_deg <- deg_paired %>%
  filter(sig) %>%
  arrange(adj.P.Val, desc(abs(logFC))) %>%
  select(cluster, cluster_source, source_file, gene, logFC, adj.P.Val, P.Value, t, B) %>%
  slice_head(n = 50)

datatable(
  top_paired_deg,
  caption = "Top paired DEG hits",
  filter = "top",
  rownames = FALSE,
  extensions = "Buttons",
  options = dt_opts_25
) %>%
  formatStyle(
    "logFC",
    color = styleInterval(0, c("blue", "red"))
  ) %>%
  formatSignif(columns = c("logFC", "adj.P.Val", "P.Value", "t", "B"), digits = 3)
```

## GSEA summary

```{r}
paired_gsea_summary <- gsea_paired %>%
  group_by(geneset, cluster_source, cluster, source_file) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    positive = sum(sig & NES > 0, na.rm = TRUE),
    negative = sum(sig & NES < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(geneset, cluster, source_file)

make_dt(
  paired_gsea_summary,
  "Paired GSEA summary by cluster and source_file",
  page_len = 10
)
```

## GSEA counts by cluster

```{r}
paired_gsea_cluster <- gsea_paired %>%
  group_by(geneset, cluster_source, cluster) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    positive = sum(sig & NES > 0, na.rm = TRUE),
    negative = sum(sig & NES < 0, na.rm = TRUE),
    n_source_files = n_distinct(source_file),
    .groups = "drop"
  ) %>%
  arrange(geneset, desc(significant))

make_dt(
  paired_gsea_cluster,
  "Paired GSEA counts by cluster",
  page_len = 10
)
```

## Plot: paired GSEA by cluster

```{r}
paired_gsea_cluster %>%
  ggplot(aes(x = reorder(cluster, significant), y = significant)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  facet_wrap(~geneset, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Paired analysis: significant GSEA pathways by cluster",
    x = "Cluster",
    y = "Significant pathways"
  )
```

## Top paired GSEA hits

```{r}
top_paired_gsea <- gsea_paired %>%
  filter(sig) %>%
  arrange(geneset, p.adjust, desc(abs(NES))) %>%
  select(geneset, cluster, cluster_source, source_file, ID, Description, NES, p.adjust, pvalue) %>%
  group_by(geneset) %>%
  slice_head(n = 30) %>%
  ungroup()

datatable(
  top_paired_gsea,
  caption = "Top paired GSEA hits",
  filter = "top",
  rownames = FALSE,
  extensions = "Buttons",
  options = dt_opts_25
) %>%
  formatStyle(
    "NES",
    color = styleInterval(0, c("blue", "red"))
  ) %>%
  formatSignif(columns = c("NES", "p.adjust", "pvalue"), digits = 3)
```

# Pooled Analysis

## DEG summary

```{r}
pooled_deg_summary <- deg_pooled %>%
  group_by(cluster_source, cluster, source_file) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    up = sum(sig & logFC > 0, na.rm = TRUE),
    down = sum(sig & logFC < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cluster, source_file)

make_dt(
  pooled_deg_summary,
  "Pooled DEG summary by cluster and source_file",
  page_len = 10
)
```

## DEG counts by cluster

```{r}
pooled_deg_cluster <- deg_pooled %>%
  group_by(cluster_source, cluster) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    up = sum(sig & logFC > 0, na.rm = TRUE),
    down = sum(sig & logFC < 0, na.rm = TRUE),
    n_source_files = n_distinct(source_file),
    .groups = "drop"
  ) %>%
  arrange(desc(significant))

make_dt(
  pooled_deg_cluster,
  "Pooled DEG counts by cluster",
  page_len = 10
)
```

## Plot: pooled DEG by cluster

```{r}
pooled_deg_cluster %>%
  ggplot(aes(x = reorder(cluster, significant), y = significant)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Pooled analysis: significant DEG by cluster",
    x = "Cluster",
    y = "Significant genes"
  )
```

## Plot: pooled DEG by source_file

```{r}
pooled_deg_summary %>%
  ggplot(aes(x = reorder(source_file, significant), y = significant, fill = cluster)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Pooled analysis: significant DEG by source file",
    x = "Source file",
    y = "Significant genes"
  )
```

## Top pooled DEG hits

```{r}
top_pooled_deg <- deg_pooled %>%
  filter(sig) %>%
  arrange(adj.P.Val, desc(abs(logFC))) %>%
  select(cluster, cluster_source, source_file, gene, logFC, adj.P.Val, P.Value, t, B) %>%
  slice_head(n = 50)

datatable(
  top_pooled_deg,
  caption = "Top pooled DEG hits",
  filter = "top",
  rownames = FALSE,
  extensions = "Buttons",
  options = dt_opts_25
) %>%
  formatStyle(
    "logFC",
    color = styleInterval(0, c("blue", "red"))
  ) %>%
  formatSignif(columns = c("logFC", "adj.P.Val", "P.Value", "t", "B"), digits = 3)
```

## GSEA summary

```{r}
pooled_gsea_summary <- gsea_pooled %>%
  group_by(geneset, cluster_source, cluster, source_file) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    positive = sum(sig & NES > 0, na.rm = TRUE),
    negative = sum(sig & NES < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(geneset, cluster, source_file)

make_dt(
  pooled_gsea_summary,
  "Pooled GSEA summary by cluster and source_file",
  page_len = 10
)
```

## GSEA counts by cluster

```{r}
pooled_gsea_cluster <- gsea_pooled %>%
  group_by(geneset, cluster_source, cluster) %>%
  summarise(
    total_rows = n(),
    significant = sum(sig, na.rm = TRUE),
    positive = sum(sig & NES > 0, na.rm = TRUE),
    negative = sum(sig & NES < 0, na.rm = TRUE),
    n_source_files = n_distinct(source_file),
    .groups = "drop"
  ) %>%
  arrange(geneset, desc(significant))

make_dt(
  pooled_gsea_cluster,
  "Pooled GSEA counts by cluster",
  page_len = 10
)
```

## Plot: pooled GSEA by cluster

```{r}
pooled_gsea_cluster %>%
  ggplot(aes(x = reorder(cluster, significant), y = significant)) +
  geom_col(fill = "purple") +
  coord_flip() +
  facet_wrap(~geneset, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Pooled analysis: significant GSEA pathways by cluster",
    x = "Cluster",
    y = "Significant pathways"
  )
```

## Top pooled GSEA hits

```{r}
top_pooled_gsea <- gsea_pooled %>%
  filter(sig) %>%
  arrange(geneset, p.adjust, desc(abs(NES))) %>%
  select(geneset, cluster, cluster_source, source_file, ID, Description, NES, p.adjust, pvalue) %>%
  group_by(geneset) %>%
  slice_head(n = 30) %>%
  ungroup()

datatable(
  top_pooled_gsea,
  caption = "Top pooled GSEA hits",
  filter = "top",
  rownames = FALSE,
  extensions = "Buttons",
  options = dt_opts_25
) %>%
  formatStyle(
    "NES",
    color = styleInterval(0, c("blue", "red"))
  ) %>%
  formatSignif(columns = c("NES", "p.adjust", "pvalue"), digits = 3)
```

# Notes

* DEG significance is defined as `adj.P.Val < 0.05`
* GSEA significance is defined as `p.adjust < 0.05`
* Results are grouped using `cluster`, `cluster_source`, and `source_file`
* Paired and pooled are shown separately because they represent different study questions

# Session Info

```{r}
sessionInfo()
```

