---
title: "Single-Cell RNA-seq Analysis: Treatment Response"
subtitle: "Paired Before/After Analysis with Differential Expression and Pathway Enrichment"
pipeline_code_author: "Foo Cheung"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: false
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 8
    smooth-scroll: true
    number-sections: true
    echo: false
execute:
  warning: false
  message: false
  cache: false
  echo: false
params:
  config_file: "./config2.yaml"
  analysis_type: "paired_cross_pool"
---

```{css, echo=FALSE}
/* ── Typography ─────────────────────────────────────────────────────────── */
body { font-family: "Inter", "Helvetica Neue", Arial, sans-serif; font-size: 15px; line-height: 1.6; }
h1 { border-bottom: 2px solid #2C7BB6; padding-bottom: 6px; margin-top: 1.5rem; }
h2 { color: #2C3E50; margin-top: 1.2rem; }
h3 { color: #34495E; }

/* ── Callout tweaks ──────────────────────────────────────────────────────── */
.callout-note    { border-left-color: #2C7BB6 !important; background: #EBF4FC !important; }
.callout-tip     { border-left-color: #1A9641 !important; background: #EAF7ED !important; }
.callout-success { border-left-color: #00897B !important; background: #E0F2F1 !important; }
.callout-warning { border-left-color: #E87722 !important; }

/* ── DT tables ───────────────────────────────────────────────────────────── */
.dataTables_wrapper { font-size: 13px; }
thead tr th { background-color: #2C7BB6 !important; color: white !important; font-weight: 600 !important; }
.dataTables_filter input { border-radius: 4px; border: 1px solid #ccc; padding: 4px 8px; }

/* ── TOC ─────────────────────────────────────────────────────────────────── */
#TOC { font-size: 13px; border-right: 1px solid #e0e0e0; }
#TOC li { line-height: 1.7; }
#TOC a:hover { color: #2C7BB6; }

/* ── Figure captions ─────────────────────────────────────────────────────── */
.figure-caption { font-style: italic; color: #666; font-size: 12px; text-align: center; }

/* ── Section spacing ─────────────────────────────────────────────────────── */
section { margin-bottom: 2rem; }

/* ── Metric cards ────────────────────────────────────────────────────────── */
.metric-grid { display: flex; flex-wrap: wrap; gap: 12px; margin: 20px 0 28px 0; }
.metric-card {
  background: #f8fafc;
  border-radius: 10px;
  padding: 16px 20px;
  text-align: center;
  flex: 1;
  min-width: 110px;
  box-shadow: 0 1px 4px rgba(0,0,0,0.08);
  transition: transform 0.15s;
}
.metric-card:hover { transform: translateY(-2px); box-shadow: 0 3px 10px rgba(0,0,0,0.12); }
.metric-value { font-size: 1.9em; font-weight: 700; line-height: 1.1; }
.metric-label { font-size: 0.78em; color: #666; margin-top: 5px; letter-spacing: 0.02em; text-transform: uppercase; }

/* ── Parameter table ─────────────────────────────────────────────────────── */
.param-table table { width: 100%; border-collapse: collapse; font-size: 13.5px; }
.param-table td, .param-table th { padding: 7px 12px; border-bottom: 1px solid #eee; }
.param-table tr:last-child td { border-bottom: none; }
.param-table tr:hover td { background: #f5f9ff; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
set.seed(42)
```

```{r load_libraries, include=FALSE}
suppressPackageStartupMessages({
  library(yaml)
  library(Seurat)
  library(SeuratObject)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(data.table)
  library(tibble)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(ggrepel)
  library(pheatmap)
  library(RColorBrewer)
  library(DT)
  library(knitr)
  library(plotly)
  library(glue)
  library(fs)
  library(rstatix)
})

`%||%` <- function(x, y) if (is.null(x)) y else x

if (file.exists("scRNAseq_helpers.R"))  source("scRNAseq_helpers.R")
if (file.exists("enhanced_helpers.R"))  source("enhanced_helpers.R")

has_sc_prop <- requireNamespace("scProportionTest", quietly = TRUE)
if (has_sc_prop) suppressPackageStartupMessages(library(scProportionTest))
```

```{r load_config, include=FALSE}
config <- yaml::read_yaml(params$config_file)

RDS_FILE          <- config$input$rds_file
OUT_ROOT          <- config$output$root_dir
REUSE_RESULTS_ONLY <- config$output$use_cached_results %||% FALSE

ASSAY          <- config$metadata$assay
CLUSTER_COL    <- config$metadata$cluster_fine
CLUSTER_COL_L1 <- config$metadata$cluster_broad
SUBJECT_COL    <- config$metadata$subject_col
TIME_COL       <- config$metadata$treatment_col
BATCH_COL      <- config$metadata$batch_col

CONTRAST <- c(config$groups$treatment, config$groups$reference)

MT_PCT_MAX    <- config$qc$mt_pct_max
NFEATURE_MIN  <- config$qc$nfeature_min
NFEATURE_MAX  <- config$qc$nfeature_max
NCOUNT_MIN    <- config$qc$ncount_min
NCOUNT_MAX    <- config$qc$ncount_max

FDR_CUTOFF      <- config$analysis$fdr_cutoff
LOGFC_CUTOFF    <- config$analysis$logfc_cutoff
N_PERMUTATIONS  <- config$analysis$n_permutations %||% 1000
MIN_CELLS       <- config$analysis$min_cells_per_cluster %||% 20

ANALYSIS_TYPE <- params$analysis_type

CUSTOM_TEXT <- if (!is.null(config$custom_text)) config$custom_text else list()
get_custom_text <- function(section) {
  if (!is.null(CUSTOM_TEXT[[section]]) && nchar(CUSTOM_TEXT[[section]]) > 0)
    return(CUSTOM_TEXT[[section]])
  NULL
}

DIR_DEG_FINE   <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL,    "deg_results")
DIR_GSEA_FINE  <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL,    "gsea_results")
DIR_PLOTS_FINE <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL,    "plots")
DIR_DEG_MAIN   <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL_L1, "deg_results")
DIR_GSEA_MAIN  <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL_L1, "gsea_results")
DIR_PLOTS_MAIN <- file.path(OUT_ROOT, ANALYSIS_TYPE, CLUSTER_COL_L1, "plots")
DIR_EXPORTS    <- file.path(OUT_ROOT, ANALYSIS_TYPE, "exports")
DIR_SIGNATURES <- file.path(OUT_ROOT, ANALYSIS_TYPE, "signature_scores")

walk(c(DIR_DEG_FINE, DIR_GSEA_FINE, DIR_PLOTS_FINE,
       DIR_DEG_MAIN, DIR_GSEA_MAIN, DIR_PLOTS_MAIN,
       DIR_EXPORTS, DIR_SIGNATURES),
     ~dir_create(.x, recurse = TRUE))

limma_fine_path <- file.path(DIR_DEG_FINE,  sprintf("%s__ALL_clusters_limma.csv",    CLUSTER_COL))
limma_main_path <- file.path(DIR_DEG_MAIN,  sprintf("%s__ALL_clusters_limma.csv",    CLUSTER_COL_L1))
go_fine_path    <- file.path(DIR_GSEA_FINE, sprintf("%s__ALL_clusters_GSEA_GO.csv",  CLUSTER_COL))
go_main_path    <- file.path(DIR_GSEA_MAIN, sprintf("%s__ALL_clusters_GSEA_GO.csv",  CLUSTER_COL_L1))
re_fine_path    <- file.path(DIR_GSEA_FINE, sprintf("%s__ALL_clusters_GSEA_REACTOME.csv", CLUSTER_COL))
re_main_path    <- file.path(DIR_GSEA_MAIN, sprintf("%s__ALL_clusters_GSEA_REACTOME.csv", CLUSTER_COL_L1))

limma_fine_available <- file_exists(limma_fine_path)
limma_main_available <- file_exists(limma_main_path)
gsea_fine_available  <- file_exists(go_fine_path) || file_exists(re_fine_path)
gsea_main_available  <- file_exists(go_main_path) || file_exists(re_main_path)

upset_pdf    <- "COMBINED_DEMUX/23-301-4_GEX1_combined_DemuxOnly_Singlets_upset_donor_assignment.pdf"
upset_png    <- sub("\\.pdf$", ".png", upset_pdf)
pipeline_png <- "dynamic_graph_20260327_102225.png"
```

```{r load_precomputed_results, include=FALSE}
limma_fine    <- NULL
limma_main    <- NULL
gsea_fine_all <- NULL
gsea_main_all <- NULL

if (limma_fine_available) limma_fine <- fread(limma_fine_path)
if (limma_main_available) limma_main <- fread(limma_main_path)

if (file_exists(go_fine_path)) {
  gsea_fine_all <- fread(go_fine_path) %>% mutate(geneset = "GO:BP")
}
if (file_exists(re_fine_path)) {
  tmp <- fread(re_fine_path) %>% mutate(geneset = "REACTOME")
  gsea_fine_all <- if (is.null(gsea_fine_all)) tmp else bind_rows(gsea_fine_all, tmp)
}
if (file_exists(go_main_path)) {
  gsea_main_all <- fread(go_main_path) %>% mutate(geneset = "GO:BP")
}
if (file_exists(re_main_path)) {
  tmp <- fread(re_main_path) %>% mutate(geneset = "REACTOME")
  gsea_main_all <- if (is.null(gsea_main_all)) tmp else bind_rows(gsea_main_all, tmp)
}
```

# Summary {.unnumbered}

```{r exec_summary, include=FALSE}
obj_meta_only <- readRDS(RDS_FILE)

n_subjects      <- length(unique(obj_meta_only@meta.data[[SUBJECT_COL]]))
n_cells         <- ncol(obj_meta_only)
n_genes         <- nrow(obj_meta_only)
n_clusters_fine <- length(unique(obj_meta_only@meta.data[[CLUSTER_COL]]))
n_clusters_broad <- length(unique(obj_meta_only@meta.data[[CLUSTER_COL_L1]]))

if (!is.null(limma_fine)) {
  n_sig_genes <- limma_fine %>%
    filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>% nrow()
  n_sig_clusters <- limma_fine %>%
    filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
    pull(cluster) %>% unique() %>% length()
} else {
  n_sig_genes    <- NA
  n_sig_clusters <- NA
}

rm(obj_meta_only); gc()
```

::: {.callout-note icon="false"}

## Processing Pipeline

```{r pipeline_plot, fig.width=14, fig.height=6, out.width="100%"}
if (file.exists(pipeline_png)) {
  knitr::include_graphics(pipeline_png)
} else {
  cat("Pipeline image not found.\n")
}
```

This pipeline schematic shows the PBMC and Ileum processing and downstream analysis workflow used for this report.

:::

```{r summary_cards, results='asis'}
make_card <- function(value, label, color = "#2C7BB6") {
  glue::glue(
    '<div class="metric-card" style="border-top: 3px solid {color};">',
    '  <div class="metric-value" style="color:{color};">{value}</div>',
    '  <div class="metric-label">{label}</div>',
    '</div>'
  )
}

sig_genes_txt <- if (!is.na(n_sig_genes))  scales::comma(n_sig_genes)  else "—"
sig_clust_txt <- if (!is.na(n_sig_clusters)) as.character(n_sig_clusters) else "—"

cards <- paste(
  make_card(n_subjects,                    "Subjects",            "#2C7BB6"),
  make_card(scales::comma(n_cells),        "Cells (pre-QC)",      "#1A9641"),
  make_card(scales::comma(n_genes),        "Genes",               "#7B2D8B"),
  make_card(n_clusters_broad,              "Broad Cell Types",    "#D7191C"),
  make_card(n_clusters_fine,               "Fine Cell Types",     "#E87722"),
  make_card(sig_genes_txt,                 "Significant DEGs",    "#00897B"),
  make_card(sig_clust_txt,                 "Cell Types w/ DEGs",  "#5E35B1"),
  sep = "\n"
)

cat(glue::glue('<div class="metric-grid">\n{cards}\n</div>\n'))
```

::: {.callout-tip icon="false"}

## Analysis Overview

::: {.param-table}

| Parameter | Value |
|---|---|
| **Study Design** | `r ANALYSIS_TYPE` |
| **Comparison** | **`r CONTRAST[1]`** vs **`r CONTRAST[2]`** |
| **FDR Cutoff** | `r FDR_CUTOFF` |
| **logFC Cutoff** | `r LOGFC_CUTOFF` |
| **Min Cells / Cluster** | `r MIN_CELLS` |
| **Mito Threshold** | ≤ `r MT_PCT_MAX`% |
| **Gene Range** | `r NFEATURE_MIN` – `r NFEATURE_MAX` |
| **UMI Range** | `r scales::comma(NCOUNT_MIN)` – `r scales::comma(NCOUNT_MAX)` |

:::

:::

```{r introduction_section, results='asis'}
intro_text <- get_custom_text("introduction")
if (!is.null(intro_text)) {
  cat("\n## Introduction {.unnumbered}\n\n")
  cat(intro_text, "\n\n")
}
```

```{r objectives_section, results='asis'}
objectives_text <- get_custom_text("objectives")
if (!is.null(objectives_text)) {
  cat("\n## Study Objectives {.unnumbered}\n\n")
  cat(objectives_text, "\n\n")
}
```

# Data Loading & QC

```{r qc_context_section, results='asis'}
qc_context <- get_custom_text("qc_context")
if (!is.null(qc_context)) cat(qc_context, "\n\n")
```

```{r load_data}
message("Loading and processing data...")

obj <- readRDS(RDS_FILE)
obj <- tryCatch(JoinLayers(obj, assay = ASSAY), error = function(e) obj)
DefaultAssay(obj) <- ASSAY

req_cols <- c(CLUSTER_COL, CLUSTER_COL_L1, SUBJECT_COL, TIME_COL)
if (!is.null(BATCH_COL)) req_cols <- c(req_cols, BATCH_COL)
miss <- setdiff(req_cols, colnames(obj@meta.data))
if (length(miss)) stop(sprintf("Missing columns: %s", paste(miss, collapse = ", ")))

get_assay_matrix <- function(seu, assay, preferred = c("counts", "data", "scale.data")) {
  aa          <- seu[[assay]]
  layer_names <- tryCatch(SeuratObject::Layers(aa), error = function(e) character(0))
  for (ly in preferred) {
    if (ly %in% layer_names)
      return(SeuratObject::GetAssayData(seu, assay = assay, layer = ly))
  }
  stop(sprintf("None of the requested layers were found in assay '%s'. Available: %s",
               assay, paste(layer_names, collapse = ", ")))
}

counts_all <- get_assay_matrix(obj, ASSAY, preferred = c("counts", "data"))
meta_all   <- obj@meta.data

if (!"percent.mito" %in% colnames(meta_all)) {
  mt_genes   <- grep("^MT-|^Mt-|^mt-", rownames(counts_all), value = TRUE)
  mt_counts  <- if (length(mt_genes)) Matrix::colSums(counts_all[mt_genes, , drop = FALSE]) else rep(0, ncol(counts_all))
  lib_counts <- Matrix::colSums(counts_all)
  meta_all$percent.mito <- ifelse(lib_counts > 0, 100 * mt_counts / lib_counts, 0)
}
if (!"nFeature_RNA" %in% colnames(meta_all)) meta_all$nFeature_RNA <- Matrix::colSums(counts_all > 0)
if (!"nCount_RNA"   %in% colnames(meta_all)) meta_all$nCount_RNA   <- Matrix::colSums(counts_all)

if (exists("calculate_complexity")) {
  obj@meta.data <- meta_all
  obj  <- calculate_complexity(obj)
  meta_all <- obj@meta.data
} else {
  detected <- Matrix::colSums(counts_all > 0)
  total    <- Matrix::colSums(counts_all)
  meta_all$log10GenesPerUMI <- log10(detected + 1) / log10(total + 1)
  obj@meta.data <- meta_all
}

if (exists("analyze_qc_failures")) {
  qc_failure_analysis <- analyze_qc_failures(obj, config$qc)
} else {
  qc_failure_analysis <- list(
    summary = data.frame(
      total_cells       = ncol(obj),
      fail_mito         = sum(obj@meta.data$percent.mito    >  MT_PCT_MAX,   na.rm = TRUE),
      fail_nfeature_low = sum(obj@meta.data$nFeature_RNA    <  NFEATURE_MIN, na.rm = TRUE),
      fail_nfeature_high = sum(obj@meta.data$nFeature_RNA   >  NFEATURE_MAX, na.rm = TRUE),
      fail_ncount_low   = sum(obj@meta.data$nCount_RNA      <  NCOUNT_MIN,   na.rm = TRUE),
      fail_ncount_high  = sum(obj@meta.data$nCount_RNA      >  NCOUNT_MAX,   na.rm = TRUE)
    )
  )
}

keep_qc <- with(obj@meta.data,
  percent.mito <= MT_PCT_MAX &
  nFeature_RNA >= NFEATURE_MIN & nFeature_RNA <= NFEATURE_MAX &
  nCount_RNA   >= NCOUNT_MIN   & nCount_RNA   <= NCOUNT_MAX)

obj      <- subset(obj, cells = rownames(obj@meta.data)[keep_qc])
meta_all <- obj@meta.data
meta_all$SubjectTimePoint <- paste(meta_all[[SUBJECT_COL]], meta_all[[TIME_COL]], sep = "_")

vec    <- meta_all$SubjectTimePoint
parsed <- utils::strcapture(
  sprintf("^(.+?)_(%s|%s)$", CONTRAST[1], CONTRAST[2]),
  vec, proto = list(subj = character(), trt = character()))

subjects       <- sort(unique(parsed$subj[!is.na(parsed$subj)]))
ordered_levels <- as.character(unlist(lapply(subjects, function(s)
  c(paste0(s, "_", CONTRAST[2]), paste0(s, "_", CONTRAST[1])))))
ordered_levels <- c(ordered_levels, setdiff(unique(vec), ordered_levels))

meta_all$SubjectTimePoint_ordered <- factor(vec, levels = ordered_levels)
obj@meta.data <- meta_all

message(sprintf("QC complete: %s cells retained", ncol(obj)))
```

```{r filter_rare_clusters}
cluster_counts_fine  <- table(obj@meta.data[[CLUSTER_COL]])
cluster_counts_broad <- table(obj@meta.data[[CLUSTER_COL_L1]])
rare_clusters_fine   <- names(cluster_counts_fine[cluster_counts_fine   < MIN_CELLS])
rare_clusters_broad  <- names(cluster_counts_broad[cluster_counts_broad < MIN_CELLS])

cells_to_keep <- rownames(obj@meta.data)[
  !(obj@meta.data[[CLUSTER_COL]]    %in% rare_clusters_fine |
    obj@meta.data[[CLUSTER_COL_L1]] %in% rare_clusters_broad)]
obj <- subset(obj, cells = cells_to_keep)

n_clusters_fine  <- length(unique(obj@meta.data[[CLUSTER_COL]]))
n_clusters_broad <- length(unique(obj@meta.data[[CLUSTER_COL_L1]]))
```

::: {.callout-success}

## Dataset Summary

* **Cells retained:** `r comma(ncol(obj))`
* **QC Filters:** Mito ≤ `r MT_PCT_MAX`%, Genes `r NFEATURE_MIN`–`r NFEATURE_MAX`, UMIs `r comma(NCOUNT_MIN)`–`r comma(NCOUNT_MAX)`
* **Analysis type:** `r ANALYSIS_TYPE`

:::

## QC Failure Summary

```{r qc_table}
datatable(
  qc_failure_analysis$summary %>% pivot_longer(everything(), names_to = "Metric", values_to = "Count"),
  options = list(pageLength = 10, dom = "t"),
  caption = "QC Filter Statistics",
  rownames = FALSE
) %>% formatRound("Count", digits = 0)
```

## Interactive QC Plot

```{r qc_plot, fig.width=10, fig.height=8}
if (exists("plot_qc_interactive")) {
  plot_qc_interactive(obj, color_by = TIME_COL, annotation_col = CLUSTER_COL)
} else {
  p1 <- VlnPlot(obj, features = "nFeature_RNA", group.by = TIME_COL, pt.size = 0)
  p2 <- VlnPlot(obj, features = "nCount_RNA",   group.by = TIME_COL, pt.size = 0)
  p3 <- VlnPlot(obj, features = "percent.mito", group.by = TIME_COL, pt.size = 0)
  wrap_plots(p1, p2, p3, ncol = 1)
}
```

## Sample QC Report

```{r sample_qc}
if (exists("generate_sample_report")) {
  sample_report <- generate_sample_report(obj, SUBJECT_COL, TIME_COL)
} else {
  sample_report <- obj@meta.data %>%
    group_by(.data[[SUBJECT_COL]], .data[[TIME_COL]]) %>%
    summarise(n_cells     = n(),
              median_UMI  = median(nCount_RNA,    na.rm = TRUE),
              median_genes = median(nFeature_RNA, na.rm = TRUE),
              median_mito = median(percent.mito,  na.rm = TRUE),
              .groups = "drop") %>%
    rename(subject = 1, condition = 2)
}

datatable(
  sample_report,
  extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"),
                 pageLength = 20, scrollX = TRUE),
  caption = "Per-Sample Quality Metrics",
  rownames = FALSE
)
```

```{r qc_notes_section, results='asis'}
qc_notes <- get_custom_text("qc_notes")
if (!is.null(qc_notes)) { cat("\n## QC Notes\n\n"); cat(qc_notes, "\n\n") }
```

# Demultiplexing and Doublet Assessment

## UpSet Plot: Doublet Finders vs Demuxalot/Vireo SNP Demultiplexing

This UpSet plot summarises overlap across doublet-calling approaches and SNP-based demultiplexing, including Demuxalot singlet assignments.

```{r upset_plot, fig.width=12, fig.height=8, out.width="100%"}
if (file.exists(upset_png)) {
  knitr::include_graphics(upset_png)
} else if (file.exists(upset_pdf)) {
  knitr::include_graphics(upset_pdf)
} else {
  cat("UpSet plot file not found.\n")
}
```

# UMAP Visualizations

## UMAPs by Metadata

```{r umap_metadata, results='asis', fig.width=12, fig.height=10}
if (exists("generate_umap_tabset")) {
  generate_umap_tabset(
    obj,
    priority_cols = c("HTO_maxID", TIME_COL, "SubjectTimePoint_ordered", "Lane",
                      "percent.mt", "nFeature_RNA", CLUSTER_COL, CLUSTER_COL_L1,
                      "cluster", "lineage_call"),
    exclude_cols  = c("orig.ident", "RNA_snn_res.0.8"),
    max_tabs = 6
  )
} else {
  cols_to_plot <- intersect(c(TIME_COL, SUBJECT_COL, CLUSTER_COL, CLUSTER_COL_L1),
                            colnames(obj@meta.data))
  for (cc in cols_to_plot) {
    print(DimPlot(obj, reduction = "umap", group.by = cc, label = FALSE, pt.size = 0.4) +
            labs(title = paste("UMAP by", cc)) + theme_minimal())
  }
}
```

## Cell Type Annotation

```{r umap_celltypes, fig.width=14, fig.height=12}
p1 <- DimPlot(obj, reduction = "umap", group.by = CLUSTER_COL,
              label = TRUE, repel = TRUE, pt.size = 1) +
  labs(title = sprintf("Fine: %s (%d types)", CLUSTER_COL, n_clusters_fine)) +
  theme_minimal()

p2 <- DimPlot(obj, reduction = "umap", group.by = CLUSTER_COL_L1,
              label = TRUE, repel = TRUE, pt.size = 1) +
  labs(title = sprintf("Broad: %s (%d types)", CLUSTER_COL_L1, n_clusters_broad)) +
  theme_minimal()

p1 / p2
```

# Cell Type Annotation Summary

## Cell Counts by Annotation Level

```{r annotation_table}
count_L1 <- table(obj@meta.data[[CLUSTER_COL_L1]])
count_L2 <- table(obj@meta.data[[CLUSTER_COL]])

anno_summary <- bind_rows(
  tibble(Level = CLUSTER_COL_L1, `Cell Type` = names(count_L1),
         `N Cells` = as.integer(count_L1),
         `% of Total` = 100 * as.integer(count_L1) / ncol(obj)),
  tibble(Level = CLUSTER_COL,    `Cell Type` = names(count_L2),
         `N Cells` = as.integer(count_L2),
         `% of Total` = 100 * as.integer(count_L2) / ncol(obj))
) %>% arrange(Level, desc(`N Cells`))

datatable(anno_summary,
  options = list(pageLength = 15, scrollX = TRUE),
  caption = sprintf("Cell type distribution (%s and %s)", CLUSTER_COL_L1, CLUSTER_COL)
) %>%
  formatRound("% of Total", digits = 2) %>%
  formatRound("N Cells",    digits = 0)
```

## Cell Type Distribution Across Subjects and Timepoints (Broad)

```{r annotation_heatmap_broad, fig.width=12, fig.height=8}
counts_df <- as.data.frame(
  table(Cell_Type = obj@meta.data[[CLUSTER_COL_L1]],
        SubjectTP = obj@meta.data$SubjectTimePoint_ordered)
) %>% filter(Freq > 0)

ggplot(counts_df, aes(x = SubjectTP, y = Cell_Type, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), size = 3, color = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", name = "Cell Count",
                      trans = "log10", labels = comma) +
  labs(title    = "Cell Type Distribution Across Samples",
       subtitle = sprintf("%s (broad) by Subject and Timepoint", CLUSTER_COL_L1),
       x = "Subject / Timepoint", y = sprintf("Cell Type (%s)", CLUSTER_COL_L1)) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank(), legend.position = "right",
        plot.title  = element_text(face = "bold"))
```

## Cell Type Distribution Across Subjects and Timepoints (Fine)

```{r annotation_heatmap_fine, fig.width=12, fig.height=8}
counts_df <- as.data.frame(
  table(Cell_Type = obj@meta.data[[CLUSTER_COL]],
        SubjectTP = obj@meta.data$SubjectTimePoint_ordered)
) %>% filter(Freq > 0)

ggplot(counts_df, aes(x = SubjectTP, y = Cell_Type, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), size = 3, color = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", name = "Cell Count",
                      trans = "log10", labels = comma) +
  labs(title    = "Cell Type Distribution Across Samples",
       subtitle = sprintf("%s (fine) by Subject and Timepoint", CLUSTER_COL),
       x = "Subject / Timepoint", y = sprintf("Cell Type (%s)", CLUSTER_COL)) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank(), legend.position = "right",
        plot.title  = element_text(face = "bold"))
```

# Composition Analysis

```{r composition}
include_comp <- isTRUE(config$report$include_composition_section)

if (include_comp) {
  if (exists("calculate_proportions")) {
    prop_data_fine  <- calculate_proportions(obj, cluster_col = CLUSTER_COL,
                                              subject_col = SUBJECT_COL, condition_col = TIME_COL)
    prop_data_broad <- calculate_proportions(obj, cluster_col = CLUSTER_COL_L1,
                                              subject_col = SUBJECT_COL, condition_col = TIME_COL)
  } else {
    prop_data_fine  <- NULL
    prop_data_broad <- NULL
  }
}
```

## Paired Proportions (Fine)

```{r comp_fine, fig.width=14, fig.height=10}
if (include_comp && !is.null(prop_data_fine) && exists("plot_paired_proportions")) {
  plot_paired_proportions(prop_data_fine, cluster_col = CLUSTER_COL,
                          subject_col = SUBJECT_COL, condition_col = TIME_COL, ncol = 4)
}
```

## Paired Proportions (Broad)

```{r comp_broad, fig.width=14, fig.height=8}
if (include_comp && !is.null(prop_data_broad) && exists("plot_paired_proportions")) {
  plot_paired_proportions(prop_data_broad, cluster_col = CLUSTER_COL_L1,
                          subject_col = SUBJECT_COL, condition_col = TIME_COL, ncol = 3)
}
```

## Statistical Testing

```{r prop_test, fig.width=10, fig.height=8}
if (include_comp) {

  # ── Fallback: paired Wilcoxon on proportions ─────────────────────────────
  fallback_proportion_test <- function(obj, cluster_col, subject_col, condition_col, contrast) {
    meta <- obj@meta.data %>%
      mutate(cluster   = .data[[cluster_col]],
             subject   = .data[[subject_col]],
             condition = .data[[condition_col]])

    props <- meta %>%
      count(subject, condition, cluster) %>%
      group_by(subject, condition) %>%
      mutate(prop = n / sum(n)) %>%
      ungroup()

    purrr::map_dfr(unique(props$cluster), function(cl) {
      d  <- props %>% filter(cluster == cl)
      g1 <- d %>% filter(condition == contrast[1]) %>% arrange(subject) %>% pull(prop)
      g2 <- d %>% filter(condition == contrast[2]) %>% arrange(subject) %>% pull(prop)
      if (length(g1) < 2 || length(g2) < 2) return(NULL)
      wt <- tryCatch(wilcox.test(g1, g2, paired = TRUE), error = function(e) NULL)
      tibble(cluster    = cl,
             mean_treat = mean(g1, na.rm = TRUE),
             mean_ref   = mean(g2, na.rm = TRUE),
             log2ratio  = log2((mean(g1, na.rm = TRUE) + 1e-6) /
                               (mean(g2, na.rm = TRUE) + 1e-6)),
             p.value    = if (!is.null(wt)) wt$p.value else NA_real_)
    }) %>%
      mutate(p.adj       = p.adjust(p.value, method = "BH"),
             significant = p.adj <= FDR_CUTOFF)
  }

  plot_fallback_proportions <- function(res, title) {
    res %>%
      mutate(cluster   = reorder(cluster, log2ratio),
             direction = ifelse(log2ratio > 0, "Up", "Down")) %>%
      ggplot(aes(x = log2ratio, y = cluster, fill = direction, alpha = significant)) +
      geom_col(width = 0.7) +
      geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_text(aes(label = case_when(p.adj < 0.01 ~ "**", p.adj < 0.05 ~ "*", TRUE ~ ""),
                    x     = log2ratio + ifelse(log2ratio >= 0, 0.03, -0.03)),
                hjust = ifelse(res$log2ratio >= 0, 0, 1), size = 4, fontface = "bold") +
      scale_fill_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5")) +
      scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
      labs(title    = title,
           subtitle = sprintf("Paired Wilcoxon · %s vs %s  (* p.adj<0.05, ** p.adj<0.01)",
                              CONTRAST[1], CONTRAST[2]),
           x    = sprintf("log2 ratio (%s / %s)", CONTRAST[1], CONTRAST[2]),
           y    = "Cell Type",
           fill = "Direction") +
      theme_minimal(base_size = 12) +
      theme(panel.grid.major.y = element_blank(),
            legend.position    = "top",
            plot.title         = element_text(face = "bold"),
            plot.subtitle      = element_text(colour = "grey40", size = 9))
  }

  # ── Use scProportionTest if available, else fallback ──────────────────────
  if (has_sc_prop) {

    prop_test_results       <- NULL
    prop_test_results_broad <- NULL

    proptest_rds <- file.path(OUT_ROOT, ANALYSIS_TYPE, "proportion_test_fine.rds")
    if (REUSE_RESULTS_ONLY && file_exists(proptest_rds)) {
      prop_test_results <- readRDS(proptest_rds)
    } else {
      try({
        prop_test_results <- scProportionTest::permutation_test(
          scProportionTest::sc_utils(obj),
          cluster_identity = CLUSTER_COL,
          sample_identity  = TIME_COL,
          sample_1         = CONTRAST[2],
          sample_2         = CONTRAST[1],
          n_permutations   = N_PERMUTATIONS)
        saveRDS(prop_test_results, proptest_rds)
      }, silent = FALSE)
    }
    if (!is.null(prop_test_results)) {
      print(scProportionTest::permutation_plot(prop_test_results) +
              labs(title    = sprintf("Proportion Changes: %s (Fine)", CLUSTER_COL),
                   subtitle = sprintf("%s permutations", scales::comma(N_PERMUTATIONS))))
    }

    proptest_broad_rds <- file.path(OUT_ROOT, ANALYSIS_TYPE, "proportion_test_broad.rds")
    if (REUSE_RESULTS_ONLY && file_exists(proptest_broad_rds)) {
      prop_test_results_broad <- readRDS(proptest_broad_rds)
    } else {
      try({
        prop_test_results_broad <- scProportionTest::permutation_test(
          scProportionTest::sc_utils(obj),
          cluster_identity = CLUSTER_COL_L1,
          sample_identity  = TIME_COL,
          sample_1         = CONTRAST[2],
          sample_2         = CONTRAST[1],
          n_permutations   = N_PERMUTATIONS)
        saveRDS(prop_test_results_broad, proptest_broad_rds)
      }, silent = FALSE)
    }
    if (!is.null(prop_test_results_broad)) {
      print(scProportionTest::permutation_plot(prop_test_results_broad) +
              labs(title    = sprintf("Proportion Changes: %s (Broad)", CLUSTER_COL_L1),
                   subtitle = sprintf("%s permutations", scales::comma(N_PERMUTATIONS))))
    }

  } else {

    message("scProportionTest unavailable — using paired Wilcoxon fallback.")
    res_fine  <- fallback_proportion_test(obj, CLUSTER_COL,    SUBJECT_COL, TIME_COL, CONTRAST)
    res_broad <- fallback_proportion_test(obj, CLUSTER_COL_L1, SUBJECT_COL, TIME_COL, CONTRAST)

    print(plot_fallback_proportions(res_fine,
            sprintf("Proportion Changes: %s (Fine)", CLUSTER_COL)))
    print(plot_fallback_proportions(res_broad,
            sprintf("Proportion Changes: %s (Broad)", CLUSTER_COL_L1)))

    datatable(
      bind_rows(res_fine  %>% mutate(level = "Fine"),
                res_broad %>% mutate(level = "Broad")) %>%
        select(level, cluster, mean_treat, mean_ref, log2ratio, p.value, p.adj, significant) %>%
        arrange(level, p.adj),
      extensions = "Buttons",
      options    = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"),
                        pageLength = 20, scrollX = TRUE),
      caption    = sprintf("Cell Type Proportion Changes: %s vs %s (Paired Wilcoxon)",
                           CONTRAST[1], CONTRAST[2]),
      rownames = FALSE
    ) %>%
      formatRound(c("mean_treat", "mean_ref", "log2ratio"), 4) %>%
      formatSignif(c("p.value", "p.adj"), 3)
  }

} else {
  cat("Composition section disabled in config.\n")
}
```

# Differential Expression

```{r deg_context_section, results='asis'}
deg_context <- get_custom_text("deg_context")
if (!is.null(deg_context)) cat(deg_context, "\n\n")
```

## DEG Summary (Fine)

```{r deg_fine_summary, fig.width=12, fig.height=6}
if (!is.null(limma_fine) && nrow(limma_fine) > 0) {
  deg_summary_fine <- limma_fine %>%
    filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
    mutate(direction = factor(ifelse(logFC > 0, "Up", "Down"), levels = c("Down", "Up"))) %>%
    count(cluster, direction)

  ggplot(deg_summary_fine, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n), position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold") +
    scale_fill_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5")) +
    coord_flip() +
    labs(title    = sprintf("DEGs per Cell Type (%s)", CLUSTER_COL),
         subtitle = ANALYSIS_TYPE, x = "Cell Type", y = "DEGs") +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## DEG Summary (Broad)

```{r deg_broad_summary, fig.width=12, fig.height=6}
if (!is.null(limma_main) && nrow(limma_main) > 0) {
  deg_summary_broad <- limma_main %>%
    filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
    mutate(direction = factor(ifelse(logFC > 0, "Up", "Down"), levels = c("Down", "Up"))) %>%
    count(cluster, direction)

  ggplot(deg_summary_broad, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n), position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold") +
    scale_fill_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5")) +
    coord_flip() +
    labs(title    = sprintf("DEGs per Cell Type (%s, Broad)", CLUSTER_COL_L1),
         subtitle = ANALYSIS_TYPE, x = "Cell Type", y = "DEGs") +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## DEG Table (Fine)

```{r deg_fine_table}
if (!is.null(limma_fine) && nrow(limma_fine) > 0) {
  cols_keep <- intersect(c("gene", "logFC", "t", "P.Value", "adj.P.Val", "cluster",
                            "method", "contrast", "source_file"), names(limma_fine))
  datatable(
    limma_fine %>% filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
      select(all_of(cols_keep)),
    extensions = c("Buttons", "Scroller"),
    options    = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"),
                      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption  = sprintf("%s DEG Results", CLUSTER_COL),
    filter   = "top",
    rownames = FALSE
  ) %>% formatRound(c("logFC", "t"), 3) %>% formatSignif(c("P.Value", "adj.P.Val"), 3)
}
```

## DEG Table (Broad)

```{r deg_broad_table}
if (!is.null(limma_main) && nrow(limma_main) > 0) {
  cols_keep <- intersect(c("gene", "logFC", "t", "P.Value", "adj.P.Val", "cluster",
                            "method", "contrast", "source_file"), names(limma_main))
  datatable(
    limma_main %>% filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
      select(all_of(cols_keep)),
    extensions = c("Buttons", "Scroller"),
    options    = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"),
                      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption  = sprintf("%s DEG Results (Broad)", CLUSTER_COL_L1),
    filter   = "top",
    rownames = FALSE
  ) %>% formatRound(c("logFC", "t"), 3) %>% formatSignif(c("P.Value", "adj.P.Val"), 3)
}
```

# GSEA

```{r gsea_context_section, results='asis'}
gsea_context <- get_custom_text("gsea_context")
if (!is.null(gsea_context)) cat(gsea_context, "\n\n")
```

## GSEA Summary (Fine)

```{r gsea_fine_summary, fig.width=12, fig.height=8}
if (!is.null(gsea_fine_all) && nrow(gsea_fine_all) > 0) {
  gsea_counts <- gsea_fine_all %>%
    mutate(direction = case_when(
      p.adjust <= FDR_CUTOFF & NES > 0 ~ "Up",
      p.adjust <= FDR_CUTOFF & NES < 0 ~ "Down",
      TRUE ~ NA_character_)) %>%
    filter(!is.na(direction)) %>%
    count(cluster, geneset, direction) %>%
    mutate(direction = factor(direction, levels = c("Down", "Up")))

  ggplot(gsea_counts, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n), position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold", size = 3) +
    scale_fill_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5")) +
    facet_wrap(~ geneset, ncol = 1, scales = "free_y") +
    coord_flip() +
    labs(title    = sprintf("Pathways per Cell Type (%s)", CLUSTER_COL),
         subtitle = ANALYSIS_TYPE, x = "Cell Type", y = "Pathways") +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## GSEA Summary (Broad)

```{r gsea_broad_summary, fig.width=12, fig.height=8}
if (!is.null(gsea_main_all) && nrow(gsea_main_all) > 0) {
  gsea_counts_broad <- gsea_main_all %>%
    mutate(direction = case_when(
      p.adjust <= FDR_CUTOFF & NES > 0 ~ "Up",
      p.adjust <= FDR_CUTOFF & NES < 0 ~ "Down",
      TRUE ~ NA_character_)) %>%
    filter(!is.na(direction)) %>%
    count(cluster, geneset, direction) %>%
    mutate(direction = factor(direction, levels = c("Down", "Up")))

  ggplot(gsea_counts_broad, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n), position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold", size = 3) +
    scale_fill_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5")) +
    facet_wrap(~ geneset, ncol = 1, scales = "free_y") +
    coord_flip() +
    labs(title    = sprintf("Pathways per Cell Type (%s, Broad)", CLUSTER_COL_L1),
         subtitle = ANALYSIS_TYPE, x = "Cell Type", y = "Pathways") +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## GSEA Table (Fine)

```{r gsea_fine_table}
if (!is.null(gsea_fine_all) && nrow(gsea_fine_all) > 0) {
  cols_keep <- intersect(c("cluster", "geneset", "Description", "NES", "pvalue",
                            "p.adjust", "setSize", "contrast", "source_file"),
                         names(gsea_fine_all))
  datatable(
    gsea_fine_all %>% filter(p.adjust <= FDR_CUTOFF) %>% select(all_of(cols_keep)),
    extensions = c("Buttons", "Scroller"),
    options    = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"),
                      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption  = sprintf("%s GSEA Results", CLUSTER_COL),
    filter   = "top",
    rownames = FALSE
  ) %>% formatRound("NES", 3) %>% formatSignif(c("pvalue", "p.adjust"), 3)
}
```

## GSEA Table (Broad)

```{r gsea_broad_table}
if (!is.null(gsea_main_all) && nrow(gsea_main_all) > 0) {
  cols_keep <- intersect(c("cluster", "geneset", "Description", "NES", "pvalue",
                            "p.adjust", "setSize", "contrast", "source_file"),
                         names(gsea_main_all))
  datatable(
    gsea_main_all %>% filter(p.adjust <= FDR_CUTOFF) %>% select(all_of(cols_keep)),
    extensions = c("Buttons", "Scroller"),
    options    = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"),
                      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption  = sprintf("%s GSEA Results (Broad)", CLUSTER_COL_L1),
    filter   = "top",
    rownames = FALSE
  ) %>% formatRound("NES", 3) %>% formatSignif(c("pvalue", "p.adjust"), 3)
}
```

# Most Variable Pathways

```{r gsea_heatmap_helper, include=FALSE}
plot_gsea_heatmap <- function(gsea_data, title_suffix = "", n_top = 25, fdr = FDR_CUTOFF) {

  required <- c("ID", "Description", "NES", "p.adjust", "cluster", "geneset")
  if (is.null(gsea_data) || nrow(gsea_data) == 0) return(NULL)
  if (!all(required %in% names(gsea_data))) return(NULL)

  top_ids <- gsea_data %>%
    filter(p.adjust <= fdr) %>%
    group_by(ID, Description) %>%
    summarise(var_NES = var(NES, na.rm = TRUE), .groups = "drop") %>%
    slice_max(order_by = var_NES, n = n_top) %>%
    pull(ID)

  if (length(top_ids) == 0) { cat("No significant variable pathways found.\n"); return(NULL) }

  plot_data <- gsea_data %>%
    filter(ID %in% top_ids) %>%
    mutate(
      label     = ifelse(nchar(Description) > 45,
                         paste0(substr(Description, 1, 42), "\u2026"), Description),
      sig_label = case_when(p.adjust < 0.001 ~ "***",
                            p.adjust < 0.01  ~ "**",
                            p.adjust < 0.05  ~ "*",
                            TRUE             ~ ""),
      NES_capped = pmax(pmin(NES, 4), -4)
    )

  # Order pathways by mean NES
  pathway_order <- plot_data %>%
    group_by(label) %>%
    summarise(mean_NES = mean(NES, na.rm = TRUE), .groups = "drop") %>%
    arrange(mean_NES) %>% pull(label)

  plot_data <- plot_data %>% mutate(label = factor(label, levels = pathway_order))

  ggplot(plot_data, aes(x = cluster, y = label, fill = NES_capped)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = sig_label), size = 3.2, colour = "grey20",
              vjust = 0.75, fontface = "bold") +
    scale_fill_gradient2(
      low      = "#3182BD",
      mid      = "white",
      high     = "#E64B35",
      midpoint = 0,
      limits   = c(-4, 4),
      oob      = scales::squish,
      name     = "NES",
      guide    = guide_colorbar(barheight = 8, barwidth = 0.8,
                                title.position = "top", title.hjust = 0.5)
    ) +
    facet_wrap(~ geneset, scales = "free_y", ncol = 1) +
    labs(
      title    = sprintf("Top Variable Pathways \u2014 %s", title_suffix),
      subtitle = "Ordered by mean NES  \u00b7  * p<0.05  ** p<0.01  *** p<0.001",
      x        = "Cell Type",
      y        = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 40, hjust = 1, size = 9, colour = "grey20"),
      axis.text.y      = element_text(size = 8.5, colour = "grey20"),
      axis.ticks       = element_blank(),
      panel.grid       = element_blank(),
      panel.spacing    = unit(0.8, "lines"),
      strip.background = element_rect(fill = "#2C3E50", colour = NA),
      strip.text       = element_text(colour = "white", face = "bold",
                                       size = 10, margin = margin(4, 4, 4, 4)),
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(size = 9, colour = "grey40", margin = margin(b = 8)),
      legend.position  = "right",
      legend.title     = element_text(size = 9, face = "bold")
    )
}
```

```{r gsea_main_var_heatmap, fig.width=13, fig.height=9}
p <- plot_gsea_heatmap(gsea_main_all, n_top = 20,
                        title_suffix = sprintf("%s (Broad)", CLUSTER_COL_L1))
if (!is.null(p)) print(p)
```

```{r gsea_fine_varheatmap, fig.width=14, fig.height=11}
p <- plot_gsea_heatmap(gsea_fine_all, n_top = 30,
                        title_suffix = sprintf("%s (Fine)", CLUSTER_COL))
if (!is.null(p)) print(p)
```

# Marker Genes

```{r markers, fig.width=14, fig.height=16}
include_marker <- isTRUE(config$report$include_marker_section)

if (include_marker && !is.null(config$markers)) {
  all_markers       <- unique(unlist(config$markers[!names(config$markers) %in% "custom_genes"]))
  available_markers <- intersect(all_markers, rownames(obj))
  if (length(available_markers) > 0) {
    FeaturePlot(obj, features = head(available_markers, 16), ncol = 4,
                pt.size = 1, order = TRUE) &
      theme_minimal(base_size = 10)
  }
}
```

# Methods

::: {.callout-note icon="false"}

## Computational Methods

**QC:** Mito ≤ `r MT_PCT_MAX`%, Genes `r NFEATURE_MIN`–`r NFEATURE_MAX`, UMIs `r comma(NCOUNT_MIN)`–`r comma(NCOUNT_MAX)`

**DE Analysis:** limma pseudobulk, paired design, FDR ≤ `r FDR_CUTOFF`, |log2FC| ≥ `r LOGFC_CUTOFF`

**GSEA:** Gene Set Enrichment Analysis, FDR ≤ `r FDR_CUTOFF`

**Software:** R `r R.version.string`, Seurat `r packageVersion("Seurat")`

:::

```{r methods_additional_section, results='asis'}
methods_additional <- get_custom_text("methods_additional")
if (!is.null(methods_additional)) {
  cat("\n## Additional Methods\n\n")
  cat(methods_additional, "\n\n")
}
```

```{r discussion_section, results='asis'}
discussion <- get_custom_text("discussion")
if (!is.null(discussion)) {
  cat("\n# Discussion {.unnumbered}\n\n")
  cat(discussion, "\n\n")
}
```

```{r conclusions_section, results='asis'}
conclusions <- get_custom_text("conclusions")
if (!is.null(conclusions)) {
  cat("\n# Conclusions {.unnumbered}\n\n")
  cat(conclusions, "\n\n")
}
```

```{r future_directions_section, results='asis'}
future_directions <- get_custom_text("future_directions")
if (!is.null(future_directions)) {
  cat("\n# Future Directions {.unnumbered}\n\n")
  cat(future_directions, "\n\n")
}
```

# Session Info

```{r session_info, results='asis'}
si <- sessionInfo()
cat(sprintf(
  "::: {.callout-note icon='false' collapse='true'}\n## Session Information\n\n**R version:** %s\n\n**Platform:** %s\n\n**Date rendered:** %s\n\n",
  R.version.string, R.version$platform, Sys.Date()
))
cat("**Key packages:**\n\n")
pkg_versions <- c("Seurat", "limma", "ggplot2", "dplyr", "data.table", "purrr", "fs")
for (pkg in pkg_versions) {
  v <- tryCatch(as.character(packageVersion(pkg)), error = function(e) "not installed")
  cat(sprintf("- `%s` %s\n", pkg, v))
}
cat("\n:::\n")
```
