---
title: "PBMC UMAP — All Reductions × All Groupings × All Sample Subsets"
author: "foo cheung"
format:
  html:
    toc: true
    toc-depth: 4
    embed-resources: true
    code-fold: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook generates **every combination** of:

- **3 UMAP reductions**: No Harmony · Pool-only Harmony · Subject + Pool Harmony  
- **4 grouping variables**: Condition · Pool · Monaco annotation · Subject  
- **6 sample filters**: All samples · HC only · Severe only · Treated only · Untreated only · HC + Severe (group comparison)

Additionally, for each reduction it produces per-subject subset panels (Monaco coloured) and per-group subset panels for all defined groups.

All figures are saved as PNG and embedded inline.

---

# Setup

```{r setup}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(scales)
library(knitr)
library(stringr)

# ── output directory ──────────────────────────────────────────────────────────
dir.create("umap_outputs", showWarnings = FALSE)
```

# Load object

```{r load}
obj_all <- readRDS("PBMC_RDS/ALL_harmony_azimuth.rds")

# Column names — adjust if yours differ
CONDITION_COL  <- "Treatment"
POOL_COL       <- "HTO_pool"
SUBJECT_COL    <- "Subject_raw"
CLUSTER_COL_L1 <- "Monaco_main"

# The three reductions to compare
REDUCTIONS <- c(
  "No Harmony"           = "umap",
  "Harmony (pool)"       = "umap_harmony_pool",
  "Harmony (subj+pool)"  = "umap_harmony_subject_pool"
)

message(sprintf("Total cells: %d", ncol(obj_all)))
message("Conditions:  ", paste(sort(unique(obj_all@meta.data[[CONDITION_COL]])), collapse = ", "))
message("Pools:       ", paste(sort(unique(obj_all@meta.data[[POOL_COL]])),      collapse = ", "))
message("Subjects:    ", paste(sort(unique(obj_all@meta.data[[SUBJECT_COL]])),   collapse = ", "))
message("Reductions:  ", paste(Reductions(obj_all), collapse = ", "))
```

# Palettes & helper labels

```{r palettes}
# ── Monaco colours ────────────────────────────────────────────────────────────
monaco_types <- sort(unique(obj_all@meta.data[[CLUSTER_COL_L1]]))
n_types      <- length(monaco_types)
base_cols12  <- c("#E64B35","#4DBBD5","#00A087","#3C5488","#F39B7F","#8491B4",
                  "#91D1C2","#DC0000","#7E6148","#B09C85","#3B4992","#EE0000")
monaco_colours <- if (n_types <= 12) {
  setNames(base_cols12[seq_len(n_types)], monaco_types)
} else {
  setNames(scales::hue_pal()(n_types), monaco_types)
}

make_palette <- function(types) {
  n <- length(types)
  setNames(if (n <= 12) base_cols12[seq_len(n)] else scales::hue_pal()(n), types)
}

# ── Pool colours ──────────────────────────────────────────────────────────────
pool_levels   <- sort(unique(obj_all@meta.data[[POOL_COL]]))
pool_colours  <- c("Pool1" = "#2C7BB6", "Pool2" = "#D7191C")
extra_pools   <- setdiff(pool_levels, names(pool_colours))
if (length(extra_pools)) {
  pool_colours <- c(pool_colours, setNames(scales::hue_pal()(length(extra_pools)), extra_pools))
}

# ── Condition colours ─────────────────────────────────────────────────────────
condition_levels <- sort(unique(obj_all@meta.data[[CONDITION_COL]]))
condition_cols   <- setNames(scales::hue_pal()(length(condition_levels)), condition_levels)

# ── Subject colours ───────────────────────────────────────────────────────────
subject_levels <- sort(unique(obj_all@meta.data[[SUBJECT_COL]]))
subject_cols   <- setNames(scales::hue_pal()(length(subject_levels)), subject_levels)

# ── Subject_Pool helper label ─────────────────────────────────────────────────
obj_all@meta.data$Subject_Pool <- paste(
  obj_all@meta.data[[SUBJECT_COL]],
  obj_all@meta.data[[POOL_COL]], sep = "_"
)
split_levels <- obj_all@meta.data %>%
  as.data.frame() %>%
  dplyr::select(all_of(c(SUBJECT_COL, POOL_COL)), Subject_Pool) %>%
  dplyr::distinct() %>%
  dplyr::arrange(.data[[SUBJECT_COL]], .data[[POOL_COL]]) %>%
  dplyr::pull(Subject_Pool)
obj_all@meta.data$Subject_Pool <- factor(obj_all@meta.data$Subject_Pool,
                                          levels = split_levels)
```

# Define sample subsets

```{r subsets}
# ── Filters are built dynamically from the actual condition levels in your data.
# No hard-coded group names — works regardless of what your conditions are called.

condition_levels_all <- sort(unique(obj_all@meta.data[[CONDITION_COL]]))
message("Building filters for conditions: ", paste(condition_levels_all, collapse = ", "))

# 1. "All samples" filter
single_filters <- list("All samples" = quote(TRUE))

# 2. One filter per individual condition  (e.g. "HC only", "Severe only", …)
for (cond in condition_levels_all) {
  nm <- sprintf("%s only", cond)
  single_filters[[nm]] <- bquote(.(as.name(CONDITION_COL)) == .(cond))
}

# Final: All samples + one filter per individual condition only
SAMPLE_FILTERS <- single_filters

message(sprintf("Total sample filters defined: %d", length(SAMPLE_FILTERS)))
message(paste(" -", names(SAMPLE_FILTERS), collapse = "\n"))

# Build subset objects once (avoids repeated subsetting)
# "All samples" uses obj_all directly; conditions use subset()
subset_objects <- lapply(names(SAMPLE_FILTERS), function(nm) {
  if (nm == "All samples") return(obj_all)
  filt <- SAMPLE_FILTERS[[nm]]
  tryCatch(
    subset(obj_all, subset = !!filt),
    error = function(e) {
      message("  [skip] subset '", nm, "' is empty or filter failed: ", conditionMessage(e))
      NULL
    }
  )
})
names(subset_objects) <- names(SAMPLE_FILTERS)

# Report
for (nm in names(subset_objects)) {
  obj <- subset_objects[[nm]]
  if (is.null(obj)) {
    message(sprintf("  %-25s  [EMPTY — skipped]", nm))
  } else {
    message(sprintf("  %-25s  n = %d cells", nm, ncol(obj)))
  }
}
```

# Helper functions

```{r helpers}
# ── Save + embed a figure ─────────────────────────────────────────────────────
save_and_show <- function(plot, filename, width, height, dpi = 200) {
  path <- file.path("umap_outputs", filename)
  ggplot2::ggsave(path, plot = plot, width = width, height = height,
                  dpi = dpi, bg = "white")
  # explicit print() is required when called inside loops with results='asis'
  print(knitr::include_graphics(path))
}

# ── Single DimPlot with standard styling ─────────────────────────────────────
make_dimplot <- function(obj, reduction, group_by, colours,
                         title = NULL, subtitle = NULL,
                         label = FALSE, label.size = 3, pt.size = 0.8) {
  p <- DimPlot(
    obj,
    reduction  = reduction,
    group.by   = group_by,
    cols       = colours,
    pt.size    = pt.size,
    label      = label,
    label.size = label.size,
    repel      = label,
    label.box  = label,
    shuffle    = TRUE
  ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, colour = "grey50"),
      legend.key.size = unit(0.4, "cm"),
      legend.text   = element_text(size = 8)
    )

  if (!is.null(title))    p <- p + labs(title = title)
  if (!is.null(subtitle)) p <- p + labs(subtitle = subtitle)
  p
}

# ── Per-subject subset panel row ──────────────────────────────────────────────
make_subject_panels <- function(obj, reduction, ncol_max = 4) {
  subjs <- sort(unique(obj@meta.data[[SUBJECT_COL]]))
  panels <- lapply(subjs, function(s) {
    sub <- tryCatch(
      subset(obj, subset = Subject_raw == s),
      error = function(e) NULL
    )
    if (is.null(sub) || ncol(sub) == 0) return(NULL)

    types_here <- sort(unique(sub@meta.data[[CLUSTER_COL_L1]]))
    cols_here  <- monaco_colours[types_here]

    DimPlot(sub, reduction = reduction,
            group.by = CLUSTER_COL_L1, cols = cols_here,
            pt.size = 1.2, shuffle = TRUE) +
      labs(title = s, x = NULL, y = NULL) +
      theme_minimal(base_size = 10) +
      theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 10),
            legend.position = "none",
            axis.text = element_text(size = 8))
  })
  panels <- Filter(Negate(is.null), panels)
  if (length(panels) == 0) return(NULL)
  wrap_plots(panels, ncol = min(ncol_max, length(panels)))
}

# ── Per-subject × per-pool subset panels ──────────────────────────────────────
# Produces one panel per subject-pool combination, labelled "Subject (Pool)"
make_subject_pool_panels <- function(obj, reduction, ncol_max = 4) {
  meta <- obj@meta.data %>% as.data.frame()

  # All unique subject × pool combos present in this object
  combos <- meta %>%
    dplyr::select(all_of(c(SUBJECT_COL, POOL_COL))) %>%
    dplyr::distinct() %>%
    dplyr::arrange(.data[[SUBJECT_COL]], .data[[POOL_COL]])

  panels <- lapply(seq_len(nrow(combos)), function(i) {
    s <- combos[[SUBJECT_COL]][i]
    p <- combos[[POOL_COL]][i]

    sub <- tryCatch(
      subset(obj, subset = Subject_raw == s & HTO_pool == p),
      error = function(e) NULL
    )
    if (is.null(sub) || ncol(sub) == 0) return(NULL)

    types_here <- sort(unique(sub@meta.data[[CLUSTER_COL_L1]]))
    cols_here  <- monaco_colours[types_here]
    panel_title <- sprintf("%s\n(%s)", s, p)

    DimPlot(sub, reduction = reduction,
            group.by = CLUSTER_COL_L1, cols = cols_here,
            pt.size = 1.2, shuffle = TRUE) +
      labs(title = panel_title, x = NULL, y = NULL) +
      theme_minimal(base_size = 10) +
      theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 9),
            legend.position = "none",
            axis.text = element_text(size = 8))
  })
  panels <- Filter(Negate(is.null), panels)
  if (length(panels) == 0) return(NULL)
  wrap_plots(panels, ncol = min(ncol_max, length(panels)))
}

# ── Shared legend strip (Monaco) ──────────────────────────────────────────────
monaco_legend_plot <- function(obj) {
  types_here <- sort(unique(obj@meta.data[[CLUSTER_COL_L1]]))
  cols_here  <- monaco_colours[types_here]
  tmp <- DimPlot(obj, reduction = names(REDUCTIONS)[1],
                 group.by = CLUSTER_COL_L1, cols = cols_here,
                 pt.size = 0.01) +
    theme(legend.position = "right",
          legend.key.size = unit(0.45, "cm"),
          legend.text     = element_text(size = 9)) +
    guides(colour = guide_legend(ncol = 1, override.aes = list(size = 3)))
  cowplot::get_legend(tmp)   # requires cowplot; falls back gracefully
}
```

---

# Section 1 — Group-level comparison: All 4 groupings × 3 reductions

For each **sample filter** (All / HC / Severe / Treated / Untreated / HC+Severe) we produce a 3-row × 4-column grid:

- **Rows** = reductions (No Harmony · Pool Harmony · Subject+Pool Harmony)  
- **Columns** = groupings (Condition · Pool · Monaco · Subject)

```{r group_level_grids, results='asis'}
for (subset_name in names(subset_objects)) {
  obj_sub <- subset_objects[[subset_name]]
  if (is.null(obj_sub)) next
  n_cells <- ncol(obj_sub)

  cat(sprintf("\n\n## %s  *(n = %s cells)*\n\n", subset_name, scales::comma(n_cells)))

  # Rebuild per-subset condition colours (only levels present)
  cond_here    <- sort(unique(obj_sub@meta.data[[CONDITION_COL]]))
  pool_here    <- sort(unique(obj_sub@meta.data[[POOL_COL]]))
  subject_here <- sort(unique(obj_sub@meta.data[[SUBJECT_COL]]))
  monaco_here  <- sort(unique(obj_sub@meta.data[[CLUSTER_COL_L1]]))

  cond_c    <- condition_cols[intersect(names(condition_cols), cond_here)]
  pool_c    <- pool_colours [intersect(names(pool_colours),    pool_here)]
  subject_c <- subject_cols [intersect(names(subject_cols),    subject_here)]
  monaco_c  <- monaco_colours[intersect(names(monaco_colours), monaco_here)]

  row_panels <- list()

  for (red_label in names(REDUCTIONS)) {
    red_key <- REDUCTIONS[red_label]

    # Skip reductions not present
    if (!red_key %in% Reductions(obj_sub)) {
      message("  Skipping reduction '", red_key, "' — not found in subset '", subset_name, "'")
      next
    }

    p_cond <- make_dimplot(obj_sub, red_key, CONDITION_COL, cond_c,
                           title    = sprintf("Condition | %s", red_label),
                           subtitle = subset_name)

    p_pool <- make_dimplot(obj_sub, red_key, POOL_COL, pool_c,
                           title    = sprintf("Pool | %s", red_label),
                           subtitle = subset_name)

    p_monaco <- make_dimplot(obj_sub, red_key, CLUSTER_COL_L1, monaco_c,
                             title    = sprintf("Monaco | %s", red_label),
                             subtitle = subset_name,
                             label = TRUE, label.size = 2.8)

    p_subject <- make_dimplot(obj_sub, red_key, SUBJECT_COL, subject_c,
                              title    = sprintf("Subject | %s", red_label),
                              subtitle = subset_name)

    row_panels[[red_label]] <- p_cond | p_pool | p_monaco | p_subject
  }

  if (length(row_panels) == 0) next

  fig <- Reduce("/", row_panels) +
    plot_annotation(
      title    = sprintf("UMAP comparison — %s", subset_name),
      subtitle = "Rows = reductions · Columns = grouping variable",
      theme    = theme(plot.title    = element_text(face = "bold", size = 14),
                       plot.subtitle = element_text(size = 10, colour = "grey40"))
    )

  safe_name <- str_replace_all(subset_name, "[^A-Za-z0-9_]", "_")
  fname     <- sprintf("grid_group_%s.png", safe_name)

  save_and_show(fig,
                filename = fname,
                width    = 22,
                height   = 6 * length(row_panels),
                dpi      = 180)
}
```

---

# Section 2 — Per-subject and per-subject × pool panels × 3 reductions

For each **sample filter** and each **reduction** we produce two panel grids:

- **2a — Per subject**: one panel per subject, Monaco coloured, true subset (no faint background)
- **2b — Per subject × pool**: one panel per subject-pool combination, so you can see intra-subject pool variation

```{r per_subject_panels, results='asis'}
for (subset_name in names(subset_objects)) {
  obj_sub <- subset_objects[[subset_name]]
  if (is.null(obj_sub)) next

  n_cells   <- ncol(obj_sub)
  subj_vec  <- sort(unique(obj_sub@meta.data[[SUBJECT_COL]]))
  pool_vec  <- sort(unique(obj_sub@meta.data[[POOL_COL]]))
  sp_combos <- nrow(unique(obj_sub@meta.data[, c(SUBJECT_COL, POOL_COL)]))

  ncol_subj <- min(5, length(subj_vec))
  ncol_sp   <- min(5, sp_combos)

  cat(sprintf("\n\n## %s  *(n = %s cells, %d subjects, %d pools)*\n\n",
              subset_name, scales::comma(n_cells), length(subj_vec), length(pool_vec)))

  for (red_label in names(REDUCTIONS)) {
    red_key <- REDUCTIONS[red_label]
    if (!red_key %in% Reductions(obj_sub)) next

    safe_name <- str_replace_all(subset_name, "[^A-Za-z0-9_]", "_")
    safe_red  <- str_replace_all(red_label,   "[^A-Za-z0-9_]", "_")

    # ── 2a: Per subject ───────────────────────────────────────────────────────
    cat(sprintf("\n\n### %s — %s — Per subject\n\n", subset_name, red_label))

    subj_grid <- make_subject_panels(obj_sub, reduction = red_key, ncol_max = ncol_subj)
    if (is.null(subj_grid)) {
      cat("*No subject panels produced.*\n\n")
    } else {
      fname    <- sprintf("subj_panels_%s_%s.png", safe_name, safe_red)
      n_rows_p <- ceiling(length(subj_vec) / ncol_subj)
      save_and_show(subj_grid,
                    filename = fname,
                    width    = min(4.5, 22 / ncol_subj) * ncol_subj,
                    height   = 4.2 * n_rows_p,
                    dpi      = 180)
    }

    # ── 2b: Per subject × pool ────────────────────────────────────────────────
    cat(sprintf("\n\n### %s — %s — Per subject × pool\n\n", subset_name, red_label))

    sp_grid <- make_subject_pool_panels(obj_sub, reduction = red_key, ncol_max = ncol_sp)
    if (is.null(sp_grid)) {
      cat("*No subject × pool panels produced.*\n\n")
    } else {
      fname    <- sprintf("subj_pool_panels_%s_%s.png", safe_name, safe_red)
      n_rows_p <- ceiling(sp_combos / ncol_sp)
      save_and_show(sp_grid,
                    filename = fname,
                    width    = min(4.5, 22 / ncol_sp) * ncol_sp,
                    height   = 4.2 * n_rows_p,
                    dpi      = 180)
    }
  }
}
```

---

# Section 3 — Side-by-side reduction comparison for each grouping

Here each panel shows the same grouping variable across all 3 reductions side-by-side, making it easy to judge how Harmony shifts the embedding for a specific colour axis.

```{r side_by_side, results='asis'}
groupings <- list(
  "Condition" = list(col = CONDITION_COL, cols_fn = function(obj) {
    lvls <- sort(unique(obj@meta.data[[CONDITION_COL]]))
    condition_cols[intersect(names(condition_cols), lvls)]
  }, label = FALSE),
  "Pool" = list(col = POOL_COL, cols_fn = function(obj) {
    lvls <- sort(unique(obj@meta.data[[POOL_COL]]))
    pool_colours[intersect(names(pool_colours), lvls)]
  }, label = FALSE),
  "Monaco" = list(col = CLUSTER_COL_L1, cols_fn = function(obj) {
    lvls <- sort(unique(obj@meta.data[[CLUSTER_COL_L1]]))
    monaco_colours[intersect(names(monaco_colours), lvls)]
  }, label = TRUE),
  "Subject" = list(col = SUBJECT_COL, cols_fn = function(obj) {
    lvls <- sort(unique(obj@meta.data[[SUBJECT_COL]]))
    subject_cols[intersect(names(subject_cols), lvls)]
  }, label = FALSE)
)

for (subset_name in names(subset_objects)) {
  obj_sub <- subset_objects[[subset_name]]
  if (is.null(obj_sub)) next

  cat(sprintf("\n\n## %s\n\n", subset_name))

  for (grp_name in names(groupings)) {
    grp  <- groupings[[grp_name]]
    cols <- grp$cols_fn(obj_sub)

    cat(sprintf("\n\n### %s — coloured by %s\n\n", subset_name, grp_name))

    panels <- list()
    for (red_label in names(REDUCTIONS)) {
      red_key <- REDUCTIONS[red_label]
      if (!red_key %in% Reductions(obj_sub)) next

      panels[[red_label]] <- make_dimplot(
        obj_sub, red_key, grp$col, cols,
        title    = red_label,
        subtitle = sprintf("%s | %s", subset_name, grp_name),
        label    = grp$label, label.size = 2.8
      )
    }

    if (length(panels) == 0) next

    fig <- Reduce("|", panels) +
      plot_annotation(
        title = sprintf("Effect of Harmony — %s | coloured by %s", subset_name, grp_name),
        theme = theme(plot.title = element_text(face = "bold", size = 13))
      )

    safe_sub <- str_replace_all(subset_name, "[^A-Za-z0-9_]", "_")
    safe_grp <- str_replace_all(grp_name,    "[^A-Za-z0-9_]", "_")
    fname    <- sprintf("sbs_%s_%s.png", safe_sub, safe_grp)

    save_and_show(fig,
                  filename = fname,
                  width    = 8 * length(panels),
                  height   = 7,
                  dpi      = 180)
  }
}
```

---

# Section 4 — Quick summary: All sample subsets × Monaco × all 3 reductions

A compact 6-row × 3-column overview so you can see how Monaco cell-type clusters look across every sample filter under each reduction at a glance.

```{r summary_monaco, results='asis'}
cat("\n\n## Monaco annotation across all subsets & reductions\n\n")

summary_rows <- list()

for (subset_name in names(subset_objects)) {
  obj_sub <- subset_objects[[subset_name]]
  if (is.null(obj_sub)) next

  monaco_here <- sort(unique(obj_sub@meta.data[[CLUSTER_COL_L1]]))
  monaco_c    <- monaco_colours[intersect(names(monaco_colours), monaco_here)]
  n_cells     <- ncol(obj_sub)

  row_p <- list()
  for (red_label in names(REDUCTIONS)) {
    red_key <- REDUCTIONS[red_label]
    if (!red_key %in% Reductions(obj_sub)) next

    row_p[[red_label]] <- make_dimplot(
      obj_sub, red_key, CLUSTER_COL_L1, monaco_c,
      title    = sprintf("%s | %s", subset_name, red_label),
      subtitle = scales::comma(n_cells),
      label = TRUE, label.size = 2.2, pt.size = 0.5
    ) + theme(legend.position = "none")
  }

  if (length(row_p) > 0) {
    summary_rows[[subset_name]] <- Reduce("|", row_p)
  }
}

if (length(summary_rows) > 0) {
  fig_summary <- Reduce("/", summary_rows) +
    plot_annotation(
      title    = "Monaco annotation — all sample filters × all reductions",
      subtitle = "Rows = sample filter · Columns = UMAP reduction (legends hidden for space)",
      theme    = theme(
        plot.title    = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, colour = "grey40")
      )
    )

  save_and_show(fig_summary,
                filename = "summary_monaco_all.png",
                width    = 8 * length(REDUCTIONS),
                height   = 6.5 * length(summary_rows),
                dpi      = 150)
}
```

---

# Output file index

```{r file_index}
files <- list.files("umap_outputs", pattern = "\\.png$", full.names = FALSE)
data.frame(
  File = files,
  Size_MB = round(file.size(file.path("umap_outputs", files)) / 1e6, 2)
) |> knitr::kable(caption = "All saved PNG outputs")
```

---

# Interpretation guide

| Section | What to look for |
|---------|-----------------|
| **Section 1 — Group grids** | Does pool separate clusters before Harmony? Does subject add structure beyond pool? |
| **Section 2 — Per-subject panels** | Is any individual subject an outlier? Do cell-type proportions shift between subjects? |
| **Section 3 — Side-by-side by grouping** | How much does Harmony collapse pool-driven separation while preserving biology? |
| **Section 4 — Monaco summary** | Are Monaco annotations stable across subsets and corrections? |

> **Rule of thumb**: Use **pool-only Harmony** as primary visualisation. Use **subject + pool** as supplementary / QC. Do **not** run DEG or GSEA from these embeddings — use raw-count pseudobulk + limma as before.

---

# Session info

```{r session}
sessionInfo()
```
