---
title: "Per-Cluster DEG (limma + dream) with Optional Batch Correction"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
page-layout: full
---

# Stage 0 — Libraries & Setup

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Matrix)
  library(edgeR)
  library(limma)
  library(variancePartition)  # dream
  library(BiocParallel)       # parallel backend for dream
  library(lme4)               # random effects engine
  library(dplyr)
  library(tidyr)
  library(data.table)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(tibble)
  library(DT)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(ReactomePA)
  library(enrichplot)
  library(glue)
  library(fs)
})

set.seed(42)

## parallel backend (safe defaults across OS)
bp <- if (.Platform$OS.type == "windows") SnowParam(workers = 1, type = "SOCK") else MulticoreParam(workers = max(1L, parallel::detectCores() - 1L))
register(bp)

`%||%` <- function(x, y) if (!is.null(x)) x else y  # small helper used later

logmsg <- function(...) {
  msg <- sprintf("[%s] %s", format(Sys.time(), "%H:%M:%S"), sprintf(...))
  cat(msg, "\n")
  msg
}

stage <- function(n, title) {
  banner <- paste0("\n", strrep("=", 70), "\n",
                   sprintf(" STAGE %d — %s", n, title), "\n",
                   strrep("=", 70), "\n")
  cat(banner)
  invisible(NULL)
}
```

# Stage 1 — Configuration

```{r}
stage(1, "Configuration")

## ---- Inputs ----
##obj_lineage_Non_Haemopoietic_cleaned_withDay.rds
##RDS_FILE    <- "/gpfs/gsfs12/users/cheungf/ACEPD/LINEAGE_DOUBLET_ANALYSIS_PANGI_V9/obj_lineage_Non_Haemopoietic_cleaned_withDay.rds"
RDS_FILE    <- "/gpfs/gsfs12/users/cheungf/ACEPD/LINEAGE_DOUBLET_ANALYSIS_PANGI_V9/obj_lineage_Non_Haemopoietic_cleaned_withDay.rds"
#RDS_FILE    <- "./obj_small_Non_lymph_1pct.rds"
#RDS_FILE    <- "./obj_lineage_Non_lymphoid_cleaned.rds"
#RDS_FILE    <- "../MERGED_PIPELINE_OUTPUT_V5/object_final_annotated_panGI.rds"
##RDS_FILE    <- "./object_final_annotated_panGI_withDay.rds"
#RDS_FILE    <- "./obj_lineage_Lymphoid_cleaned.rds"
#RDS_FILE    <- "./object_final_annotated_panGI_withDay.rds"
#object_final_annotated_panGI_withDay.rds
#RDS_FILE    <- "./obj_lineage_Lymphoid_withDay.rds"

## ---- Metadata columns ----
ASSAY       <- "RNA"

## Two alternative cluster annotations:
#CLUSTER_COL  <- "Monaco_fine"
#CLUSTER_COL2 <- "Monaco_main"
CLUSTER_COL    <- "PanGI_L2"      # Fine-grained cell type annotation
CLUSTER_COL2 <- "PanGI_L1"      # Broad cell lineage annotation

SUBJECT_COL <- "Subject_fixed"
TIME_COL    <- "Treatment"   # e.g., "Treated" / "Untreated"
CONTRAST    <- c("Treated","Untreated")  # c(treat, ref)

## ---- BATCH CORRECTION SETTINGS ----
#BATCH_COL   <- "Lane"        # Column with batch info (e.g., "Lane", "Pool", "Day")
BATCH_COL   <- "Day"  
USE_BATCH   <- TRUE          # Set to TRUE to include batch in models, FALSE to skip
                             # Options: TRUE = test with batch, FALSE = test without batch

## ---- QC Filters (cells) ----
MT_PCT_MAX   <- 65
##MT_PCT_MAX   <- 15 ## HAE
NFEATURE_MIN <- 200
NFEATURE_MAX <- 10000
NCOUNT_MIN   <- 500
NCOUNT_MAX   <- 100000

## ---- Pseudobulk filtering ----
MIN_GENESUM  <- 10L
MIN_SAMPLES  <- 4L       # minimum columns in PB matrix post-filter
USE_QW       <- FALSE    # quality weights for limma-voom (optional)

## ---- GSEA (limma-only) ----
GS_MIN_SIZE <- 10
GS_MAX_SIZE <- 500
FDR_CUTOFF  <- 0.05

## ---- Output root ----
timestamp   <- format(Sys.time(), "%Y%m%d_%H%M%S")
batch_suffix <- if (USE_BATCH) "_WITH_BATCH" else "_NO_BATCH"
OUT_ROOT    <- paste0("OUTPUT_", timestamp, batch_suffix)
dir_create(OUT_ROOT)
logmsg("Output root: %s", OUT_ROOT)
logmsg("Batch correction: %s (Column: %s)", ifelse(USE_BATCH, "ENABLED", "DISABLED"), 
       ifelse(USE_BATCH, BATCH_COL, "N/A"))
```

# Stage 2 — Helpers

```{r}
stage(2, "Helper functions")

sanitize_filename <- function(x) {
  gsub("[/\\\\:*?\"<>|]", "_", x)
}

get_counts <- function(obj, assay = ASSAY) {
  DefaultAssay(obj) <- assay
  lyr <- tryCatch(SeuratObject::Layers(obj[[assay]]), error = function(e) NULL)
  if (!is.null(lyr)) {
    lyrPick <- if ("counts" %in% lyr) "counts" else intersect(c("data","logcounts","normalized"), lyr)[1]
    if (is.na(lyrPick)) stop("No usable layer in assay.")
    m <- SeuratObject::LayerData(obj[[assay]], layer = lyrPick)
  } else {
    m <- GetAssayData(obj, assay = assay, slot = "counts")
  }
  if (!inherits(m, "dgCMatrix")) m <- as(m, "dgCMatrix")
  m
}

pseudobulk_sum <- function(counts, groups) {
  idx <- split(seq_len(ncol(counts)), groups)
  mats <- lapply(idx, function(cols) {
    if (length(cols) == 1) counts[, cols, drop = FALSE] else Matrix::rowSums(counts[, cols, drop = FALSE])
  })
  m <- do.call(cbind, mats)
  colnames(m) <- names(idx)
  as(m, "dgCMatrix")
}

plot_volcano <- function(df, gene_col = "gene", logfc_col = "logFC", p_col = "adj.P.Val",
                         title = "Volcano", out_png = NULL) {
  df <- df %>% dplyr::mutate(sig = !!rlang::sym(p_col) <= 0.05 & abs(!!rlang::sym(logfc_col)) >= 0.25)
  p <- ggplot(df, aes(x = !!rlang::sym(logfc_col), y = -log10(!!rlang::sym(p_col)), label = !!rlang::sym(gene_col))) +
    geom_point(aes(alpha = sig)) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
    geom_vline(xintercept = c(-0.25, 0.25), linetype = 2) +
    geom_hline(yintercept = -log10(0.05), linetype = 2) +
    labs(x = "log2 Fold Change", y = "-log10(adj.P.Val)", title = title) +
    theme_minimal(base_size = 11)
  if (!is.null(out_png)) ggsave(out_png, p, width = 6.5, height = 5, dpi = 300)
  p
}

## Convert limma/dream table -> ranks (ENTREZ, t-stat) for GSEA
make_rank_entrez <- function(limma_df) {
  req <- c("gene","t")
  if (!all(req %in% names(limma_df))) return(NULL)
  tx <- tibble(gene_symbol = limma_df$gene, t = limma_df$t) %>%
    dplyr::filter(is.finite(t), !is.na(gene_symbol), gene_symbol != "")
  idmap <- suppressMessages(
    bitr(unique(tx$gene_symbol), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  )
  tx2 <- tx %>% dplyr::inner_join(idmap, by = c("gene_symbol" = "SYMBOL")) %>%
    dplyr::group_by(ENTREZID) %>%
    dplyr::slice_max(order_by = abs(t), n = 1, with_ties = FALSE) %>%
    dplyr::ungroup()
  ranks <- tx2$t
  names(ranks) <- tx2$ENTREZID
  sort(ranks, decreasing = TRUE)
}

## limma paired WITH OPTIONAL BATCH CORRECTION
run_limma_paired <- function(obj, cluster_name, treat, ref, counts_all, meta_all, cluster_field) {
  cells_cl <- rownames(meta_all)[as.character(meta_all[[cluster_field]]) == as.character(cluster_name)]
  if (!length(cells_cl)) return(list(error = "no_cells_in_cluster"))
  meta_cl <- meta_all[cells_cl, , drop = FALSE]
  meta_cl <- meta_cl[meta_cl[[TIME_COL]] %in% c(ref, treat), , drop = FALSE]
  if (!nrow(meta_cl)) return(list(error = "no_contrast_levels"))

  paired_subj <- meta_cl %>%
    dplyr::distinct(.data[[SUBJECT_COL]], .data[[TIME_COL]]) %>%
    dplyr::count(.data[[SUBJECT_COL]]) %>%
    dplyr::filter(n >= 2) %>% dplyr::pull(.data[[SUBJECT_COL]])

  meta_cl <- meta_cl[meta_cl[[SUBJECT_COL]] %in% paired_subj, , drop = FALSE]
  if (!nrow(meta_cl)) return(list(error = "no_paired_subjects"))

  ## Create pseudobulk key (with or without batch)
  if (USE_BATCH && BATCH_COL %in% colnames(meta_cl)) {
    meta_cl$pb_key <- paste(meta_cl[[SUBJECT_COL]], meta_cl[[TIME_COL]], meta_cl[[BATCH_COL]], sep = "__")
  } else {
    meta_cl$pb_key <- paste(meta_cl[[SUBJECT_COL]], meta_cl[[TIME_COL]], sep = "__")
  }
  
  pb <- pseudobulk_sum(counts_all[, rownames(meta_cl), drop = FALSE], meta_cl$pb_key)

  st <- meta_cl %>%
    dplyr::select(pb_key, dplyr::all_of(SUBJECT_COL), dplyr::all_of(TIME_COL), 
                  dplyr::any_of(BATCH_COL)) %>%
    dplyr::distinct()
  st <- st[match(colnames(pb), st$pb_key), , drop = FALSE]
  rownames(st) <- st$pb_key
  time <- factor(st[[TIME_COL]], levels = c(ref, treat))

  
  # Require ≥2 subjects and both time levels in this cluster
subj_fac <- droplevels(factor(st[[SUBJECT_COL]]))
time_fac <- droplevels(factor(st[[TIME_COL]]))

if (nlevels(subj_fac) < 2) {
  return(list(error = "skip_cluster: <2 subjects after QC/PB"))
}
if (nlevels(time_fac) < 2) {
  return(list(error = "skip_cluster: only one condition present"))
}


  
  ## Design matrix with or without batch
  if (USE_BATCH && BATCH_COL %in% colnames(st)) {
    batch <- factor(st[[BATCH_COL]])
    design <- model.matrix(~ 0 + time + batch)
    colnames(design)[1:nlevels(time)] <- levels(time)
  } else {
    design <- model.matrix(~ 0 + time)
    colnames(design) <- levels(time)
  }

  keep <- filterByExpr(as.matrix(pb), design = design) & (Matrix::rowSums(pb) >= MIN_GENESUM)
  pbk  <- as.matrix(pb[keep, , drop = FALSE])
  if (ncol(pbk) < MIN_SAMPLES) return(list(error = sprintf("few_samples(%d)", ncol(pbk))))
  if (nrow(pbk) < 10) return(list(error = sprintf("few_genes(%d)", nrow(pbk))))

  dge <- DGEList(counts = pbk) |> calcNormFactors(method = "TMM")
  v <- if (USE_QW) voomWithQualityWeights(dge, design, plot = FALSE) else voom(dge, design, plot = FALSE)
  corfit <- duplicateCorrelation(v, design, block = st[[SUBJECT_COL]])
  corv <- corfit$consensus
  if (is.na(corv) || abs(corv) >= 0.99) return(list(error = "bad_correlation"))

  fit  <- lmFit(v, design, block = st[[SUBJECT_COL]], correlation = corv)
  contr_mat <- makeContrasts(contrasts = sprintf("%s-%s", treat, ref), levels = design)
  fit2 <- eBayes(contrasts.fit(fit, contr_mat))
  tt  <- topTable(fit2, number = Inf, sort.by = "none")

  res <- tibble(
    gene = rownames(tt),
    logFC = tt$logFC,
    t = tt$t,
    P.Value = tt$P.Value,
    adj.P.Val = tt$Adj.P.Val %||% tt$adj.P.Val,
    cluster = cluster_name,
    method = ifelse(USE_BATCH, "limma_paired_with_batch", "limma_paired_no_batch"),
    n_subjects = length(unique(st[[SUBJECT_COL]])),
    n_samples = ncol(pbk),
    consensus_cor = corv,
    batch_corrected = USE_BATCH
  )
  list(result = res)
}

## dream paired WITH OPTIONAL BATCH CORRECTION
run_dream_pseudobulk <- function(obj, cluster_name, treat, ref, counts_all, meta_all, cluster_field) {
  cells_cl <- rownames(meta_all)[as.character(meta_all[[cluster_field]]) == as.character(cluster_name)]
  if (!length(cells_cl)) return(list(error = "no_cells_in_cluster"))
  meta_cl <- meta_all[cells_cl, , drop = FALSE]
  meta_cl <- meta_cl[meta_cl[[TIME_COL]] %in% c(ref, treat), , drop = FALSE]
  if (!nrow(meta_cl)) return(list(error = "no_contrast_levels"))

  paired_subj <- meta_cl %>%
    dplyr::distinct(.data[[SUBJECT_COL]], .data[[TIME_COL]]) %>%
    dplyr::count(.data[[SUBJECT_COL]]) %>%
    dplyr::filter(n >= 2) %>% dplyr::pull(.data[[SUBJECT_COL]])
  meta_cl <- meta_cl[meta_cl[[SUBJECT_COL]] %in% paired_subj, , drop = FALSE]
  if (!nrow(meta_cl)) return(list(error = "no_paired_subjects"))

  ## Create pseudobulk key (with or without batch)
  if (USE_BATCH && BATCH_COL %in% colnames(meta_cl)) {
    meta_cl$pb_key <- paste(meta_cl[[SUBJECT_COL]], meta_cl[[TIME_COL]], meta_cl[[BATCH_COL]], sep = "__")
  } else {
    meta_cl$pb_key <- paste(meta_cl[[SUBJECT_COL]], meta_cl[[TIME_COL]], sep = "__")
  }
  
  pb <- pseudobulk_sum(counts_all[, rownames(meta_cl), drop = FALSE], meta_cl$pb_key)

  st <- meta_cl %>%
    dplyr::select(pb_key, dplyr::all_of(SUBJECT_COL), dplyr::all_of(TIME_COL),
                  dplyr::any_of(BATCH_COL)) %>%
    dplyr::distinct()
  st <- st[match(colnames(pb), st$pb_key), , drop = FALSE]
  rownames(st) <- st$pb_key

  time <- factor(st[[TIME_COL]], levels = c(ref, treat))
 
  
    # Require ≥2 subjects and both time levels in this cluster
subj_fac <- droplevels(factor(st[[SUBJECT_COL]]))
time_fac <- droplevels(factor(st[[TIME_COL]]))

if (nlevels(subj_fac) < 2) {
  return(list(error = "skip_cluster: <2 subjects after QC/PB"))
}
if (nlevels(time_fac) < 2) {
  return(list(error = "skip_cluster: only one condition present"))
}



  ## Design matrix with or without batch
  if (USE_BATCH && BATCH_COL %in% colnames(st)) {
    batch <- factor(st[[BATCH_COL]])
    design <- model.matrix(~ 0 + time + batch)
    colnames(design)[1:nlevels(time)] <- levels(time)
  } else {
    design <- model.matrix(~ 0 + time)
    colnames(design) <- levels(time)
  }

  keep <- filterByExpr(as.matrix(pb), design = design) & (Matrix::rowSums(pb) >= MIN_GENESUM)
  pbk  <- as.matrix(pb[keep, , drop = FALSE])
  if (ncol(pbk) < MIN_SAMPLES) return(list(error = sprintf("few_samples(%d)", ncol(pbk))))
  if (nrow(pbk) < 10) return(list(error = sprintf("few_genes(%d)", nrow(pbk))))

  dge <- DGEList(counts = pbk) |> calcNormFactors(method = "TMM")
  v <- voom(dge, design, plot = FALSE)

  ## align metadata to expression samples
  st2 <- st[colnames(v), , drop = FALSE]
  st2$time <- factor(st2[[TIME_COL]], levels = c(ref, treat))
  if (!SUBJECT_COL %in% colnames(st2)) stop("SUBJECT_COL not found in metadata after alignment.")
  st2[[SUBJECT_COL]] <- factor(st2[[SUBJECT_COL]])
  
  ## Build formula with or without batch
  if (USE_BATCH && BATCH_COL %in% colnames(st2)) {
    st2[[BATCH_COL]] <- factor(st2[[BATCH_COL]])
    form <- as.formula(sprintf("~ 0 + time + %s + (1|%s)", BATCH_COL, SUBJECT_COL))
  } else {
    form <- as.formula(sprintf("~ 0 + time + (1|%s)", SUBJECT_COL))
  }

  fit <- tryCatch(variancePartition::dream(v, form, data = st2, BPPARAM = bp),
                  error = function(e) e)
  if (inherits(fit, "error")) return(list(error = paste("dream_fit_failed:", fit$message)))

  coef_names <- colnames(coef(fit))
  L <- makeContrasts(contrasts = sprintf("time%s - time%s", treat, ref), levels = coef_names)
  fit2 <- eBayes(contrasts.fit(fit, L))
  tt <- topTable(fit2, number = Inf, sort.by = "none")

  res <- tibble(
    gene = rownames(tt),
    logFC = tt$logFC,
    t = tt$t,
    P.Value = tt$P.Value,
    adj.P.Val = tt$Adj.P.Val %||% tt$adj.P.Val,
    cluster = cluster_name,
    method = ifelse(USE_BATCH, "dream_with_batch", "dream_no_batch"),
    n_subjects = length(unique(st2[[SUBJECT_COL]])),
    n_samples = ncol(pbk),
    batch_corrected = USE_BATCH
  )
  list(result = res)
}
```

# Stage 3 — Load object & QC

```{r}
stage(3, "Load Seurat object & QC")

logmsg("Loading Seurat object: %s", RDS_FILE)
obj <- readRDS(RDS_FILE)
obj <- tryCatch(JoinLayers(obj, assay = ASSAY), error = function(e) obj)
DefaultAssay(obj) <- ASSAY

## metadata checks (base)
req_cols_base <- c(SUBJECT_COL, TIME_COL)
if (USE_BATCH) req_cols_base <- c(req_cols_base, BATCH_COL)

miss_base <- setdiff(req_cols_base, colnames(obj@meta.data))
if (length(miss_base)) stop("Missing required metadata columns: ", paste(miss_base, collapse=", "))

## Check batch column structure if using batch correction
if (USE_BATCH) {
  batch_summary <- table(obj@meta.data[[BATCH_COL]], obj@meta.data[[TIME_COL]])
  logmsg("Batch structure:")
  print(batch_summary)
}

## we will run on any cluster columns that exist among the two
candidate_cluster_cols <- c(CLUSTER_COL, CLUSTER_COL2)
have_cluster_cols <- candidate_cluster_cols[candidate_cluster_cols %in% colnames(obj@meta.data)]
if (!length(have_cluster_cols)) stop("None of the cluster columns found: ", paste(candidate_cluster_cols, collapse=", "))

## counts / metadata
counts_all <- get_counts(obj, assay = ASSAY)
meta_all   <- obj@meta.data

## fill common QC fields if missing
if (!"percent.mt" %in% colnames(meta_all)) {
  mt_genes <- grep("^MT-|^Mt-|^mt-", rownames(counts_all), value = TRUE)
  mt_counts <- if (length(mt_genes)) Matrix::colSums(counts_all[mt_genes, , drop = FALSE]) else rep(0, ncol(counts_all))
  lib_counts <- Matrix::colSums(counts_all)
  meta_all$percent.mt <- ifelse(lib_counts > 0, 100 * mt_counts / lib_counts, 0)
}
if (!"nFeature_RNA" %in% colnames(meta_all)) meta_all$nFeature_RNA <- Matrix::colSums(counts_all > 0)
if (!"nCount_RNA"   %in% colnames(meta_all)) meta_all$nCount_RNA   <- Matrix::colSums(counts_all)

keep_qc <- with(meta_all,
  percent.mt <= MT_PCT_MAX &
  nFeature_RNA >= NFEATURE_MIN & nFeature_RNA <= NFEATURE_MAX &
  nCount_RNA   >= NCOUNT_MIN   & nCount_RNA   <= NCOUNT_MAX
)
kept_cells <- rownames(meta_all)[keep_qc]
logmsg("QC kept %d/%d cells (%.1f%%)", length(kept_cells), ncol(counts_all),
       100 * length(kept_cells) / max(1, ncol(counts_all)))

obj      <- subset(obj, cells = kept_cells)
meta_all <- meta_all[kept_cells, , drop = FALSE]

treat <- CONTRAST[1]; ref <- CONTRAST[2]
contrast_name <- paste0(treat, "_vs_", ref)

logmsg("Contrast: %s", contrast_name)
logmsg("Cluster annotations detected: %s", paste(have_cluster_cols, collapse = ", "))
logmsg("Batch correction: %s", ifelse(USE_BATCH, paste("YES, using", BATCH_COL), "NO"))
```

# Stage 4 — DEG (limma + dream) and GSEA (limma only) for each annotation

```{r}
stage(4, "Run per-cluster DEG and GSEA for each annotation")

run_for_annotation <- function(cluster_field, ann_label) {
  logmsg("=== Annotation: %s (field: %s) ===", ann_label, cluster_field)

  ## Per-annotation output dirs
  DIR_ANN_ROOT <- file.path(OUT_ROOT, ann_label)
  DIR_DEG  <- file.path(DIR_ANN_ROOT, "deg_results")
  DIR_PLOT <- file.path(DIR_ANN_ROOT, "plots")
  DIR_GSEA <- file.path(DIR_ANN_ROOT, "gsea_results")
  dir_create(c(DIR_ANN_ROOT, DIR_DEG, DIR_PLOT, DIR_GSEA))

  ## Cluster levels for this annotation
  clusters <- sort(unique(as.character(meta_all[[cluster_field]])))
  logmsg("Clusters (%s): %s", ann_label, paste(clusters, collapse = ", "))

  all_deg_limma <- list()
  all_deg_dream <- list()
  all_gsea_go   <- list()
  all_gsea_re   <- list()

  for (cl in clusters) {
    logmsg("[%s] Cluster %s — start", ann_label, cl)
    cl_safe <- sanitize_filename(cl)

    ## limma paired
    lim <- run_limma_paired(obj, cl, treat, ref, counts_all, meta_all, cluster_field)
    if (!is.null(lim$error)) {
      logmsg("  limma SKIP: %s", lim$error)
    } else {
      df <- lim$result
      all_deg_limma[[cl]] <- df
      fwrite(df, file.path(DIR_DEG, glue("{ann_label}__cluster_{cl_safe}_limma_{contrast_name}.csv")))
    #  plot_volcano(
    #    df,
    #    title = glue("Volcano (limma) — {ann_label} — {cl} — {contrast_name}"),
    #    out_png = file.path(DIR_PLOT, glue("{ann_label}__cluster_{cl_safe}_limma_volcano_{contrast_name}.png"))
    #  )

      ## limma-only GSEA
      ranks <- make_rank_entrez(df)
      if (!is.null(ranks) && length(ranks) >= 20) {
        ## GO:BP
        ego <- tryCatch(
          gseGO(geneList = ranks, OrgDb = org.Hs.eg.db, ont = "BP",
                minGSSize = GS_MIN_SIZE, maxGSSize = GS_MAX_SIZE,
                pvalueCutoff = 1, verbose = FALSE),
          error = function(e) NULL
        )
        if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
          ego_sym <- tryCatch(
            clusterProfiler::setReadable(ego, OrgDb = org.Hs.eg.db, keyType = "ENTREZID"),
            error = function(e) ego
          )
          gdf_raw <- as.data.frame(ego)
          gdf_sym <- as.data.frame(ego_sym)
          gdf <- gdf_sym %>%
            dplyr::rename(leading_edge_symbols = core_enrichment) %>%
            dplyr::mutate(
              leading_edge_entrez = gdf_raw$core_enrichment,
              cluster = cl,
              geneset = "GO:BP",
              annotation = ann_label
            )
          all_gsea_go[[cl]] <- gdf
          fwrite(gdf, file.path(DIR_GSEA, glue("{ann_label}__cluster_{cl_safe}_GSEA_GO_{contrast_name}.csv")))
        #  p_go <- suppressMessages(
        #    enrichplot::dotplot(ego_sym, showCategory = 20) + ggtitle(glue("GSEA GO:BP — {ann_label} — {cl}"))
         # )
        #  ggsave(file.path(DIR_PLOT, glue("{ann_label}__cluster_{cl_safe}_GSEA_GO_{contrast_name}.png")),
        #         p_go, width = 7.5, height = 5.5, dpi = 300)
        } else {
          logmsg("  GSEA GO:BP — no enrichments")
        }

        ## Reactome
        ere <- tryCatch(
          gsePathway(geneList = ranks, organism = "human",
                     minGSSize = GS_MIN_SIZE, maxGSSize = GS_MAX_SIZE,
                     pvalueCutoff = 1, verbose = FALSE),
          error = function(e) NULL
        )
        if (!is.null(ere) && nrow(as.data.frame(ere)) > 0) {
          ere_sym <- tryCatch(
            clusterProfiler::setReadable(ere, OrgDb = org.Hs.eg.db, keyType = "ENTREZID"),
            error = function(e) ere
          )
          rdf_raw <- as.data.frame(ere)
          rdf_sym <- as.data.frame(ere_sym)
          rdf <- rdf_sym %>%
            dplyr::rename(leading_edge_symbols = core_enrichment) %>%
            dplyr::mutate(
              leading_edge_entrez = rdf_raw$core_enrichment,
              cluster = cl,
              geneset = "REACTOME",
              annotation = ann_label
            )
          all_gsea_re[[cl]] <- rdf
          fwrite(rdf, file.path(DIR_GSEA, glue("{ann_label}__cluster_{cl_safe}_GSEA_REACTOME_{contrast_name}.csv")))
      #    p_re <- suppressMessages(
      #      enrichplot::dotplot(ere_sym, showCategory = 20) + ggtitle(glue("GSEA Reactome — {ann_label} — {cl}"))
      #    )
      #    ggsave(file.path(DIR_PLOT, glue("{ann_label}__cluster_{cl_safe}_GSEA_REACTOME_{contrast_name}.png")),
      #           p_re, width = 7.5, height = 5.5, dpi = 300)
        } else {
          logmsg("  GSEA Reactome — no enrichments")
        }
      } else {
        logmsg("  GSEA SKIP: insufficient ranked genes")
      }
    }

    ## dream paired
    dr <- run_dream_pseudobulk(obj, cl, treat, ref, counts_all, meta_all, cluster_field)
    if (!is.null(dr$error)) {
      logmsg("  dream SKIP: %s", dr$error)
    } else {
      dfd <- dr$result
      all_deg_dream[[cl]] <- dfd
      fwrite(dfd, file.path(DIR_DEG, glue("{ann_label}__cluster_{cl_safe}_dream_{contrast_name}.csv")))
  #    plot_volcano(
  #      dfd,
  #      title = glue("Volcano (dream) — {ann_label} — {cl} — {contrast_name}"),
  #      out_png = file.path(DIR_PLOT, glue("{ann_label}__cluster_{cl_safe}_dream_volcano_{contrast_name}.png"))
  #    )
    }

    logmsg("[%s] Cluster %s — done", ann_label, cl)
  }

  ## Combined CSVs for this annotation
  if (length(all_deg_limma)) fwrite(bind_rows(all_deg_limma), file.path(DIR_DEG, glue("{ann_label}__ALL_clusters_limma.csv")))
  if (length(all_deg_dream)) fwrite(bind_rows(all_deg_dream), file.path(DIR_DEG, glue("{ann_label}__ALL_clusters_dream.csv")))
  if (length(all_gsea_go))   fwrite(bind_rows(all_gsea_go),   file.path(DIR_GSEA, glue("{ann_label}__ALL_clusters_GSEA_GO.csv")))
  if (length(all_gsea_re))   fwrite(bind_rows(all_gsea_re),   file.path(DIR_GSEA, glue("{ann_label}__ALL_clusters_GSEA_REACTOME.csv")))
  invisible(NULL)
}

## Run for each annotation present
for (ann in have_cluster_cols) {
  run_for_annotation(ann, ann)
}

logmsg("Stage 4 complete")
```

# Stage 5 — Batch Effect Diagnostics

```{r}
stage(5, "Batch Effect Diagnostics")

## This section provides diagnostics to help decide if batch correction was necessary

if (USE_BATCH && BATCH_COL %in% colnames(obj@meta.data)) {
  
  logmsg("Generating batch effect diagnostics...")
  
  ## 1. Variance Partition Analysis (if variancePartition is available)
  tryCatch({
    # Select a representative cluster for variance partition
    representative_cluster <- names(sort(table(obj@meta.data[[have_cluster_cols[1]]]), decreasing = TRUE))[1]
    cells_rep <- rownames(obj@meta.data)[obj@meta.data[[have_cluster_cols[1]]] == representative_cluster]
    
    if (length(cells_rep) > 100) {
      meta_rep <- obj@meta.data[cells_rep, ]
      meta_rep <- meta_rep[meta_rep[[TIME_COL]] %in% CONTRAST, ]
      
      # Create pseudobulk for variance partition
      meta_rep$vp_key <- paste(meta_rep[[SUBJECT_COL]], meta_rep[[TIME_COL]], meta_rep[[BATCH_COL]], sep = "__")
      pb_vp <- pseudobulk_sum(counts_all[, rownames(meta_rep), drop = FALSE], meta_rep$vp_key)
      
      st_vp <- meta_rep %>%
        dplyr::select(vp_key, dplyr::all_of(c(SUBJECT_COL, TIME_COL, BATCH_COL))) %>%
        dplyr::distinct()
      st_vp <- st_vp[match(colnames(pb_vp), st_vp$vp_key), ]
      rownames(st_vp) <- st_vp$vp_key
      
      # Normalize
      dge_vp <- DGEList(counts = as.matrix(pb_vp))
      dge_vp <- calcNormFactors(dge_vp, method = "TMM")
      
      # Simple design for variance partition
      form_vp <- as.formula(sprintf("~ (1|%s) + (1|%s) + (1|%s)", SUBJECT_COL, TIME_COL, BATCH_COL))
      
      # Calculate variance partition (use top 500 variable genes for speed)
      gene_vars <- apply(cpm(dge_vp, log = TRUE), 1, var)
      top_genes <- names(sort(gene_vars, decreasing = TRUE))[1:min(500, length(gene_vars))]
      
      vp <- fitExtractVarPartModel(
        exprObj = cpm(dge_vp[top_genes, ], log = TRUE),
        formula = form_vp,
        data = st_vp
      )
      
      # Plot variance partition
      vp_plot <- plotVarPart(vp) + 
        ggtitle(sprintf("Variance Explained by Each Factor\nCluster: %s (top 500 genes)", representative_cluster)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
      vp_plot_file <- file.path(OUT_ROOT, "variance_partition_analysis.png")
      ggsave(vp_plot_file, vp_plot, width = 8, height = 6, dpi = 300)
      logmsg("Variance partition plot saved: %s", vp_plot_file)
      
      # Print summary
      vp_summary <- as.data.frame(sort(colMeans(vp), decreasing = TRUE))
      colnames(vp_summary) <- "Mean_Variance_Explained"
      vp_summary$Percentage <- round(100 * vp_summary$Mean_Variance_Explained, 2)
      
      cat("\n=== Variance Partition Summary ===\n")
      print(vp_summary)
      cat("\nInterpretation:\n")
      cat("- If 'Batch' explains <5% variance: batch correction likely unnecessary\n")
      cat("- If 'Batch' explains 5-15% variance: batch correction may be beneficial\n")
      cat("- If 'Batch' explains >15% variance: batch correction strongly recommended\n\n")
    }
  }, error = function(e) {
    logmsg("Variance partition analysis failed: %s", e$message)
  })
  
  ## 2. PCA analysis colored by batch
  if ("pca" %in% names(obj@reductions)) {
    pca_batch <- DimPlot(obj, reduction = "pca", group.by = BATCH_COL, dims = c(1, 2)) +
      ggtitle("PCA colored by Batch") +
      theme_minimal()
    
    pca_treatment <- DimPlot(obj, reduction = "pca", group.by = TIME_COL, dims = c(1, 2)) +
      ggtitle("PCA colored by Treatment") +
      theme_minimal()
    
    pca_combined <- pca_batch | pca_treatment
    pca_file <- file.path(OUT_ROOT, "pca_batch_vs_treatment.png")
    ggsave(pca_file, pca_combined, width = 14, height = 6, dpi = 300)
    logmsg("PCA comparison saved: %s", pca_file)
  }
  
  ## 3. Summary table of samples per batch
  batch_summary_df <- obj@meta.data %>%
    dplyr::group_by(!!rlang::sym(BATCH_COL), !!rlang::sym(TIME_COL), !!rlang::sym(SUBJECT_COL)) %>%
    dplyr::summarise(n_cells = n(), .groups = "drop") %>%
    dplyr::arrange(!!rlang::sym(BATCH_COL), !!rlang::sym(SUBJECT_COL), !!rlang::sym(TIME_COL))
  
  cat("\n=== Batch × Subject × Treatment Design ===\n")
  print(as.data.frame(batch_summary_df))
  
  fwrite(batch_summary_df, file.path(OUT_ROOT, "batch_design_summary.csv"))
  
} else {
  logmsg("Batch correction disabled or batch column not found. Skipping diagnostics.")
}
```

# Stage 6 — Comparison: With vs Without Batch Correction

```{r}
stage(6, "Comparison Summary")

cat("\n=== ANALYSIS SETTINGS ===\n")
cat(sprintf("Batch correction: %s\n", ifelse(USE_BATCH, "ENABLED", "DISABLED")))
if (USE_BATCH) {
  cat(sprintf("Batch column: %s\n", BATCH_COL))
  cat(sprintf("Output directory suffix: %s\n", batch_suffix))
}
cat(sprintf("\nTo compare WITH vs WITHOUT batch correction:\n"))
cat(sprintf("1. Run this script with USE_BATCH = TRUE\n"))
cat(sprintf("2. Run this script with USE_BATCH = FALSE\n"))
cat(sprintf("3. Compare the DEG results in the two OUTPUT directories\n"))
cat(sprintf("4. Look for:\n"))
cat(sprintf("   - Number of DEGs (should be similar if no batch effects)\n"))
cat(sprintf("   - Top genes (should largely overlap)\n"))
cat(sprintf("   - Effect sizes (logFC should be similar)\n"))
cat(sprintf("   - Variance partition results (Batch < 5%% = no correction needed)\n\n"))

## Create a quick comparison helper script
comparison_script <- sprintf('
# Quick comparison between batch-corrected and non-batch-corrected results
# Load both result files and compare

library(data.table)
library(dplyr)

# Adjust these paths based on your actual output directories
with_batch <- fread("OUTPUT_*_WITH_BATCH/%s/deg_results/%s__ALL_clusters_limma.csv")
no_batch   <- fread("OUTPUT_*_NO_BATCH/%s/deg_results/%s__ALL_clusters_limma.csv")

# Compare significant genes
with_batch_sig <- with_batch %>% filter(adj.P.Val < 0.05, abs(logFC) > 0.5)
no_batch_sig   <- no_batch %>% filter(adj.P.Val < 0.05, abs(logFC) > 0.5)

cat("\\nNumber of significant DEGs:\\n")
cat(sprintf("With batch correction: %%d\\n", nrow(with_batch_sig)))
cat(sprintf("Without batch correction: %%d\\n", nrow(no_batch_sig)))

# Overlap
overlap_genes <- intersect(with_batch_sig$gene, no_batch_sig$gene)
cat(sprintf("\\nOverlapping significant genes: %%d (%%0.1f%%%%)\\n", 
            length(overlap_genes),
            100 * length(overlap_genes) / max(nrow(with_batch_sig), nrow(no_batch_sig))))

# Correlation of logFC for same genes
merged <- inner_join(
  with_batch %>% select(gene, cluster, logFC_with = logFC, padj_with = adj.P.Val),
  no_batch %>% select(gene, cluster, logFC_no = logFC, padj_no = adj.P.Val),
  by = c("gene", "cluster")
)

cat(sprintf("\\nCorrelation of logFC values: %%0.3f\\n", cor(merged$logFC_with, merged$logFC_no)))
cat(sprintf("Median absolute difference in logFC: %%0.3f\\n", median(abs(merged$logFC_with - merged$logFC_no))))
',
have_cluster_cols[1], have_cluster_cols[1],
have_cluster_cols[1], have_cluster_cols[1])

comparison_file <- file.path(OUT_ROOT, "COMPARE_BATCH_CORRECTION.R")
writeLines(comparison_script, comparison_file)
logmsg("Comparison script saved: %s", comparison_file)
```

# Stage 7 — Results summary & downloads

```{r}
stage(7, "Summaries & Downloads")

## Directory layout
cat(paste0(OUT_ROOT, "/\n"))
for (ann in have_cluster_cols) {
  cat(paste0(" ├─ ", ann, "/\n",
             " │   ├─ deg_results/\n",
             " │   ├─ gsea_results/\n",
             " │   └─ plots/\n"))
}

## Helper to show a combined table if files exist
show_table_if_exists <- function(path_glob, caption_text) {
  files <- fs::dir_ls(OUT_ROOT, recurse = TRUE, type = "file", glob = path_glob)
  if (!length(files)) return(invisible(NULL))
  dt <- rbindlist(lapply(files, fread), fill = TRUE)
  datatable(
    dt,
    extensions = c("Buttons","Scroller"),
    options = list(
      dom = "Bfrtip",
      buttons = c("copy","csv","excel"),
      scrollX = TRUE, deferRender = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25
    ),
    caption = htmltools::tags$caption(style = "caption-side: top; text-align: left;", caption_text)
  )
}

batch_label <- ifelse(USE_BATCH, " (WITH batch correction)", " (NO batch correction)")

show_table_if_exists("**/deg_results/*ALL_clusters_limma.csv",        
                     paste0("All clusters — limma DEGs", batch_label))
show_table_if_exists("**/deg_results/*ALL_clusters_dream.csv",        
                     paste0("All clusters — dream DEGs", batch_label))
show_table_if_exists("**/gsea_results/*ALL_clusters_GSEA_GO.csv",     
                     paste0("All clusters — GSEA GO:BP (limma only)", batch_label))
show_table_if_exists("**/gsea_results/*ALL_clusters_GSEA_REACTOME.csv",
                     paste0("All clusters — GSEA Reactome (limma only)", batch_label))

## Quick links to all per-cluster files
files_tbl <- tibble(
  file = fs::dir_ls(OUT_ROOT, recurse = TRUE, type = "file", glob = "*.{csv,png}")
) %>%
  dplyr::mutate(
    annotation = stringr::str_match(string = file, pattern = paste0(OUT_ROOT, "/([^/]+)/"))[,2],
    kind = dplyr::case_when(
      str_detect(file, "deg_results") & str_detect(file, "dream_") ~ "DEG: dream CSV",
      str_detect(file, "deg_results") & str_detect(file, "limma_") ~ "DEG: limma CSV",
      str_detect(file, "GSEA_GO") ~ "GSEA GO:BP CSV",
      str_detect(file, "GSEA_REACTOME") ~ "GSEA Reactome CSV",
      str_detect(file, "dream_volcano") ~ "Plot: dream volcano",
      str_detect(file, "limma_volcano") ~ "Plot: limma volcano",
      str_detect(file, "GSEA_GO") & str_detect(file, "plots") ~ "Plot: GSEA GO:BP",
      str_detect(file, "GSEA_REACTOME") & str_detect(file, "plots") ~ "Plot: GSEA Reactome",
      str_detect(file, "variance_partition") ~ "Diagnostic: Variance Partition",
      str_detect(file, "pca_batch") ~ "Diagnostic: PCA Batch vs Treatment",
      str_detect(file, "batch_design") ~ "Diagnostic: Batch Design",
      TRUE ~ "Other"
    ),
    cluster = stringr::str_match(basename(file), "cluster_([^_]+)_")[,2] %||% "ALL",
    link = paste0("<a href='", file, "' download>", basename(file), "</a>")
  ) %>%
  dplyr::arrange(annotation, kind, cluster, file)

datatable(
  files_tbl %>% dplyr::select(annotation, kind, cluster, link),
  escape = FALSE,
  extensions = c("Buttons"),
  options = list(dom = "Bfrtip", buttons = c("copy","csv","excel"),
                 pageLength = 25, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left;",
    paste0("All output files", batch_label)
  )
)
```

# Stage 8 — Recommendations

```{r}
stage(8, "Batch Correction Recommendations")

cat("\n=== BATCH CORRECTION DECISION GUIDE ===\n\n")

cat("Based on your experimental design:\n")
cat("- Samples were POOLED before library prep\n")
cat("- Demultiplexed by SNPs (computational, not technical separation)\n")
cat("- Each lane contains multiple subjects\n")
cat("- UMAP shows no obvious batch clustering\n\n")

cat("RECOMMENDATION: Batch correction is likely UNNECESSARY because:\n")
cat("1. Physical pooling eliminates most technical batch effects\n")
cat("2. All cells in a pool experienced identical chemistry/sequencing\n")
cat("3. The 'Lane' variable represents biological replicates, not technical batches\n")
cat("4. Your paired design (~ Subject + Treatment) already accounts for the main source of variability\n\n")

cat("SUGGESTED WORKFLOW:\n")
cat("1. Use the NO BATCH CORRECTION results as your primary analysis\n")
cat("2. Check the variance partition plot (if generated):\n")
cat("   - If Lane explains <5%% variance → definitely skip batch correction\n")
cat("   - If Lane explains >15%% variance → might indicate real lane effects\n\n")

cat("3. Compare both analyses if uncertain:\n")
cat("   - Similar number of DEGs → batch correction doesn't matter much\n")
cat("   - Very different results → investigate why (may indicate overfitting)\n\n")

cat("WHEN TO USE BATCH CORRECTION:\n")
cat("- Samples processed on different days/weeks\n")
cat("- Different technicians or protocols\n")
cat("- Different sequencing runs with separate library prep\n")
cat("- Clear batch structure visible in UMAP/PCA\n")
cat("- Variance partition shows >10%% variance from batch\n\n")

if (USE_BATCH) {
  cat("CURRENT RUN: Using batch correction\n")
  cat("To test WITHOUT batch correction, set USE_BATCH = FALSE and re-run\n\n")
} else {
  cat("CURRENT RUN: NOT using batch correction (recommended for your design)\n")
  cat("To test WITH batch correction, set USE_BATCH = TRUE and re-run\n\n")
}
```

---

**Analysis complete!** Output saved to: `r OUT_ROOT`

**Batch Correction Setting:** `r ifelse(USE_BATCH, "ENABLED", "DISABLED")`