---
title: "Single-Cell RNA-seq Analysis: Treatment Response"
subtitle: "Paired Before/After Analysis with Differential Expression and Pathway Enrichment"
author: "Your Name"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 8
    smooth-scroll: true
    number-sections: true
execute:
  warning: false
  message: false
  cache: false
params:
  config_file: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
set.seed(42)
```

```{r load_libraries, include=FALSE}
suppressPackageStartupMessages({
  library(yaml); library(Seurat); library(SeuratObject); library(Matrix)
  library(dplyr); library(tidyr); library(data.table); library(tibble); library(stringr); library(purrr)
  library(ggplot2); library(patchwork); library(scales); library(ggrepel); library(pheatmap)
  library(RColorBrewer); library(DT); library(knitr); library(plotly); library(EnhancedVolcano)
  library(enrichplot); library(glue); library(fs); library(rstatix)
  library(logger); library(progress); library(future); library(future.apply)
  library(scProportionTest, lib = "../../lib")
})
source("scRNAseq_helpers.R")
source("enhanced_helpers.R")
```

```{r load_config}
config <- yaml::read_yaml(params$config_file)
#config <- yaml::read_yaml("./config4.yaml")
#config <- yaml::read_yaml("./config5C.yaml")
RDS_FILE <- config$input$rds_file
OUT_ROOT <- config$output$root_dir
REUSE_RESULTS_ONLY <- config$output$use_cached_results
ASSAY <- config$metadata$assay
CLUSTER_COL <- config$metadata$cluster_fine
CLUSTER_COL_L1 <- config$metadata$cluster_broad
SUBJECT_COL <- config$metadata$subject_col
TIME_COL <- config$metadata$treatment_col
BATCH_COL <- config$metadata$batch_col
CONTRAST <- c(config$groups$treatment, config$groups$reference)
MT_PCT_MAX <- config$qc$mt_pct_max
NFEATURE_MIN <- config$qc$nfeature_min
NFEATURE_MAX <- config$qc$nfeature_max
NCOUNT_MIN <- config$qc$ncount_min
NCOUNT_MAX <- config$qc$ncount_max
FDR_CUTOFF <- config$analysis$fdr_cutoff
LOGFC_CUTOFF <- config$analysis$logfc_cutoff
N_PERMUTATIONS <- config$analysis$n_permutations
MIN_CELLS <- config$analysis$min_cells_per_cluster

# Extract custom text sections (if provided)
CUSTOM_TEXT <- if (!is.null(config$custom_text)) config$custom_text else list()
get_custom_text <- function(section) {
  if (!is.null(CUSTOM_TEXT[[section]]) && nchar(CUSTOM_TEXT[[section]]) > 0) {
    return(CUSTOM_TEXT[[section]])
  }
  return(NULL)
}

DIR_DEG_FINE <- file.path(OUT_ROOT, CLUSTER_COL, "deg_results")
DIR_GSEA_FINE <- file.path(OUT_ROOT, CLUSTER_COL, "gsea_results")
DIR_PLOTS_FINE <- file.path(OUT_ROOT, CLUSTER_COL, "plots")
DIR_DEG_MAIN <- file.path(OUT_ROOT, CLUSTER_COL_L1, "deg_results")
DIR_GSEA_MAIN <- file.path(OUT_ROOT, CLUSTER_COL_L1, "gsea_results")
DIR_PLOTS_MAIN <- file.path(OUT_ROOT, CLUSTER_COL_L1, "plots")
DIR_EXPORTS <- file.path(OUT_ROOT, "exports")
DIR_SIGNATURES <- file.path(OUT_ROOT, "signature_scores")

walk(c(DIR_DEG_FINE, DIR_GSEA_FINE, DIR_PLOTS_FINE, DIR_DEG_MAIN, DIR_GSEA_MAIN, 
       DIR_PLOTS_MAIN, DIR_EXPORTS, DIR_SIGNATURES), ~dir_create(.x, recurse = TRUE))

limma_fine_path <- file.path(DIR_DEG_FINE, sprintf("%s__ALL_clusters_limma.csv", CLUSTER_COL))
limma_main_path <- file.path(DIR_DEG_MAIN, sprintf("%s__ALL_clusters_limma.csv", CLUSTER_COL_L1))
go_fine_path <- file.path(DIR_GSEA_FINE, sprintf("%s__ALL_clusters_GSEA_GO.csv", CLUSTER_COL))
go_main_path <- file.path(DIR_GSEA_MAIN, sprintf("%s__ALL_clusters_GSEA_GO.csv", CLUSTER_COL_L1))
re_fine_path <- file.path(DIR_GSEA_FINE, sprintf("%s__ALL_clusters_GSEA_REACTOME.csv", CLUSTER_COL))
re_main_path <- file.path(DIR_GSEA_MAIN, sprintf("%s__ALL_clusters_GSEA_REACTOME.csv", CLUSTER_COL_L1))

limma_fine_available <- file_exists(limma_fine_path)
limma_main_available <- file_exists(limma_main_path)
gsea_fine_available <- file_exists(go_fine_path) || file_exists(re_fine_path)
gsea_main_available <- file_exists(go_main_path) || file_exists(re_main_path)

if (limma_main_available) limma_main <- fread(limma_main_path)
if (limma_fine_available) limma_fine <- fread(limma_fine_path)
if (file_exists(re_fine_path)) gsea_fine_all <- fread(re_fine_path)
if (file_exists(re_main_path)) gsea_main_all <- fread(re_main_path)

message(sprintf("Configuration loaded: %s (fine) and %s (broad)", CLUSTER_COL, CLUSTER_COL_L1))
message(sprintf("Comparison: %s vs %s", CONTRAST[1], CONTRAST[2]))
```

# Executive Summary {.unnumbered}

```{r exec_summary, include=FALSE}
obj <- readRDS(RDS_FILE)
obj <- tryCatch(JoinLayers(obj, assay = ASSAY), error = function(e) obj)
DefaultAssay(obj) <- ASSAY
n_subjects <- length(unique(obj@meta.data[[SUBJECT_COL]]))
n_cells <- ncol(obj)
n_genes <- nrow(obj)
n_clusters_fine <- length(unique(obj@meta.data[[CLUSTER_COL]]))
n_clusters_broad <- length(unique(obj@meta.data[[CLUSTER_COL_L1]]))
if (limma_fine_available) {
  n_sig_genes <- limma_fine %>% filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>% nrow()
  n_sig_clusters <- limma_fine %>% filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>% 
    pull(cluster) %>% unique() %>% length()
} else {
  n_sig_genes <- NA; n_sig_clusters <- NA
}
```

::: {.callout-note icon=false}

## Analysis Overview

**Study Design:** Paired analysis comparing **`r CONTRAST[1]`** vs **`r CONTRAST[2]`**

**Key Metrics:**

* Subjects: `r n_subjects` paired samples
* Cells: `r comma(n_cells)` (post-QC)
* Genes: `r comma(n_genes)`
* Cell Types: `r n_clusters_broad` broad, `r n_clusters_fine` fine
* Significant DEGs: `r if(!is.na(n_sig_genes)) comma(n_sig_genes) else "Pending"` genes across `r if(!is.na(n_sig_clusters)) n_sig_clusters else "?"` cell types
:::

```{r introduction_section, results='asis', echo=FALSE}
intro_text <- get_custom_text("introduction")
if (!is.null(intro_text)) {
  cat("\n## Introduction {.unnumbered}\n\n")
  cat(intro_text, "\n\n")
}
```

```{r objectives_section, results='asis', echo=FALSE}
objectives_text <- get_custom_text("objectives")
if (!is.null(objectives_text)) {
  cat("\n## Study Objectives {.unnumbered}\n\n")
  cat(objectives_text, "\n\n")
}
```

# Data Loading & QC

```{r qc_context_section, results='asis', echo=FALSE}
qc_context <- get_custom_text("qc_context")
if (!is.null(qc_context)) {
  cat(qc_context, "\n\n")
}
```

```{r load_data}
message("Loading and processing data...")
req_cols <- c(CLUSTER_COL, CLUSTER_COL_L1, SUBJECT_COL, TIME_COL)
if (!is.null(BATCH_COL)) req_cols <- c(req_cols, BATCH_COL)
miss <- setdiff(req_cols, colnames(obj@meta.data))
if (length(miss)) stop(sprintf("Missing columns: %s", paste(miss, collapse=', ')))

has_counts <- tryCatch("counts" %in% Layers(obj[[ASSAY]]), error = function(e) FALSE)
counts_all <- if (has_counts) GetAssayData(obj, assay = ASSAY, slot = "counts") else GetAssayData(obj, assay = ASSAY, slot = "data")

meta_all <- obj@meta.data
if (!"percent.mito" %in% colnames(meta_all)) {
  mt_genes <- grep("^MT-|^Mt-|^mt-", rownames(counts_all), value = TRUE)
  mt_counts <- if (length(mt_genes)) Matrix::colSums(counts_all[mt_genes, , drop = FALSE]) else rep(0, ncol(counts_all))
  lib_counts <- Matrix::colSums(counts_all)
  meta_all$percent.mito <- ifelse(lib_counts > 0, 100 * mt_counts / lib_counts, 0)
}
if (!"nFeature_RNA" %in% colnames(meta_all)) meta_all$nFeature_RNA <- Matrix::colSums(counts_all > 0)
if (!"nCount_RNA" %in% colnames(meta_all)) meta_all$nCount_RNA <- Matrix::colSums(counts_all)
obj@meta.data <- meta_all
obj <- calculate_complexity(obj)
qc_failure_analysis <- analyze_qc_failures(obj, config$qc)

keep_qc <- with(obj@meta.data, percent.mito <= MT_PCT_MAX & nFeature_RNA >= NFEATURE_MIN & 
                nFeature_RNA <= NFEATURE_MAX & nCount_RNA >= NCOUNT_MIN & nCount_RNA <= NCOUNT_MAX)
obj <- subset(obj, cells = rownames(obj@meta.data)[keep_qc])

meta_all <- obj@meta.data
meta_all$SubjectTimePoint <- paste(meta_all[[SUBJECT_COL]], meta_all[[TIME_COL]], sep = "_")
vec <- meta_all$SubjectTimePoint
parsed <- utils::strcapture(sprintf("^(.+?)_(%s|%s)$", CONTRAST[1], CONTRAST[2]), vec,
                            proto = list(subj = character(), trt = character()))
subjects <- sort(unique(parsed$subj[!is.na(parsed$subj)]))
ordered_levels <- as.character(unlist(lapply(subjects, function(s) c(paste0(s, "_", CONTRAST[2]), paste0(s, "_", CONTRAST[1])))))
ordered_levels <- c(ordered_levels, setdiff(unique(vec), ordered_levels))
meta_all$SubjectTimePoint_ordered <- factor(vec, levels = ordered_levels)
obj@meta.data <- meta_all
message(sprintf("QC complete: %s cells retained", ncol(obj)))
```

::: {.callout-success}

## Dataset Summary

* **Cells:** `r comma(ncol(obj))` high-quality cells from `r n_subjects` subjects
* **QC Filters:** Mito ≤ `r MT_PCT_MAX`%, Genes `r NFEATURE_MIN`-`r NFEATURE_MAX`, UMIs `r comma(NCOUNT_MIN)`-`r comma(NCOUNT_MAX)`
:::

## QC Failure Summary

```{r qc_table}
datatable(qc_failure_analysis$summary %>% pivot_longer(everything(), names_to = "Metric", values_to = "Count"),
  options = list(pageLength = 10, dom = 't'), caption = "QC Filter Statistics", rownames = FALSE) %>%
  formatRound("Count", digits = 0)
```

## Interactive QC Plot

```{r qc_plot, fig.width=10, fig.height=8}
plot_qc_interactive(obj, color_by = TIME_COL, annotation_col = CLUSTER_COL)
```

## Sample QC Report

```{r sample_qc}
sample_report <- generate_sample_report(obj, SUBJECT_COL, TIME_COL)
datatable(sample_report, extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 20, scrollX = TRUE),
  caption = "Per-Sample Quality Metrics", rownames = FALSE) %>%
  formatRound(c("median_UMI", "median_genes", "median_mito", "median_ribo", "mean_novelty"), digits = 2) %>%
  formatStyle("outlier", backgroundColor = styleEqual(c(TRUE, FALSE), c("#fee5d9", "white")))
```

```{r qc_notes_section, results='asis', echo=FALSE}
qc_notes <- get_custom_text("qc_notes")
if (!is.null(qc_notes)) {
  cat("\n## QC Notes\n\n")
  cat(qc_notes, "\n\n")
}
```

# UMAP Visualizations

## UMAPs by Metadata

```{r umap_metadata, results='asis', fig.width=12, fig.height=10}
generate_umap_tabset(obj, priority_cols = c(TIME_COL, "SubjectTimePoint_ordered", CLUSTER_COL, CLUSTER_COL_L1),
  exclude_cols = c("orig.ident", "RNA_snn_res.0.8"), max_tabs = 6)
```

## Cell Type Annotation

```{r umap_celltypes, fig.width=14, fig.height=12}
p1 <- DimPlot(obj, reduction = "umap", group.by = CLUSTER_COL, label = TRUE, repel = TRUE, pt.size = 1) +
  labs(title = sprintf("Fine: %s (%d types)", CLUSTER_COL, n_clusters_fine)) + theme_minimal()
p2 <- DimPlot(obj, reduction = "umap", group.by = CLUSTER_COL_L1, label = TRUE, repel = TRUE, pt.size = 1) +
  labs(title = sprintf("Broad: %s (%d types)", CLUSTER_COL_L1, n_clusters_broad)) + theme_minimal()
p1 / p2
```

# Cell Type Annotation Summary

## Cell Counts by Annotation Level

```{r annotation_table}
count_L1 <- table(obj@meta.data[[CLUSTER_COL_L1]])
count_L2 <- table(obj@meta.data[[CLUSTER_COL]])

anno_summary <- bind_rows(
  tibble(Level = CLUSTER_COL_L1, 
         `Cell Type` = names(count_L1), 
         `N Cells` = as.integer(count_L1),
         `% of Total` = 100 * count_L1 / ncol(obj)),
  tibble(Level = CLUSTER_COL, 
         `Cell Type` = names(count_L2), 
         `N Cells` = as.integer(count_L2),
         `% of Total` = 100 * count_L2 / ncol(obj))
) %>%
  arrange(Level, desc(`N Cells`))

datatable(anno_summary,
          options = list(pageLength = 15, scrollX = TRUE),
          caption = sprintf("**Table 2:** Cell type distribution across annotation levels (%s and %s)", 
                           CLUSTER_COL_L1, CLUSTER_COL)) %>%
  formatPercentage("% of Total", digits = 2) %>%
  formatRound("N Cells", digits = 0)
```

## Cell Type Distribution Across Subjects and Timepoints

```{r annotation_heatmap, fig.width=12, fig.height=8}
counts_df <- as.data.frame(
  table(
    Cell_Type = obj@meta.data[[CLUSTER_COL_L1]],
    SubjectTP = obj@meta.data$SubjectTimePoint_ordered
  )
) %>% 
  dplyr::filter(Freq > 0)

ggplot(counts_df, aes(x = SubjectTP, y = Cell_Type, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), size = 3, color = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", 
                      name = "Cell Count", trans = "log10",
                      labels = comma) +
  labs(
    title = "Cell Type Distribution Across Samples",
    subtitle = sprintf("%s (broad lineages) by Subject and Timepoint", CLUSTER_COL_L1),
    x = "Subject / Timepoint",
    y = sprintf("Cell Type (%s)", CLUSTER_COL_L1)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r annotation_heatmap2, fig.width=12, fig.height=8}
counts_df <- as.data.frame(
  table(
    Cell_Type = obj@meta.data[[CLUSTER_COL]],
    SubjectTP = obj@meta.data$SubjectTimePoint_ordered
  )
) %>% 
  dplyr::filter(Freq > 0)

ggplot(counts_df, aes(x = SubjectTP, y = Cell_Type, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), size = 3, color = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", 
                      name = "Cell Count", trans = "log10",
                      labels = comma) +
  labs(
    title = "Cell Type Distribution Across Samples",
    subtitle = sprintf("%s (fine lineages) by Subject and Timepoint", CLUSTER_COL),
    x = "Subject / Timepoint",
    y = sprintf("Cell Type (%s)", CLUSTER_COL)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

# Composition Analysis

```{r composition}
if (config$report$include_composition_section) {
  prop_data_fine <- calculate_proportions(obj, cluster_col = CLUSTER_COL, subject_col = SUBJECT_COL, condition_col = TIME_COL)
  prop_data_broad <- calculate_proportions(obj, cluster_col = CLUSTER_COL_L1, subject_col = SUBJECT_COL, condition_col = TIME_COL)
}
```

## Paired Proportions (Fine)

```{r comp_fine, fig.width=14, fig.height=10}
if (config$report$include_composition_section) {
  plot_paired_proportions(prop_data_fine, cluster_col = CLUSTER_COL, subject_col = SUBJECT_COL, 
    condition_col = TIME_COL, ncol = 4)
}
```

## Paired Proportions (Broad)

```{r comp_broad, fig.width=14, fig.height=8}
if (config$report$include_composition_section) {
  plot_paired_proportions(prop_data_broad, cluster_col = CLUSTER_COL_L1, subject_col = SUBJECT_COL, 
    condition_col = TIME_COL, ncol = 3)
}
```

## Statistical Testing

```{r prop_test, fig.width=10, fig.height=8}
# ---------- Fine (CLUSTER_COL) ----------
proptest_rds <- file.path(OUT_ROOT, "proportion_test_fine.rds")
if (REUSE_RESULTS_ONLY && file_exists(proptest_rds)) {
  prop_test_results <- readRDS(proptest_rds)
} else if (requireNamespace("scProportionTest", quietly = TRUE)) {
  prop_test <- sc_utils(obj)
  prop_test_results <- permutation_test(
    prop_test,
    cluster_identity = CLUSTER_COL,
    sample_identity  = TIME_COL,
    sample_1 = CONTRAST[2],  # reference
    sample_2 = CONTRAST[1],  # treatment
    n_permutations = N_PERMUTATIONS
  )
  saveRDS(prop_test_results, proptest_rds)
}
if (exists("prop_test_results")) {
  permutation_plot(prop_test_results) +
    labs(
      title    = sprintf("Proportion Changes: %s (Fine)", CLUSTER_COL),
      subtitle = sprintf("%s permutations", scales::comma(N_PERMUTATIONS))
    )
}

# ---------- Broad (CLUSTER_COL_L1) ----------
proptest_broad_rds <- file.path(OUT_ROOT, "proportion_test_broad.rds")
if (REUSE_RESULTS_ONLY && file_exists(proptest_broad_rds)) {
  prop_test_results_broad <- readRDS(proptest_broad_rds)
} else if (requireNamespace("scProportionTest", quietly = TRUE)) {
  prop_test_broad <- sc_utils(obj)
  prop_test_results_broad <- permutation_test(
    prop_test_broad,
    cluster_identity = CLUSTER_COL_L1,
    sample_identity  = TIME_COL,
    sample_1 = CONTRAST[2],  # reference
    sample_2 = CONTRAST[1],  # treatment
    n_permutations = N_PERMUTATIONS
  )
  saveRDS(prop_test_results_broad, proptest_broad_rds)
}
if (exists("prop_test_results_broad")) {
  permutation_plot(prop_test_results_broad) +
    labs(
      title    = sprintf("Proportion Changes: %s (Broad)", CLUSTER_COL_L1),
      subtitle = sprintf("%s permutations", scales::comma(N_PERMUTATIONS))
    )
}
```

# Differential Expression

```{r deg_context_section, results='asis', echo=FALSE}
deg_context <- get_custom_text("deg_context")
if (!is.null(deg_context)) {
  cat(deg_context, "\n\n")
}
```

```{r load_deg}
if (config$report$include_deg_section) {
  if (limma_fine_available) message(sprintf("Loaded %s DEG results (fine)", nrow(limma_fine)))
  if (limma_main_available) message(sprintf("Loaded %s DEG results (broad)", nrow(limma_main)))
}
```

## DEG Summary (Fine)

```{r deg_fine_summary, fig.width=12, fig.height=6}
if (limma_fine_available) {
  deg_summary_fine <- limma_fine %>% filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
    mutate(direction = factor(ifelse(logFC > 0, "Up", "Down"), levels = c("Down", "Up"))) %>% count(cluster, direction)
  ggplot(deg_summary_fine, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) + geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white", fontface = "bold") +
    coord_flip() + labs(title = sprintf("DEGs per Cell Type (%s)", CLUSTER_COL), x = "Cell Type", y = "DEGs") +
    theme_minimal() + theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## DEG Summary (Broad)

```{r deg_broad_summary, fig.width=12, fig.height=6}
if (limma_main_available) {
  deg_summary_broad <- limma_main %>%
    dplyr::filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>%
    dplyr::mutate(direction = factor(ifelse(logFC > 0, "Up", "Down"),
                                     levels = c("Down", "Up"))) %>%
    dplyr::count(cluster, direction)

  ggplot(deg_summary_broad, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n),
              position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold") +
    coord_flip() +
    labs(
      title = sprintf("DEGs per Cell Type (%s, Broad)", CLUSTER_COL_L1),
      x = "Cell Type", y = "DEGs"
    ) +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## DEG Table (Fine)

```{r deg_fine_table}
if (limma_fine_available) {
  datatable(limma_fine %>%
    dplyr::filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>% 
      select(gene, logFC, t, P.Value, adj.P.Val, cluster, method),
    extensions = c("Buttons", "Scroller"),
    options = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"), scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption = sprintf("%s DEG Results", CLUSTER_COL), filter = "top", rownames = FALSE) %>%
    formatRound(c("logFC", "t"), 3) %>% formatSignif(c("P.Value", "adj.P.Val"), 3)
}
```

## DEG Table (Broad)

```{r deg_broad_table}
if (limma_main_available) {
  DT::datatable(
    limma_main %>%
    dplyr::filter(adj.P.Val <= FDR_CUTOFF, abs(logFC) >= LOGFC_CUTOFF) %>% 
      dplyr::select(gene, logFC, t, P.Value, adj.P.Val, cluster, method),
    extensions = c("Buttons", "Scroller"),
    options = list(
      dom = "Bfrtip",
      buttons = list("copy", "csv", "excel"),
      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25
    ),
    caption = sprintf("%s DEG Results (Broad)", CLUSTER_COL_L1),
    filter = "top", rownames = FALSE
  ) %>%
    formatRound(c("logFC", "t"), 3) %>%
    formatSignif(c("P.Value", "adj.P.Val"), 3)
}
```

# GSEA

```{r gsea_context_section, results='asis', echo=FALSE}
gsea_context <- get_custom_text("gsea_context")
if (!is.null(gsea_context)) {
  cat(gsea_context, "\n\n")
}
```

```{r load_gsea}
if (config$report$include_gsea_section) {
  if (gsea_fine_available) {
    gsea_fine_all <- NULL
    if (file_exists(go_fine_path)) gsea_fine_all <- fread(go_fine_path) %>% mutate(geneset = "GO:BP")
    if (file_exists(re_fine_path)) {
      re <- fread(re_fine_path) %>% mutate(geneset = "REACTOME")
      gsea_fine_all <- if (is.null(gsea_fine_all)) re else bind_rows(gsea_fine_all, re)
    }
  }
  if (gsea_main_available) {
    gsea_main_all <- NULL
    if (file_exists(go_main_path)) gsea_main_all <- fread(go_main_path) %>% mutate(geneset = "GO:BP")
    if (file_exists(re_main_path)) {
      re <- fread(re_main_path) %>% mutate(geneset = "REACTOME")
      gsea_main_all <- if (is.null(gsea_main_all)) re else bind_rows(gsea_main_all, re)
    }
  }
}
```

## GSEA Summary (Fine)

```{r gsea_fine_summary, fig.width=12, fig.height=8}
if (gsea_fine_available && !is.null(gsea_fine_all)) {
  gsea_counts <- gsea_fine_all %>% mutate(direction = case_when(p.adjust <= FDR_CUTOFF & NES > 0 ~ "Up",
    p.adjust <= FDR_CUTOFF & NES < 0 ~ "Down", TRUE ~ NA_character_)) %>% filter(!is.na(direction)) %>%
    count(cluster, geneset, direction) %>% mutate(direction = factor(direction, levels = c("Down", "Up")))
  ggplot(gsea_counts, aes(x = reorder(cluster, n), y = n, fill = direction)) + geom_col(width = 0.7) +
    geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 3) +
    facet_wrap(~ geneset, ncol = 1, scales = "free_y") + coord_flip() +
    labs(title = sprintf("Pathways per Cell Type (%s)", CLUSTER_COL), x = "Cell Type", y = "Pathways") +
    theme_minimal() + theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## GSEA Summary (Broad)

```{r gsea_broad_summary, fig.width=12, fig.height=8}
if (gsea_main_available && !is.null(gsea_main_all)) {
  gsea_counts_broad <- gsea_main_all %>%
    dplyr::mutate(
      direction = dplyr::case_when(
        p.adjust <= FDR_CUTOFF & NES > 0  ~ "Up",
        p.adjust <= FDR_CUTOFF & NES < 0 ~ "Down",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(direction)) %>%
    dplyr::count(cluster, geneset, direction) %>%
    dplyr::mutate(direction = factor(direction, levels = c("Down", "Up")))

  ggplot(gsea_counts_broad, aes(x = reorder(cluster, n), y = n, fill = direction)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n),
              position = position_stack(vjust = 0.5),
              color = "white", fontface = "bold", size = 3) +
    facet_wrap(~ geneset, ncol = 1, scales = "free_y") +
    coord_flip() +
    labs(
      title = sprintf("Pathways per Cell Type (%s, Broad)", CLUSTER_COL_L1),
      x = "Cell Type", y = "Pathways"
    ) +
    theme_minimal() +
    theme(legend.position = "top", panel.grid.major.y = element_blank())
}
```

## GSEA Table (Fine)

```{r gsea_fine_table}
if (gsea_fine_available && !is.null(gsea_fine_all)) {
  datatable(gsea_fine_all %>%  filter(p.adjust <= FDR_CUTOFF)  %>%
              select(cluster, geneset, Description, NES, pvalue, p.adjust, setSize),
    extensions = c("Buttons", "Scroller"),
    options = list(dom = "Bfrtip", buttons = list("copy", "csv", "excel"), scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25),
    caption = sprintf("%s GSEA Results", CLUSTER_COL), filter = "top", rownames = FALSE) %>%
    formatRound("NES", 3) %>% formatSignif(c("pvalue", "p.adjust"), 3)
}
```

## GSEA Table (Broad)

```{r gsea_broad_table}
if (gsea_main_available && !is.null(gsea_main_all)) {
  DT::datatable(
    gsea_main_all %>%  filter(p.adjust <= FDR_CUTOFF) %>% 
      dplyr::select(cluster, geneset, Description, NES, pvalue, p.adjust, setSize),
    extensions = c("Buttons", "Scroller"),
    options = list(
      dom = "Bfrtip",
      buttons = list("copy", "csv", "excel"),
      scrollX = TRUE, scrollY = 500, scroller = TRUE, pageLength = 25
    ),
    caption = sprintf("%s GSEA Results (Broad)", CLUSTER_COL_L1),
    filter = "top", rownames = FALSE
  ) %>%
    formatRound("NES", 3) %>%
    formatSignif(c("pvalue", "p.adjust"), 3)
}
```

# Most Variable Pathways

```{r gsea_main_var_heatmap, fig.width=12, fig.height=8}
if (gsea_main_available) {
  # Select top variable pathways
  top_var_pathways_main <- gsea_main_all %>%
    dplyr::filter(p.adjust <= FDR_CUTOFF) %>%
    dplyr::group_by(ID, Description) %>%
    dplyr::summarise(
      var_NES = var(NES, na.rm = TRUE),
      mean_abs_NES = mean(abs(NES), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::slice_max(order_by = var_NES, n = 20) %>%
    dplyr::pull(ID)
  
  heatmap_data_main <- gsea_main_all %>%
    dplyr::filter(ID %in% top_var_pathways_main, p.adjust <= FDR_CUTOFF) %>%
    dplyr::mutate(
      Description_short = ifelse(nchar(Description) > 50,
                                 paste0(substr(Description, 1, 47), "..."),
                                 Description)
    )
  
  if (nrow(heatmap_data_main) > 0) {
    ggplot(heatmap_data_main, aes(x = cluster, y = Description_short, fill = NES)) +
      geom_tile(color = "white", linewidth = 0.5) +
      geom_text(aes(label = ifelse(p.adjust < 0.01, "**", 
                                   ifelse(p.adjust < 0.05, "*", ""))),
                size = 4, color = "black") +
      scale_fill_gradient2(low = "#4DBBD5", mid = "white", high = "#E64B35",
                          midpoint = 0, name = "NES") +
      facet_wrap(~ geneset, scales = "free_y", ncol = 1) +
      labs(
        title = sprintf("Top Variable Pathways Across Cell Types (%s)", CLUSTER_COL_L1),
        subtitle = "* p.adj < 0.05, ** p.adj < 0.01",
        x = "Cell Type",
        y = "Pathway"
      ) +
      theme_minimal(base_size = 10) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "grey90"),
        plot.title = element_text(face = "bold")
      )
  }
}
```  


```{r gsea_fine_varheatmap, fig.width=14, fig.height=10}
if (gsea_fine_available) {
  # Select top variable pathways
  top_var_pathways_fine <- gsea_fine_all %>%
    dplyr::filter(p.adjust <= FDR_CUTOFF) %>%
    dplyr::group_by(ID, Description) %>%
    dplyr::summarise(
      var_NES = var(NES, na.rm = TRUE),
      mean_abs_NES = mean(abs(NES), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::slice_max(order_by = var_NES, n = 30) %>%
    dplyr::pull(ID)
  
  heatmap_data_fine <- gsea_fine_all %>%
    dplyr::filter(ID %in% top_var_pathways_fine, p.adjust <= FDR_CUTOFF) %>%
    dplyr::mutate(
      Description_short = ifelse(nchar(Description) > 50,
                                 paste0(substr(Description, 1, 47), "..."),
                                 Description)
    )
  
  if (nrow(heatmap_data_fine) > 0) {
    ggplot(heatmap_data_fine, aes(x = cluster, y = Description_short, fill = NES)) +
      geom_tile(color = "white", linewidth = 0.5) +
      geom_text(aes(label = ifelse(p.adjust < 0.01, "**", 
                                   ifelse(p.adjust < 0.05, "*", ""))),
                size = 4, color = "black") +
      scale_fill_gradient2(low = "#4DBBD5", mid = "white", high = "#E64B35",
                          midpoint = 0, name = "NES") +
      facet_wrap(~ geneset, scales = "free_y", ncol = 1) +
      labs(
        title = sprintf("Top Variable Pathways Across Cell Types (%s)", CLUSTER_COL),
        subtitle = "* p.adj < 0.05, ** p.adj < 0.01",
        x = "Cell Type",
        y = "Pathway"
      ) +
      theme_minimal(base_size = 10) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "grey90"),
        plot.title = element_text(face = "bold")
      )
  }
}
```


# Marker Genes

```{r markers, fig.width=14, fig.height=16}
if (config$report$include_marker_section) {
  all_markers <- unique(unlist(config$markers[!names(config$markers) %in% "custom_genes"]))
  available_markers <- intersect(all_markers, rownames(obj))
  if (length(available_markers) > 0) {
    FeaturePlot(obj, features = head(available_markers, 16), ncol = 4, pt.size = 1, order = TRUE) &
      theme_minimal(base_size = 10)
  }
}
```

# Methods

::: {.callout-note icon=false}

## Computational Methods

**QC:** Mito ≤ `r MT_PCT_MAX`%, Genes `r NFEATURE_MIN`-`r NFEATURE_MAX`, UMIs `r comma(NCOUNT_MIN)`-`r comma(NCOUNT_MAX)`

**DE Analysis:** limma pseudobulk, paired design, FDR ≤ `r FDR_CUTOFF`, |log₂FC| ≥ `r LOGFC_CUTOFF`

**GSEA:** Gene Set Enrichment Analysis, FDR ≤ `r FDR_CUTOFF`

**Software:** R `r R.version.string`, Seurat `r packageVersion("Seurat")`, limma `r packageVersion("limma")`
:::

```{r methods_additional_section, results='asis', echo=FALSE}
methods_additional <- get_custom_text("methods_additional")
if (!is.null(methods_additional)) {
  cat("\n## Additional Methods\n\n")
  cat(methods_additional, "\n\n")
}
```

```{r discussion_section, results='asis', echo=FALSE}
discussion <- get_custom_text("discussion")
if (!is.null(discussion)) {
  cat("\n# Discussion {.unnumbered}\n\n")
  cat(discussion, "\n\n")
}
```

```{r conclusions_section, results='asis', echo=FALSE}
conclusions <- get_custom_text("conclusions")
if (!is.null(conclusions)) {
  cat("\n# Conclusions {.unnumbered}\n\n")
  cat(conclusions, "\n\n")
}
```

```{r future_directions_section, results='asis', echo=FALSE}
future_directions <- get_custom_text("future_directions")
if (!is.null(future_directions)) {
  cat("\n# Future Directions {.unnumbered}\n\n")
  cat(future_directions, "\n\n")
}
```

# Session Info

```{r session}
sessionInfo()
```

---

**Report generated:** `r Sys.time()` | **Config:** `r params$config_file`

```
