---
title: "CellChat CD4-CD8 Crosstalk Analysis"
subtitle: "Bidirectional T Cell Communication: Treated vs Untreated"
author: "Generated with CellChat v2.1.1"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: 
      light: flatly
      dark: darkly
    css: styles.css
    embed-resources: true
    smooth-scroll: true
    link-external-newwindow: true
execute:
  warning: false
  message: false
  cache: false
  echo: false
---

```{css, echo=FALSE}
/* Custom styling */
.callout-note {
  border-left: 4px solid #4A90E2;
  background-color: #f0f7ff;
}

.callout-important {
  border-left: 4px solid #E74C3C;
  background-color: #fff5f5;
}

.callout-tip {
  border-left: 4px solid #2ECC71;
  background-color: #f0fff4;
}

.callout-warning {
  border-left: 4px solid #F39C12;
  background-color: #fffbf0;
}

.metric-card {
  background: linear-gradient(135deg, #667eea 0%, #764ba2 100%);
  color: white;
  padding: 20px;
  border-radius: 8px;
  margin: 10px 0;
  box-shadow: 0 4px 6px rgba(0,0,0,0.1);
}

.figure-caption {
  font-style: italic;
  color: #666;
  margin-top: 10px;
}

h1 { 
  color: #2c3e50;
  border-bottom: 3px solid #3498db;
  padding-bottom: 10px;
}

h2 { 
  color: #34495e;
  margin-top: 30px;
  border-bottom: 2px solid #95a5a6;
  padding-bottom: 5px;
}

h3 { 
  color: #7f8c8d;
}

.panel-tabset > .panel-tabs {
  border-bottom: 2px solid #3498db;
}

.dataTables_wrapper {
  font-size: 0.9em;
}

table {
  border-collapse: collapse;
  width: 100%;
  margin: 20px 0;
}

th {
  background-color: #3498db;
  color: white;
  padding: 12px;
  text-align: left;
}

tr:nth-child(even) {
  background-color: #f8f9fa;
}

pre {
  background-color: #f6f8fa;
  border: 1px solid #e1e4e8;
  border-radius: 6px;
  padding: 16px;
}

.side-by-side {
  display: flex;
  justify-content: space-around;
  gap: 2%;
}

.side-by-side > div {
  width: 48%;
}
```

::: {.callout-note icon=false}
## Analysis Overview

This analysis identifies **bidirectional communication between CD4 and CD8 T cells** by analyzing ligand-receptor interactions. Understanding how these cells communicate is essential for deciphering T cell coordination and how treatment affects their crosstalk.

**Report Structure:**
* **Section 1:** Global signaling context across all cell types
* **Section 2:** CD4 → CD8 signaling (CD4 cells sending signals TO CD8 cells)
* **Section 3:** CD8 → CD4 signaling (CD8 cells sending signals TO CD4 cells)
:::

---

::: {.callout-tip icon=false}
## Key Terminology

* **Ligand-receptor pairs:** Specific molecular interactions where one cell expresses a ligand and another expresses its receptor
* **Pathways:** Groups of related ligand-receptor pairs (e.g., TNF, MIF, CLEC)
* **Significant:** Statistical confidence (p < 0.05) that the interaction is real
* **Treated vs Untreated:** How treatment changes communication patterns
* **Probability:** CellChat's measure of interaction strength based on gene expression
:::

---

## Setup

```{r setup}
#| code-summary: "Load libraries and configure paths"

suppressPackageStartupMessages({
  library(CellChat)
  library(patchwork)
  library(ComplexHeatmap)
  library(circlize)
  library(dplyr)
  library(data.table)
  library(fs)
  library(ggplot2)
  library(DT)
})

OUTPUT_DIR <- file.path(
  "/Users/cheungf/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/R/23_301",
  "OUTPUT_CELLCHAT_23_JAN_26"
)

CD4_LABEL <- "Conventional_CD4"
CD8_LABEL <- "Conventional_CD8"

PVALUE_CUTOFF <- 0.05
TOP_PATHWAYS  <- 12

DOC_DIR <- "/Users/cheungf/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/R/23_301"
PLOT_DIR_REL  <- "plots_cd4_cd8"
TABLE_DIR_REL <- "tables_cd4_cd8"
PLOT_DIR  <- file.path(DOC_DIR,  PLOT_DIR_REL)
TABLE_DIR <- file.path(DOC_DIR, TABLE_DIR_REL)
dir_create(PLOT_DIR)
dir_create(TABLE_DIR)

logmsg <- function(...) cat(sprintf("[%s] %s\n", format(Sys.time(), "%H:%M:%S"), sprintf(...)))

conds <- c("Treated","Untreated")

cellchat_list <- list(
  Treated   = readRDS(file.path(OUTPUT_DIR, "cellchat_Treated_FINAL.rds")),
  Untreated = readRDS(file.path(OUTPUT_DIR, "cellchat_Untreated_FINAL.rds"))
)
cellchat_list <- cellchat_list[conds]
stopifnot(all(conds %in% names(cellchat_list)))

cellchat_merged <- readRDS(file.path(OUTPUT_DIR, "cellchat_merged_WITH_Step10_ALL.rds"))

logmsg("✓ Loaded Treated/Untreated FINAL + merged Step10 RDS.")
```

---

## Helper Functions

```{r helpers}
#| code-summary: "Utility functions for plotting and analysis"

save_png_safe <- function(file, width=1800, height=1200, res=200, expr) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  ok <- TRUE
  error_msg <- NULL
  png(filename=file, width=width, height=height, res=res)
  tryCatch(
    expr,
    error = function(e) { 
      ok <<- FALSE
      error_msg <<- e$message
      message("⚠️ PLOT FAILED: ", basename(file), " | ERROR: ", e$message) 
    }
  )
  dev.off()

  if (!ok || !file.exists(file) || file.info(file)$size == 0) {
    if (file.exists(file)) unlink(file)
    if (!is.null(error_msg)) {
      message("  Full error: ", error_msg)
    }
    return(NA_character_)
  }
  file
}

assert_celltypes <- function(cc, labels) {
  lv <- levels(cc@idents)
  missing <- setdiff(labels, lv)
  if (length(missing) > 0) {
    stop("Missing celltype labels in CellChat object: ", paste(missing, collapse=", "),
         "\nAvailable levels include: ", paste(head(lv, 40), collapse=", "), if (length(lv)>40) " ...")
  }
}

get_comm <- function(cc, source, target, p.cut=0.05) {
  df <- suppressWarnings(subsetCommunication(cc, sources.use=source, targets.use=target))
  if (is.null(df) || nrow(df) == 0) return(list(all=data.frame(), sig=data.frame()))

  df <- as.data.frame(df)
  if (!("pval" %in% colnames(df))) {
    pcol <- grep("^p", colnames(df), value=TRUE)[1]
    if (!is.na(pcol)) df$pval <- df[[pcol]]
  }

  df <- df %>%
    mutate(pval = as.numeric(pval)) %>%
    filter(!is.na(pval)) %>%
    arrange(pval)

  list(all=df, sig=df %>% filter(pval < p.cut))
}

rank_pathways <- function(df_sig) {
  if (is.null(df_sig) || nrow(df_sig) == 0) return(data.frame())
  pn <- if ("pathway_name" %in% colnames(df_sig)) "pathway_name" else if ("pathway" %in% colnames(df_sig)) "pathway" else NULL
  if (is.null(pn)) return(data.frame())

  df_sig %>%
    group_by(.data[[pn]]) %>%
    summarise(
      n_pairs = dplyr::n(),
      sum_prob = sum(prob, na.rm=TRUE),
      .groups="drop"
    ) %>%
    rename(pathway = .data[[pn]]) %>%
    arrange(desc(sum_prob), desc(n_pairs))
}

bubble_compare <- function(merged, source, target, out_file, title_text) {
  save_png_safe(out_file, width=2200, height=1200, res=200, expr={
    p <- netVisual_bubble(
      merged,
      sources.use = source,
      targets.use = target,
      comparison = c(1,2),
      angle.x = 45,
      remove.isolate = TRUE
    ) + ggtitle(title_text)
    print(p)
  })
}

mini_circle <- function(cc, out_file, title_text, measure=c("count","weight")) {
  measure <- match.arg(measure)

  mat <- cc@net[[measure]]
  if (is.null(rownames(mat)) || is.null(colnames(mat))) {
    stop("cc@net[[measure]] has no row/col names. Need named cell types to subset.")
  }

  cells <- c(CD4_LABEL, CD8_LABEL)
  missing <- setdiff(cells, rownames(mat))
  if (length(missing) > 0) {
    stop("Missing in net matrix: ", paste(missing, collapse=", "),
         "\nAvailable: ", paste(rownames(mat), collapse=", "))
  }

  mat2 <- mat[cells, cells, drop = FALSE]
  vw <- as.numeric(table(cc@idents)[cells])
  names(vw) <- cells

  save_png_safe(out_file, width=1600, height=1600, res=200, expr={
    netVisual_circle(
      mat2,
      vertex.weight = vw,
      weight.scale = TRUE,
      label.edge = FALSE,
      title.name = title_text
    )
  })
}

make_chords <- function(cc, pathways, prefix, source_label, target_label) {
  out_files <- c()
  valid_pw <- cc@netP$pathways
  pathways2 <- intersect(pathways, valid_pw)

  dropped <- setdiff(pathways, pathways2)
  if (length(dropped) > 0) {
    message(prefix, " dropped (not in cc@netP$pathways): ", paste(dropped, collapse=", "))
  }

  for (pw in pathways2) {
    out <- file.path(PLOT_DIR, sprintf("chord_%s_%s.png", prefix, gsub("[^A-Za-z0-9]+","_", pw)))
    made <- save_png_safe(out, width=1200, height=1200, res=200, expr={
      netVisual_chord_gene(
        cc,
        sources.use = source_label,
        targets.use = target_label,
        signaling = pw,
        lab.cex = 0.6,
        legend.pos.x = 10
      )
      title(paste0(prefix, ": ", pw))
    })
    if (!is.na(made)) out_files <- c(out_files, made)
  }

  out_files
}

make_heatmaps <- function(merged, pathways, prefix, source_label, target_label) {
  out_files <- c()
  for (pw in pathways) {
    out <- file.path(PLOT_DIR, sprintf("heatmap_%s_%s.png", prefix, gsub("[^A-Za-z0-9]+","_", pw)))
    made <- save_png_safe(out, width=4000, height=2400, res=200, expr={
      netVisual_heatmap(
        merged,
        signaling = pw,
        sources.use = source_label,
        targets.use = target_label,
        comparison = c(1,2),
        measure = "weight"
      )
      title(paste0(prefix, ": ", pw))
    })
    if (!is.na(made)) out_files <- c(out_files, made)
  }
  out_files
}
```

---

## Validation

```{r validate}
#| code-summary: "Verify CD4/CD8 cell types exist in data"

assert_celltypes(cellchat_list$Treated,   c(CD4_LABEL, CD8_LABEL))
assert_celltypes(cellchat_list$Untreated, c(CD4_LABEL, CD8_LABEL))
logmsg("✓ Found CD4/CD8 labels in both conditions: %s and %s", CD4_LABEL, CD8_LABEL)
```

---

# 1. Global Signaling Context

## 1.1 Pathway Contribution Heatmaps

::: {.callout-note icon=false}
### What This Shows

Which signaling pathways are most active across **all cell types** in your sample. This provides context for understanding CD4-CD8 specific interactions.

**How to Read:**
- Each row = signaling pathway (e.g., TNF, MHC-I, CLEC)
- Darker/larger bars = stronger signaling activity
- **Outgoing** = cells sending signals out
- **Incoming** = cells receiving signals
- Compare Treated vs Untreated to identify treatment effects
:::

```{r signaling-heatmaps}
f_out_t <- file.path(PLOT_DIR, "signaling_outgoing_treated.png")
f_out_u <- file.path(PLOT_DIR, "signaling_outgoing_untreated.png")
f_inc_t <- file.path(PLOT_DIR, "signaling_incoming_treated.png")
f_inc_u <- file.path(PLOT_DIR, "signaling_incoming_untreated.png")

made_out_t <- save_png_safe(f_out_t, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Treated, pattern = "outgoing", width = 10, height = 20)
  draw(ht)
})

made_out_u <- save_png_safe(f_out_u, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Untreated, pattern = "outgoing", width = 10, height = 20)
  draw(ht)
})

made_inc_t <- save_png_safe(f_inc_t, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Treated, pattern = "incoming", width = 10, height = 20)
  draw(ht)
})

made_inc_u <- save_png_safe(f_inc_u, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Untreated, pattern = "incoming", width = 10, height = 20)
  draw(ht)
})
```

### Outgoing Signaling Patterns

::: {.side-by-side}

::: {}
**Treated**

```{r}
#| out-width: "100%"
if (!is.na(made_out_t) && file.exists(made_out_t)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(made_out_t)))
}
```
:::

::: {}
**Untreated**

```{r}
#| out-width: "100%"
if (!is.na(made_out_u) && file.exists(made_out_u)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(made_out_u)))
}
```
:::

:::

### Incoming Signaling Patterns

::: {.side-by-side}

::: {}
**Treated**

```{r}
#| out-width: "100%"
if (!is.na(made_inc_t) && file.exists(made_inc_t)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(made_inc_t)))
}
```
:::

::: {}
**Untreated**

```{r}
#| out-width: "100%"
if (!is.na(made_inc_u) && file.exists(made_inc_u)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(made_inc_u)))
}
```
:::

:::

---

## 1.2 CD4-CD8 Interaction Strengths

::: {.callout-tip icon=false}
### What This Shows

Direct communication strength between **only CD4 and CD8 cells** (filtering out all other cell types).

**How to Read:**
- Lines show communication between CD4 and CD8
- Thicker lines = stronger communication
- **Counts** = number of different signaling interactions
- **Weights** = overall strength (accounts for number and intensity)
- Compare across tabs to see treatment effects on CD4-CD8 crosstalk
:::

```{r cd4cd8-circles}
f1a <- file.path(PLOT_DIR, "cd4cd8_circle_Treated_count.png")
f1b <- file.path(PLOT_DIR, "cd4cd8_circle_Treated_weight.png")
f1c <- file.path(PLOT_DIR, "cd4cd8_circle_Untreated_count.png")
f1d <- file.path(PLOT_DIR, "cd4cd8_circle_Untreated_weight.png")

mini_circle(cellchat_list$Treated,   f1a, "Treated (Counts)",  "count")
mini_circle(cellchat_list$Treated,   f1b, "Treated (Weights)", "weight")
mini_circle(cellchat_list$Untreated, f1c, "Untreated (Counts)","count")
mini_circle(cellchat_list$Untreated, f1d, "Untreated (Weights)","weight")
```

::: {.panel-tabset}

#### Treated - Counts

```{r}
if (file.exists(f1a)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f1a)))
```

#### Treated - Weights

```{r}
if (file.exists(f1b)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f1b)))
```

#### Untreated - Counts

```{r}
if (file.exists(f1c)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f1c)))
```

#### Untreated - Weights

```{r}
if (file.exists(f1d)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f1d)))
```

:::

---

# 2. CD4 → CD8 Signaling

::: {.callout-important icon=false}
## Directional Analysis

This section examines how **CD4 T cells communicate TO CD8 T cells**. CD4 cells express ligands, CD8 cells express receptors.
:::

## 2.1 Significant Ligand-Receptor Pairs

::: {.callout-note icon=false}
### Interpretation Guide

**Bubble Plot:**
- Each dot = one ligand-receptor pair
- Dot size = interaction strength  
- Dot color = statistical significance
- Left vs right = Treated vs Untreated
- Dots appearing in only one condition indicate treatment effects

**Tables:**
List all significant interactions with complete statistics. Use search/filter to find specific genes.
:::

```{r cd4-cd8-comm}
comm_t_cd4_cd8 <- get_comm(cellchat_list$Treated,   CD4_LABEL, CD8_LABEL, PVALUE_CUTOFF)
comm_u_cd4_cd8 <- get_comm(cellchat_list$Untreated, CD4_LABEL, CD8_LABEL, PVALUE_CUTOFF)

logmsg("CD4→CD8 sig pairs: Treated=%s, Untreated=%s",
       nrow(comm_t_cd4_cd8$sig), nrow(comm_u_cd4_cd8$sig))

write.csv(comm_t_cd4_cd8$sig, file.path(TABLE_DIR, "sig_pairs_CD4_to_CD8_Treated.csv"), row.names=FALSE)
write.csv(comm_u_cd4_cd8$sig, file.path(TABLE_DIR, "sig_pairs_CD4_to_CD8_Untreated.csv"), row.names=FALSE)
```

### Comparative Bubble Plot

```{r}
f_bub_cd4_cd8 <- file.path(PLOT_DIR, "bubble_CD4_to_CD8_Treated_vs_Untreated.png")
bubble_compare(cellchat_merged, CD4_LABEL, CD8_LABEL, f_bub_cd4_cd8, "CD4 → CD8 (Treated vs Untreated)")

if (file.exists(f_bub_cd4_cd8)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_bub_cd4_cd8)))
```

### Detailed Interaction Tables

::: {.panel-tabset}

#### Treated (p < 0.05)

```{r}
DT::datatable(
  comm_t_cd4_cd8$sig,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

#### Untreated (p < 0.05)

```{r}
DT::datatable(
  comm_u_cd4_cd8$sig,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

:::

---

## 2.2 Pathway-Level Analysis

::: {.callout-tip icon=false}
### Understanding Pathways

**Pathway Tables:**
- **n_pairs:** Number of different ligand-receptor combinations in this pathway
- **sum_prob:** Overall pathway strength (higher = more active)
- Pathways ranked by cumulative strength

**Chord Diagrams:**
- Ribbons connect sender to receiver
- Ribbon width = interaction strength
- Compare Treated vs Untreated side-by-side

**Heatmaps:**
Show interaction strength for specific ligand-receptor pairs within pathways (darker = stronger)
:::

```{r cd4-cd8-pathways}
pw_t_cd4_cd8 <- rank_pathways(comm_t_cd4_cd8$sig)
pw_u_cd4_cd8 <- rank_pathways(comm_u_cd4_cd8$sig)

write.csv(pw_t_cd4_cd8, file.path(TABLE_DIR, "pathways_CD4_to_CD8_Treated.csv"), row.names=FALSE)
write.csv(pw_u_cd4_cd8, file.path(TABLE_DIR, "pathways_CD4_to_CD8_Untreated.csv"), row.names=FALSE)

top_pw_cd4_cd8 <- unique(c(
  head(pw_t_cd4_cd8$pathway, TOP_PATHWAYS),
  head(pw_u_cd4_cd8$pathway, TOP_PATHWAYS)
))
top_pw_cd4_cd8 <- top_pw_cd4_cd8[!is.na(top_pw_cd4_cd8)]

logmsg("Top pathways for CD4->CD8: %s", paste(top_pw_cd4_cd8, collapse=", "))

chords_t_cd4_cd8 <- make_chords(cellchat_list$Treated,   top_pw_cd4_cd8, "Treated_CD4_to_CD8",   CD4_LABEL, CD8_LABEL)
chords_u_cd4_cd8 <- make_chords(cellchat_list$Untreated, top_pw_cd4_cd8, "Untreated_CD4_to_CD8", CD4_LABEL, CD8_LABEL)

hm_cd4_cd8 <- make_heatmaps(
  cellchat_merged,
  head(top_pw_cd4_cd8, min(length(top_pw_cd4_cd8), TOP_PATHWAYS)),
  "CD4_to_CD8",
  CD4_LABEL,
  CD8_LABEL
)
```

### Pathway Summary Tables

::: {.panel-tabset}

#### Treated Pathways

```{r}
DT::datatable(
  pw_t_cd4_cd8,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

#### Untreated Pathways

```{r}
DT::datatable(
  pw_u_cd4_cd8,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

:::

### Chord Plots - All Pathways Combined

```{r cd4-cd8-merged-chords}
f_chord_t_cd4_cd8 <- file.path(PLOT_DIR, "chord_merged_Treated_CD4_to_CD8.png")
f_chord_u_cd4_cd8 <- file.path(PLOT_DIR, "chord_merged_Untreated_CD4_to_CD8.png")

save_png_safe(f_chord_t_cd4_cd8, width=1400, height=1400, res=200, expr={
  netVisual_chord_gene(
    cellchat_list$Treated,
    sources.use = CD4_LABEL,
    targets.use = CD8_LABEL,
    lab.cex = 0.6,
    legend.pos.x = 10,
    title.name = "Treated: CD4 → CD8 (All Pathways)"
  )
})

save_png_safe(f_chord_u_cd4_cd8, width=1400, height=1400, res=200, expr={
  netVisual_chord_gene(
    cellchat_list$Untreated,
    sources.use = CD4_LABEL,
    targets.use = CD8_LABEL,
    lab.cex = 0.6,
    legend.pos.x = 10,
    title.name = "Untreated: CD4 → CD8 (All Pathways)"
  )
})
```

::: {.side-by-side}

::: {}
**Treated - All Pathways**

```{r}
#| out-width: "100%"
if (file.exists(f_chord_t_cd4_cd8)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_chord_t_cd4_cd8)))
}
```
:::

::: {}
**Untreated - All Pathways**

```{r}
#| out-width: "100%"
if (file.exists(f_chord_u_cd4_cd8)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_chord_u_cd4_cd8)))
}
```
:::

:::

### Pathway-Specific Heatmaps

```{r}
#| results: asis
#| fig-show: hold
#| out-width: "50%"
if (length(hm_cd4_cd8) > 0) {
  for (f in hm_cd4_cd8) {
    if (file.exists(f)) {
      rel_path <- file.path(PLOT_DIR_REL, basename(f))
      cat("![", gsub("heatmap_CD4_to_CD8_", "", tools::file_path_sans_ext(basename(f))), "](", rel_path, ")", sep="")
    }
  }
} else {
  cat("No heatmaps were generated for CD4→CD8 pathways.\n\n")
}
```

### Communication Matrix Heatmaps (All Cell Types)

::: {.callout-note icon=false}
### Contextual View

Shows the **complete signaling landscape** for each pathway - not just CD4-CD8, but all cell type interactions.

**How to Read:**
- **Y-axis (rows):** Sender cells expressing the ligand
- **X-axis (columns):** Receiver cells expressing the receptor
- **Color intensity:** Communication strength (darker = stronger)
- **Side-by-side panels:** Left = Treated, Right = Untreated
- Look for CD4 and CD8 rows/columns to see their specific interactions
- Helps determine if CD4-CD8 crosstalk is unique or part of broader patterns
:::

```{r cd4-cd8-comm-matrix}
comm_matrix_cd4_cd8 <- list()

available_pathways_t <- cellchat_list$Treated@netP$pathways
available_pathways_u <- cellchat_list$Untreated@netP$pathways
all_available <- unique(c(available_pathways_t, available_pathways_u))

all_pw_cd4_cd8 <- unique(c(pw_t_cd4_cd8$pathway, pw_u_cd4_cd8$pathway))
pathways_to_plot <- intersect(all_pw_cd4_cd8, all_available)

logmsg("CD4→CD8 Communication matrix: generating heatmaps for ALL %d pathways", length(pathways_to_plot))

for (pw in pathways_to_plot) {
  logmsg("=== CD4→CD8: Processing pathway: '%s' ===", pw)
  out <- file.path(PLOT_DIR, sprintf("comm_matrix_%s_%s.png", "CD4_to_CD8", gsub("[^A-Za-z0-9]+","_", pw)))
  
  made <- save_png_safe(out, width=3200, height=1600, res=200, expr={
    ht1 <- netVisual_heatmap(
      cellchat_list$Treated,
      signaling = pw,
      measure = "weight",
      font.size = 7,
      title.name = paste0("Treated - ", pw)
    )
    
    ht2 <- netVisual_heatmap(
      cellchat_list$Untreated,
      signaling = pw,
      measure = "weight",
      font.size = 7,
      title.name = paste0("Untreated - ", pw)
    )
    
    draw(ht1 + ht2, ht_gap = unit(0.5, "cm"))
  })
  
  if (!is.na(made)) {
    comm_matrix_cd4_cd8 <- c(comm_matrix_cd4_cd8, made)
    logmsg("✓ SUCCESS: Created %s", basename(made))
  } else {
    logmsg("✗ FAILED: Could not create heatmap for %s", pw)
  }
}

logmsg("Total communication matrix heatmaps created for CD4→CD8: %d", length(comm_matrix_cd4_cd8))
```

::: {.panel-tabset}

```{r}
#| results: asis
if (length(comm_matrix_cd4_cd8) == 0) {
  cat("**No communication matrix heatmaps were generated.** Check the console output for log messages.\n\n")
} else {
  for (f in comm_matrix_cd4_cd8) {
    if (file.exists(f)) {
      pw_name <- gsub("comm_matrix_CD4_to_CD8_", "", tools::file_path_sans_ext(basename(f)))
      cat("#### ", pw_name, "\n\n")
      rel_path <- file.path(PLOT_DIR_REL, basename(f))
      cat("![](", rel_path, "){width=90%}\n\n")
    }
  }
}
```

:::

---

# 3. CD8 → CD4 Signaling

::: {.callout-important icon=false}
## Reverse Direction Analysis

This section examines how **CD8 T cells communicate TO CD4 T cells** (reverse of Section 2). CD8 cells express ligands, CD4 cells express receptors.

**Why This Matters:**
Bidirectional communication reveals whether cells engage in two-way conversations or if signaling is primarily unidirectional.
:::

## 3.1 Significant Ligand-Receptor Pairs

```{r cd8-cd4-comm}
comm_t_cd8_cd4 <- get_comm(cellchat_list$Treated,   CD8_LABEL, CD4_LABEL, PVALUE_CUTOFF)
comm_u_cd8_cd4 <- get_comm(cellchat_list$Untreated, CD8_LABEL, CD4_LABEL, PVALUE_CUTOFF)

logmsg("CD8→CD4 sig pairs: Treated=%s, Untreated=%s",
       nrow(comm_t_cd8_cd4$sig), nrow(comm_u_cd8_cd4$sig))

write.csv(comm_t_cd8_cd4$sig, file.path(TABLE_DIR, "sig_pairs_CD8_to_CD4_Treated.csv"), row.names=FALSE)
write.csv(comm_u_cd8_cd4$sig, file.path(TABLE_DIR, "sig_pairs_CD8_to_CD4_Untreated.csv"), row.names=FALSE)
```

### Comparative Bubble Plot

```{r}
f_bub_cd8_cd4 <- file.path(PLOT_DIR, "bubble_CD8_to_CD4_Treated_vs_Untreated.png")
bubble_compare(cellchat_merged, CD8_LABEL, CD4_LABEL, f_bub_cd8_cd4, "CD8 → CD4 (Treated vs Untreated)")

if (file.exists(f_bub_cd8_cd4)) knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_bub_cd8_cd4)))
```

### Detailed Interaction Tables

::: {.panel-tabset}

#### Treated (p < 0.05)

```{r}
DT::datatable(
  comm_t_cd8_cd4$sig,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

#### Untreated (p < 0.05)

```{r}
DT::datatable(
  comm_u_cd8_cd4$sig,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

:::

---

## 3.2 Pathway-Level Analysis

::: {.callout-warning icon=false}
### Comparing Directionality

**Key Questions:**
1. Are the same pathways active in both directions (CD4→CD8 AND CD8→CD4)?
2. Are some pathways unique to one direction?
3. Does treatment affect directionality?

**Example:** If MHC-II appears strongly in Section 2 but weakly here, CD4 cells primarily signal TO CD8, not vice versa.
:::

```{r cd8-cd4-pathways}
pw_t_cd8_cd4 <- rank_pathways(comm_t_cd8_cd4$sig)
pw_u_cd8_cd4 <- rank_pathways(comm_u_cd8_cd4$sig)

write.csv(pw_t_cd8_cd4, file.path(TABLE_DIR, "pathways_CD8_to_CD4_Treated.csv"), row.names=FALSE)
write.csv(pw_u_cd8_cd4, file.path(TABLE_DIR, "pathways_CD8_to_CD4_Untreated.csv"), row.names=FALSE)

top_pw_cd8_cd4 <- unique(c(
  head(pw_t_cd8_cd4$pathway, TOP_PATHWAYS),
  head(pw_u_cd8_cd4$pathway, TOP_PATHWAYS)
))
top_pw_cd8_cd4 <- top_pw_cd8_cd4[!is.na(top_pw_cd8_cd4)]

chords_t_cd8_cd4 <- make_chords(cellchat_list$Treated,   top_pw_cd8_cd4, "Treated_CD8_to_CD4",   CD8_LABEL, CD4_LABEL)
chords_u_cd8_cd4 <- make_chords(cellchat_list$Untreated, top_pw_cd8_cd4, "Untreated_CD8_to_CD4", CD8_LABEL, CD4_LABEL)

hm_cd8_cd4 <- make_heatmaps(
  cellchat_merged,
  head(top_pw_cd8_cd4, min(length(top_pw_cd8_cd4), TOP_PATHWAYS)),
  "CD8_to_CD4",
  CD8_LABEL,
  CD4_LABEL
)
```

### Pathway Summary Tables

::: {.panel-tabset}

#### Treated Pathways

```{r}
DT::datatable(
  pw_t_cd8_cd4,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

#### Untreated Pathways

```{r}
DT::datatable(
  pw_u_cd8_cd4,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover'
)
```

:::

### Chord Plots - All Pathways Combined

```{r cd8-cd4-merged-chords}
f_chord_t_cd8_cd4 <- file.path(PLOT_DIR, "chord_merged_Treated_CD8_to_CD4.png")
f_chord_u_cd8_cd4 <- file.path(PLOT_DIR, "chord_merged_Untreated_CD8_to_CD4.png")

save_png_safe(f_chord_t_cd8_cd4, width=1400, height=1400, res=200, expr={
  netVisual_chord_gene(
    cellchat_list$Treated,
    sources.use = CD8_LABEL,
    targets.use = CD4_LABEL,
    lab.cex = 0.6,
    legend.pos.x = 10,
    title.name = "Treated: CD8 → CD4 (All Pathways)"
  )
})

save_png_safe(f_chord_u_cd8_cd4, width=1400, height=1400, res=200, expr={
  netVisual_chord_gene(
    cellchat_list$Untreated,
    sources.use = CD8_LABEL,
    targets.use = CD4_LABEL,
    lab.cex = 0.6,
    legend.pos.x = 10,
    title.name = "Untreated: CD8 → CD4 (All Pathways)"
  )
})
```

::: {.side-by-side}

::: {}
**Treated - All Pathways**

```{r}
#| out-width: "100%"
if (file.exists(f_chord_t_cd8_cd4)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_chord_t_cd8_cd4)))
}
```
:::

::: {}
**Untreated - All Pathways**

```{r}
#| out-width: "100%"
if (file.exists(f_chord_u_cd8_cd4)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_chord_u_cd8_cd4)))
}
```
:::

:::

### Pathway-Specific Heatmaps

```{r}
#| results: asis
#| fig-show: hold
#| out-width: "50%"
if (length(hm_cd8_cd4) > 0) {
  for (f in hm_cd8_cd4) {
    if (file.exists(f)) {
      rel_path <- file.path(PLOT_DIR_REL, basename(f))
      cat("![", gsub("heatmap_CD8_to_CD4_", "", tools::file_path_sans_ext(basename(f))), "](", rel_path, ")", sep="")
    }
  }
} else {
  cat("No heatmaps were generated for CD8→CD4 pathways.\n\n")
}
```

### Communication Matrix Heatmaps (All Cell Types)

```{r cd8-cd4-comm-matrix}
comm_matrix_cd8_cd4 <- list()

available_pathways_t <- cellchat_list$Treated@netP$pathways
available_pathways_u <- cellchat_list$Untreated@netP$pathways
all_available <- unique(c(available_pathways_t, available_pathways_u))

all_pw_cd8_cd4 <- unique(c(pw_t_cd8_cd4$pathway, pw_u_cd8_cd4$pathway))
pathways_to_plot <- intersect(all_pw_cd8_cd4, all_available)

logmsg("CD8→CD4 Communication matrix: generating heatmaps for ALL %d pathways", length(pathways_to_plot))

for (pw in pathways_to_plot) {
  logmsg("=== CD8→CD4: Processing pathway: '%s' ===", pw)
  out <- file.path(PLOT_DIR, sprintf("comm_matrix_%s_%s.png", "CD8_to_CD4", gsub("[^A-Za-z0-9]+","_", pw)))
  
  made <- save_png_safe(out, width=3200, height=1600, res=200, expr={
    ht1 <- netVisual_heatmap(
      cellchat_list$Treated,
      signaling = pw,
      measure = "weight",
      font.size = 7,
      title.name = paste0("Treated - ", pw)
    )
    
    ht2 <- netVisual_heatmap(
      cellchat_list$Untreated,
      signaling = pw,
      measure = "weight",
      font.size = 7,
      title.name = paste0("Untreated - ", pw)
    )
    
    draw(ht1 + ht2, ht_gap = unit(0.5, "cm"))
  })
  
  if (!is.na(made)) {
    comm_matrix_cd8_cd4 <- c(comm_matrix_cd8_cd4, made)
    logmsg("✓ SUCCESS: Created %s", basename(made))
  } else {
    logmsg("✗ FAILED: Could not create heatmap for %s", pw)
  }
}

logmsg("Total communication matrix heatmaps created for CD8→CD4: %d", length(comm_matrix_cd8_cd4))
```

::: {.panel-tabset}

```{r}
#| results: asis
if (length(comm_matrix_cd8_cd4) == 0) {
  cat("**No communication matrix heatmaps were generated.** Check the console output for log messages.\n\n")
} else {
  for (f in comm_matrix_cd8_cd4) {
    if (file.exists(f)) {
      pw_name <- gsub("comm_matrix_CD8_to_CD4_", "", tools::file_path_sans_ext(basename(f)))
      cat("#### ", pw_name, "\n\n")
      rel_path <- file.path(PLOT_DIR_REL, basename(f))
      cat("![](", rel_path, "){width=90%}\n\n")
    }
  }
}
```

:::

---

## Session Info {.appendix}

```{r}
sessionInfo()
```
