---
title: "CellChat Analysis: CD4 and CD8 Communication Overview"
subtitle: "Comprehensive T Cell Interaction Landscape"
author: "Generated with CellChat v2.1.1"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: 
      light: flatly
      dark: darkly
    css: styles.css
    embed-resources: true
    smooth-scroll: true
    link-external-newwindow: true
execute:
  warning: false
  message: false
  cache: false
  echo: false
---

```{css, echo=FALSE}
/* Custom styling */
.callout-note {
  border-left: 4px solid #4A90E2;
  background-color: #f0f7ff;
}

.callout-important {
  border-left: 4px solid #E74C3C;
  background-color: #fff5f5;
}

.callout-tip {
  border-left: 4px solid #2ECC71;
  background-color: #f0fff4;
}

.callout-warning {
  border-left: 4px solid #F39C12;
  background-color: #fffbf0;
}

h1 { 
  color: #2c3e50;
  border-bottom: 3px solid #3498db;
  padding-bottom: 10px;
}

h2 { 
  color: #34495e;
  margin-top: 30px;
  border-bottom: 2px solid #95a5a6;
  padding-bottom: 5px;
}

h3 { 
  color: #7f8c8d;
}

.panel-tabset > .panel-tabs {
  border-bottom: 2px solid #3498db;
}

.dataTables_wrapper {
  font-size: 0.9em;
}

table {
  border-collapse: collapse;
  width: 100%;
  margin: 20px 0;
}

th {
  background-color: #3498db;
  color: white;
  padding: 12px;
  text-align: left;
}

tr:nth-child(even) {
  background-color: #f8f9fa;
}

pre {
  background-color: #f6f8fa;
  border: 1px solid #e1e4e8;
  border-radius: 6px;
  padding: 16px;
}
```

::: {.callout-note icon=false}
## About This Report

**CellChat v2.1.1** infers cell-cell communication using single-cell RNA-seq data by comparing gene expression to a curated database of ~3,300 validated molecular interactions (ligands, receptors, and cofactors).

This report provides a **comprehensive overview** of CD4 and CD8 T cell communication patterns with **all other cell types** in the tissue microenvironment, comparing Treated vs Untreated conditions.
:::

---

::: {.callout-important icon=false}
## Analysis Scope

Unlike the detailed bidirectional CD4-CD8 crosstalk analysis, this report examines:

- **CD4 T cells** communicating with all cell types in the dataset
- **CD8 T cells** communicating with all cell types in the dataset
- Both incoming (receiving signals) and outgoing (sending signals) patterns
- Complete ligand-receptor interaction tables (all p-values)
- Global signaling pathway contribution patterns
:::

---

## Setup

```{r setup}
#| code-summary: "Load libraries and configure paths"

suppressPackageStartupMessages({
  library(CellChat)
  library(patchwork)
  library(ComplexHeatmap)
  library(circlize)
  library(dplyr)
  library(data.table)
  library(fs)
  library(ggplot2)
  library(DT)
})

OUTPUT_DIR <- file.path(
  "/Users/cheungf/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/R/23_301",
  "OUTPUT_CELLCHAT_23_JAN_26"
)

CD4_LABEL <- "Conventional_CD4"
CD8_LABEL <- "Conventional_CD8"

DOC_DIR <- "/Users/cheungf/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/R/23_301"
PLOT_DIR_REL  <- "plots_cd4_cd8_overview"
TABLE_DIR_REL <- "tables_cd4_cd8_overview"
PLOT_DIR  <- file.path(DOC_DIR, PLOT_DIR_REL)
TABLE_DIR <- file.path(DOC_DIR, TABLE_DIR_REL)
dir_create(PLOT_DIR)
dir_create(TABLE_DIR)

PVALUE_CUTOFF <- 0.05

logmsg <- function(...) cat(sprintf("[%s] %s\n", format(Sys.time(), "%H:%M:%S"), sprintf(...)))

cellchat_list <- list(
  Treated   = readRDS(file.path(OUTPUT_DIR, "cellchat_Treated_FINAL.rds")),
  Untreated = readRDS(file.path(OUTPUT_DIR, "cellchat_Untreated_FINAL.rds"))
)

cellchat_merged <- readRDS(file.path(OUTPUT_DIR, "cellchat_merged_WITH_Step10_ALL.rds"))

logmsg("✓ Loaded Treated/Untreated FINAL + merged RDS.")
```

```{r helpers}
#| code-summary: "Utility functions"

save_png_safe <- function(file, width=1800, height=1200, res=200, expr) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  ok <- TRUE
  png(filename=file, width=width, height=height, res=res)
  tryCatch(expr, error = function(e) { 
    ok <<- FALSE
    message("⚠️ PLOT FAILED: ", basename(file), " | ", e$message) 
  })
  dev.off()
  if (!ok || !file.exists(file) || file.info(file)$size == 0) {
    if (file.exists(file)) unlink(file)
    return(NA_character_)
  }
  file
}

get_comm_involving <- function(cc, cell_type, p.cut=0.05) {
  df_all <- suppressWarnings(subsetCommunication(cc))
  if (is.null(df_all) || nrow(df_all) == 0) return(list(all=data.frame(), sig=data.frame()))
  
  df_all <- as.data.frame(df_all)
  if (!("pval" %in% colnames(df_all))) {
    pcol <- grep("^p", colnames(df_all), value=TRUE)[1]
    if (!is.na(pcol)) df_all$pval <- df_all[[pcol]]
  }
  
  df <- df_all %>%
    mutate(pval = as.numeric(pval)) %>%
    filter(!is.na(pval)) %>%
    filter(source == cell_type | target == cell_type) %>%
    arrange(pval)
  
  list(all=df, sig=df %>% filter(pval < p.cut))
}
```

---

# 1. Significant Ligand-Receptor Interactions

::: {.callout-tip icon=false}
## Statistical Significance

Interactions with **p-value < 0.05** are considered statistically significant, meaning there's high confidence these communication events are real and not due to chance.

CellChat identifies significant communications by:
1. Identifying over-expressed ligands and receptors in each cell type
2. Performing permutation testing to assess statistical confidence
3. Filtering for biologically meaningful interactions
:::

## 1.1 CD4 T Cells with All Partners

::: {.callout-note icon=false}
### Analysis Scope

This section shows **all significant interactions involving CD4 T cells** communicating with any other cell type in the dataset (epithelial cells, immune cells, stromal cells, etc.).
:::

```{r cd4-all-data}
comm_t_cd4_all <- get_comm_involving(cellchat_list$Treated, CD4_LABEL, PVALUE_CUTOFF)
comm_u_cd4_all <- get_comm_involving(cellchat_list$Untreated, CD4_LABEL, PVALUE_CUTOFF)

logmsg("CD4 T (all partners): Treated=%d sig, Untreated=%d sig", 
       nrow(comm_t_cd4_all$sig), nrow(comm_u_cd4_all$sig))

write.csv(comm_t_cd4_all$sig, file.path(TABLE_DIR, "sig_CD4_all_Treated.csv"), row.names=FALSE)
write.csv(comm_u_cd4_all$sig, file.path(TABLE_DIR, "sig_CD4_all_Untreated.csv"), row.names=FALSE)

cd4_sources <- unique(c(
  comm_t_cd4_all$sig$source[comm_t_cd4_all$sig$target == CD4_LABEL],
  comm_u_cd4_all$sig$source[comm_u_cd4_all$sig$target == CD4_LABEL]
))

cd4_targets <- unique(c(
  comm_t_cd4_all$sig$target[comm_t_cd4_all$sig$source == CD4_LABEL],
  comm_u_cd4_all$sig$target[comm_u_cd4_all$sig$source == CD4_LABEL]
))
```

### CD4 Sending Signals (CD4 → Others)

::: {.callout-note icon=false}
**Interpretation:** This plot shows all cell types that **receive signals from CD4 cells**. Each bubble represents a ligand-receptor pair where CD4 expresses the ligand.
:::

```{r}
#| fig-width: 16
#| fig-height: 10
#| fig-cap: "CD4 T cells sending signals to other cell types (Treated vs Untreated comparison)"

netVisual_bubble(
  cellchat_merged,
  sources.use = CD4_LABEL,
  targets.use = cd4_targets,
  comparison = c(1, 2),
  angle.x = 45,
  remove.isolate = FALSE
) + 
  ggtitle("CD4 → Others (Treated vs Untreated)") 
```

### CD4 Receiving Signals (Others → CD4)

::: {.callout-note icon=false}
**Interpretation:** This plot shows all cell types that **send signals to CD4 cells**. Each bubble represents a ligand-receptor pair where CD4 expresses the receptor.
:::

```{r}
#| fig-width: 16
#| fig-height: 10
#| fig-cap: "Other cell types sending signals to CD4 T cells (Treated vs Untreated comparison)"

netVisual_bubble(
  cellchat_merged,
  sources.use = cd4_sources,
  targets.use = CD4_LABEL,
  comparison = c(1, 2),
  angle.x = 45,
  remove.isolate = FALSE
) + 
  ggtitle("Others → CD4 (Treated vs Untreated)") 
```

---

## 1.2 CD8 T Cells with All Partners

::: {.callout-note icon=false}
### Analysis Scope

This section shows **all significant interactions involving CD8 T cells** communicating with any other cell type in the dataset.
:::

```{r cd8-all-data}
comm_t_cd8_all <- get_comm_involving(cellchat_list$Treated, CD8_LABEL, PVALUE_CUTOFF)
comm_u_cd8_all <- get_comm_involving(cellchat_list$Untreated, CD8_LABEL, PVALUE_CUTOFF)

logmsg("CD8 T (all partners): Treated=%d sig, Untreated=%d sig", 
       nrow(comm_t_cd8_all$sig), nrow(comm_u_cd8_all$sig))

write.csv(comm_t_cd8_all$sig, file.path(TABLE_DIR, "sig_CD8_all_Treated.csv"), row.names=FALSE)
write.csv(comm_u_cd8_all$sig, file.path(TABLE_DIR, "sig_CD8_all_Untreated.csv"), row.names=FALSE)

cd8_sources <- unique(c(
  comm_t_cd8_all$sig$source[comm_t_cd8_all$sig$target == CD8_LABEL],
  comm_u_cd8_all$sig$source[comm_u_cd8_all$sig$target == CD8_LABEL]
))

cd8_targets <- unique(c(
  comm_t_cd8_all$sig$target[comm_t_cd8_all$sig$source == CD8_LABEL],
  comm_u_cd8_all$sig$target[comm_u_cd8_all$sig$source == CD8_LABEL]
))
```

### CD8 Sending Signals (CD8 → Others)

::: {.callout-note icon=false}
**Interpretation:** This plot shows all cell types that **receive signals from CD8 cells**. Each bubble represents a ligand-receptor pair where CD8 expresses the ligand.
:::

```{r}
#| fig-width: 16
#| fig-height: 10
#| fig-cap: "CD8 T cells sending signals to other cell types (Treated vs Untreated comparison)"

netVisual_bubble(
  cellchat_merged,
  sources.use = CD8_LABEL,
  targets.use = cd8_targets,
  comparison = c(1, 2),
  angle.x = 45,
  remove.isolate = FALSE
) + 
  ggtitle("CD8 → Others (Treated vs Untreated)") #+
#  theme_minimal(base_size = 12) +
#  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### CD8 Receiving Signals (Others → CD8)

::: {.callout-note icon=false}
**Interpretation:** This plot shows all cell types that **send signals to CD8 cells**. Each bubble represents a ligand-receptor pair where CD8 expresses the receptor.
:::

```{r}
#| fig-width: 16
#| fig-height: 10
#| fig-cap: "Other cell types sending signals to CD8 T cells (Treated vs Untreated comparison)"

netVisual_bubble(
  cellchat_merged,
  sources.use = cd8_sources,
  targets.use = CD8_LABEL,
  comparison = c(1, 2),
  angle.x = 45,
  remove.isolate = FALSE
) + 
  ggtitle("Others → CD8 (Treated vs Untreated)")
```

---

# 2. Complete Interaction Tables

::: {.callout-important icon=false}
## Unfiltered Data

These tables show **all inferred interactions** (not filtered by p-value threshold). This allows you to:
- Explore borderline significant interactions (0.05 < p < 0.10)
- Identify trends that might not reach statistical significance
- Export complete datasets for custom filtering
- Perform your own statistical analyses
:::

## 2.1 CD4 T Cells - All Interactions

```{r cd4-all-table}
comm_t_cd4_full <- get_comm_involving(cellchat_list$Treated, CD4_LABEL, p.cut = 1)
comm_u_cd4_full <- get_comm_involving(cellchat_list$Untreated, CD4_LABEL, p.cut = 1)
```

::: {.panel-tabset}

### Treated - All Interactions

```{r}
DT::datatable(
  comm_t_cd4_full$all,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover',
  caption = "Treated: All CD4 T cell Interactions (all p-values)"
)
```

### Untreated - All Interactions

```{r}
DT::datatable(
  comm_u_cd4_full$all,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover',
  caption = "Untreated: All CD4 T cell Interactions (all p-values)"
)
```

:::

---

## 2.2 CD8 T Cells - All Interactions

```{r cd8-all-table}
comm_t_cd8_full <- get_comm_involving(cellchat_list$Treated, CD8_LABEL, p.cut = 1)
comm_u_cd8_full <- get_comm_involving(cellchat_list$Untreated, CD8_LABEL, p.cut = 1)
```

::: {.panel-tabset}

### Treated - All Interactions

```{r}
DT::datatable(
  comm_t_cd8_full$all,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover',
  caption = "Treated: All CD8 T cell Interactions (all p-values)"
)
```

### Untreated - All Interactions

```{r}
DT::datatable(
  comm_u_cd8_full$all,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  class = 'cell-border stripe hover',
  caption = "Untreated: All CD8 T cell Interactions (all p-values)"
)
```

:::

---

# 3. Global Signaling Pathway Contributions

::: {.callout-note icon=false}
## Pathway Heatmaps

These heatmaps visualize which signaling pathways contribute most to **outgoing** (sending) and **incoming** (receiving) signals for each cell type across the entire dataset.

**How to Read:**
- **Rows:** Signaling pathways (e.g., TNF, MHC-I, CLEC)
- **Columns:** Cell types
- **Color intensity:** Contribution strength (darker = stronger)
- Look for CD4 and CD8 columns to see their pathway profiles
- Compare Treated vs Untreated to identify treatment-induced changes
:::

```{r contributing-signals}
f_out_t <- file.path(PLOT_DIR, "signaling_outgoing_treated.png")
f_out_u <- file.path(PLOT_DIR, "signaling_outgoing_untreated.png")
f_inc_t <- file.path(PLOT_DIR, "signaling_incoming_treated.png")
f_inc_u <- file.path(PLOT_DIR, "signaling_incoming_untreated.png")

save_png_safe(f_out_t, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Treated, pattern = "outgoing", width = 10, height = 20)
  draw(ht)
})

save_png_safe(f_out_u, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Untreated, pattern = "outgoing", width = 10, height = 20)
  draw(ht)
})

save_png_safe(f_inc_t, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Treated, pattern = "incoming", width = 10, height = 20)
  draw(ht)
})

save_png_safe(f_inc_u, width=1800, height=2000, res=200, expr={
  ht <- netAnalysis_signalingRole_heatmap(cellchat_list$Untreated, pattern = "incoming", width = 10, height = 20)
  draw(ht)
})
```

::: {.panel-tabset}

## Treated

### Outgoing Pathways

```{r}
#| out-width: "90%"
if (file.exists(f_out_t)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_out_t)))
}
```

### Incoming Pathways

```{r}
#| out-width: "90%"
if (file.exists(f_inc_t)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_inc_t)))
}
```

## Untreated

### Outgoing Pathways

```{r}
#| out-width: "90%"
if (file.exists(f_out_u)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_out_u)))
}
```

### Incoming Pathways

```{r}
#| out-width: "90%"
if (file.exists(f_inc_u)) {
  knitr::include_graphics(file.path(PLOT_DIR_REL, basename(f_inc_u)))
}
```

:::

---

# 4. References

1. Suoqin Jin et al., *Inference and analysis of cell-cell communication using CellChat*. Nature Communications, 12:1088 (2021). https://www.nature.com/articles/s41467-021-21246-9

---

## Session Info {.appendix}

```{r}
sessionInfo()
```
