---
title: "PBMC DEG and Pathway Analysis"
subtitle: "Healthy Controls vs Mild Untreated vs Severe Patients"
author: ""
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 22
    fig-height: 18
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This report compares PBMC cell-type-specific pathway enrichment and differential gene expression across:

- 4 healthy controls
- 4 mild untreated patients
- 4 severe patients

The analysis includes Hallmark, Reactome, and GO Biological Process pathway comparisons, direct single-contrast pathway plots for each contrast, and gene-level DEG comparisons.

# Setup and Functions

```{r}
#| label: setup-functions

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(ggrepel)
  library(fs)
  library(purrr)
  library(scales)
  library(knitr)     # for kable
  library(kableExtra) # for styled tables
})

BASE_DIR <- "deg_gsea_pooled_vs_within_pool_Monaco_fine_pruned_v3_consistent_disease_direction"
OUT_DIR  <- "figures_publication"
dir_create(OUT_DIR)

PATHWAY_DBS <- c("Hallmark", "Reactome", "GO_BP")

FDR_STRICT <- 0.05
FDR_LOOSE  <- 0.10
TOP_N      <- 5

DPI <- 300
WIDTH <- 22
HEIGHT_PER_ROW <- 4.5
NCOL_FACET <- 4

CONTRAST_SUFFIX <- "_pooled"
STRATEGY_DIR <- "pooled"

FORCE_LABEL_PATTERNS <- c(
  "interferon",
  "innate immune",
  "antiviral",
  "type i ifn",
  "type ii ifn",
  "ifn.alpha",
  "ifn.beta",
  "ifn.gamma",
  "jak.stat",
  "isg15",
  "irf"
)

FORCE_LABEL_REGEX <- paste(FORCE_LABEL_PATTERNS, collapse = "|")

CELLTYPE_ORDER <- c(
  "Classical_monocytes",
  "Non_classical_monocytes",
  "Intermediate_monocytes",
  "Myeloid_dendritic_cells",
  "Plasmacytoid_dendritic_cells",
  "Natural_killer_cells",
  "Terminal_effector_CD8_T_cells",
  "Effector_memory_CD8_T_cells",
  "Central_memory_CD8_T_cells",
  "Naive_CD8_T_cells",
  "Terminal_effector_CD4_T_cells",
  "Naive_CD4_T_cells",
  "Follicular_helper_T_cells",
  "T_regulatory_cells",
  "Th1_cells",
  "Th2_cells",
  "Th17_cells",
  "Th1_Th17_cells",
  "MAIT_cells",
  "Non_Vd2_gd_T_cells",
  "Vd2_gd_T_cells",
  "Naive_B_cells",
  "Non-switched_memory_B_cells",
  "Switched_memory_B_cells",
  "Exhausted_B_cells",
  "Plasmablasts",
  "Low-density_neutrophils",
  "Progenitor_cells"
)

CELLTYPE_LABELS <- c(
  Classical_monocytes = "Classical Monocytes",
  Non_classical_monocytes = "Non-classical Monocytes",
  Intermediate_monocytes = "Intermediate Monocytes",
  Myeloid_dendritic_cells = "Myeloid DCs",
  Plasmacytoid_dendritic_cells = "Plasmacytoid DCs",
  Natural_killer_cells = "NK Cells",
  Terminal_effector_CD8_T_cells = "Effector CD8 T",
  Effector_memory_CD8_T_cells = "Effector Memory CD8 T",
  Central_memory_CD8_T_cells = "Central Memory CD8 T",
  Naive_CD8_T_cells = "Naive CD8 T",
  Terminal_effector_CD4_T_cells = "Effector CD4 T",
  Naive_CD4_T_cells = "Naive CD4 T",
  Follicular_helper_T_cells = "Follicular Helper T",
  T_regulatory_cells = "T Regulatory",
  Th1_cells = "Th1",
  Th2_cells = "Th2",
  Th17_cells = "Th17",
  Th1_Th17_cells = "Th1/Th17",
  MAIT_cells = "MAIT Cells",
  Non_Vd2_gd_T_cells = "Non-Vδ2 γδ T",
  Vd2_gd_T_cells = "Vδ2 γδ T",
  Naive_B_cells = "Naive B Cells",
  `Non-switched_memory_B_cells` = "Non-switched Memory B",
  Switched_memory_B_cells = "Switched Memory B",
  Exhausted_B_cells = "Exhausted B Cells",
  Plasmablasts = "Plasmablasts",
  `Low-density_neutrophils` = "Low-density Neutrophils",
  Progenitor_cells = "Progenitor Cells"
)

quadrant_colors <- c(
  "Q1: Both Up" = "#C0392B",
  "Q2: Y-Up / X-Down" = "#2980B9",
  "Q3: Both Down" = "#1A5276",
  "Q4: X-Up / Y-Down" = "#922B21",
  "On Axis" = "grey75"
)

sig_shapes <- c(
  "Both significant" = 16,
  "X-contrast only" = 17,
  "Y-contrast only" = 15,
  "Neither significant" = 1
)

sig_alpha <- c(
  "Both <= 0.05" = 1.00,
  "Both / mixed 0.05-0.1" = 0.55,
  "X only <= 0.05" = 1.00,
  "X only 0.05-0.1" = 0.55,
  "Y only <= 0.05" = 1.00,
  "Y only 0.05-0.1" = 0.55,
  "Not significant" = 0.15
)

theme_pub <- function(base_size = 13) {
  theme_bw(base_size = base_size) +
    theme(
      strip.text = element_text(face = "bold", size = base_size - 1),
      strip.background = element_rect(fill = "grey94", color = "grey70"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      panel.border = element_rect(color = "grey60"),
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 2),
      plot.title = element_text(size = base_size + 2, face = "bold"),
      plot.subtitle = element_text(size = base_size - 1, color = "grey35"),
      legend.title = element_text(face = "bold", size = base_size - 1),
      legend.text = element_text(size = base_size - 2),
      legend.key.size = unit(0.45, "cm"),
      plot.margin = margin(8, 8, 8, 8)
    )
}

clean_cluster_names <- function(x) {
  x %>%
    str_replace_all("\\+", "") %>%
    str_replace_all("__", "_") %>%
    str_replace_all(" ", "_") %>%
    str_replace_all("\\.", "_") %>%
    str_replace_all("_+", "_") %>%
    str_replace("^cluster_", "") %>%
    str_replace("^DEG_", "") %>%
    str_replace("_DEG$", "") %>%
    str_replace("^ALL_clusters_", "") %>%
    str_replace("^_|_$", "") %>%
    str_trim()
}

clean_pathway_label <- function(x) {
  x %>%
    str_remove("^HALLMARK_") %>%
    str_remove("^GOBP_") %>%
    str_remove("^REACTOME_") %>%
    str_replace_all("_", " ") %>%
    str_to_title() %>%
    str_wrap(width = 35)
}

read_safe <- function(path) {
  if (!file.exists(path)) return(NULL)
  dt <- tryCatch(data.table::fread(path, fill = TRUE), error = function(e) NULL)
  if (is.null(dt) || nrow(dt) == 0) return(NULL)
  as.data.frame(dt)
}

discover_celltypes <- function() {
  base <- file.path(BASE_DIR, STRATEGY_DIR, "deg_by_celltype")
  if (!dir_exists(base)) stop("Cannot find: ", base)
  dir_ls(base, type = "directory") %>% basename() %>% sort()
}

gsea_path <- function(ct, contrast, db) {
  file.path(
    BASE_DIR,
    STRATEGY_DIR,
    "gsea_by_celltype",
    ct,
    paste0(ct, "_", contrast, CONTRAST_SUFFIX, "_", db, "_results.csv")
  )
}

deg_path <- function(ct, contrast) {
  file.path(
    BASE_DIR,
    STRATEGY_DIR,
    "deg_by_celltype",
    ct,
    paste0("DEG_", ct, "_", contrast, CONTRAST_SUFFIX, ".csv")
  )
}

fig_height <- function(df, ncol = NCOL_FACET) {
  n_ct <- length(unique(df$cluster_raw))
  n_row <- ceiling(n_ct / ncol)
  max(HEIGHT_PER_ROW * 2, n_row * HEIGHT_PER_ROW)
}

save_png <- function(p, fname, width = WIDTH, height) {
  path <- file.path(OUT_DIR, paste0(fname, ".png"))
  ggsave(path, p, width = width, height = height, dpi = DPI, limitsize = FALSE)
  cat("  Saved:", path, "\n")
}

annotate_sig <- function(df) {
  df %>%
    mutate(
      sig_status = case_when(
        !is.na(padj_x) & !is.na(padj_y) &
          padj_x <= FDR_LOOSE & padj_y <= FDR_LOOSE ~ "Both significant",
        !is.na(padj_x) & padj_x <= FDR_LOOSE &
          (is.na(padj_y) | padj_y > FDR_LOOSE) ~ "X-contrast only",
        !is.na(padj_y) & padj_y <= FDR_LOOSE &
          (is.na(padj_x) | padj_x > FDR_LOOSE) ~ "Y-contrast only",
        TRUE ~ "Neither significant"
      ),
      sig_tier = case_when(
        !is.na(padj_x) & !is.na(padj_y) &
          padj_x <= FDR_STRICT & padj_y <= FDR_STRICT ~ "Both <= 0.05",
        !is.na(padj_x) & !is.na(padj_y) &
          padj_x <= FDR_LOOSE & padj_y <= FDR_LOOSE ~ "Both / mixed 0.05-0.1",
        !is.na(padj_x) & padj_x <= FDR_STRICT &
          (is.na(padj_y) | padj_y > FDR_LOOSE) ~ "X only <= 0.05",
        !is.na(padj_x) & padj_x > FDR_STRICT &
          padj_x <= FDR_LOOSE &
          (is.na(padj_y) | padj_y > FDR_LOOSE) ~ "X only 0.05-0.1",
        !is.na(padj_y) & padj_y <= FDR_STRICT &
          (is.na(padj_x) | padj_x > FDR_LOOSE) ~ "Y only <= 0.05",
        !is.na(padj_y) & padj_y > FDR_STRICT &
          padj_y <= FDR_LOOSE &
          (is.na(padj_x) | padj_x > FDR_LOOSE) ~ "Y only 0.05-0.1",
        TRUE ~ "Not significant"
      ),
      quadrant = case_when(
        lfc_x > 0 & lfc_y > 0 ~ "Q1: Both Up",
        lfc_x < 0 & lfc_y > 0 ~ "Q2: Y-Up / X-Down",
        lfc_x < 0 & lfc_y < 0 ~ "Q3: Both Down",
        lfc_x > 0 & lfc_y < 0 ~ "Q4: X-Up / Y-Down",
        TRUE ~ "On Axis"
      ),
      label_strength = abs(lfc_x) + abs(lfc_y),
      neg_log10_padj = -log10(
        pmax(
          (pmin(replace_na(padj_x, 1), 1) +
             pmin(replace_na(padj_y, 1), 1)) / 2,
          1e-10
        )
      )
    )
}

load_gsea_pair <- function(celltypes, contrast_x, contrast_y, db) {
  all_rows <- list()

  for (ct in celltypes) {
    dx <- read_safe(gsea_path(ct, contrast_x, db))
    dy <- read_safe(gsea_path(ct, contrast_y, db))

    if (is.null(dx) || is.null(dy)) next
    if (!all(c("Description", "NES", "p.adjust") %in% colnames(dx))) next
    if (!all(c("Description", "NES", "p.adjust") %in% colnames(dy))) next

    merged <- inner_join(
      dx %>% dplyr::select(Description, lfc_x = NES, padj_x = p.adjust),
      dy %>% dplyr::select(Description, lfc_y = NES, padj_y = p.adjust),
      by = "Description"
    ) %>%
      filter(is.finite(lfc_x), is.finite(lfc_y)) %>%
      group_by(Description) %>%
      slice_max(abs(lfc_x) + abs(lfc_y), n = 1, with_ties = FALSE) %>%
      ungroup()

    if (nrow(merged) < 5) next

    merged <- merged %>%
      annotate_sig() %>%
      mutate(
        cluster_raw = clean_cluster_names(ct),
        pathway_clean = clean_pathway_label(Description)
      )

    all_rows[[ct]] <- merged
  }

  if (length(all_rows) == 0) return(NULL)
  bind_rows(all_rows)
}

# FIX: Added contrast_label argument so direction labels and axis labels
# are dynamic per contrast rather than hardcoded to "Severe vs Mild"
load_gsea_single <- function(celltypes, contrast, db) {
  all_rows <- list()

  for (ct in celltypes) {
    dx <- read_safe(gsea_path(ct, contrast, db))

    if (is.null(dx)) next
    if (!all(c("Description", "NES", "p.adjust") %in% colnames(dx))) next

    # Parse a human-readable label from the contrast string for axis/direction labels
    contrast_label <- contrast %>%
      str_replace_all("_vs_", " vs ") %>%
      str_replace_all("MildUntreated", "Mild Untreated") %>%
      str_replace_all("HC", "HC")

    df <- dx %>%
      dplyr::select(Description, NES, p.adjust) %>%
      filter(is.finite(NES)) %>%
      mutate(
        cluster_raw = clean_cluster_names(ct),
        pathway_clean = clean_pathway_label(Description),
        # FIX: direction labels now use the dynamic contrast_label
        direction = case_when(
          NES > 0 ~ paste0("Up in ", contrast_label),
          NES < 0 ~ paste0("Down in ", contrast_label),
          TRUE ~ "Neutral"
        ),
        sig_tier = case_when(
          p.adjust <= FDR_STRICT ~ "FDR \u2264 0.05",
          p.adjust <= FDR_LOOSE  ~ "FDR 0.05\u20130.10",
          TRUE                   ~ "Not significant"
        ),
        neg_log10_padj = -log10(pmax(p.adjust, 1e-10)),
        label_strength = abs(NES),
        # Store contrast_label so the plot function can use it
        contrast_label = contrast_label
      )

    if (nrow(df) < 5) next
    all_rows[[ct]] <- df
  }

  if (length(all_rows) == 0) return(NULL)
  bind_rows(all_rows)
}

load_deg_pair <- function(celltypes, contrast_x, contrast_y) {
  all_rows <- list()

  for (ct in celltypes) {
    dx <- read_safe(deg_path(ct, contrast_x))
    dy <- read_safe(deg_path(ct, contrast_y))

    if (is.null(dx) || is.null(dy)) next
    if (!all(c("gene", "logFC", "adj.P.Val") %in% colnames(dx))) next
    if (!all(c("gene", "logFC", "adj.P.Val") %in% colnames(dy))) next

    merged <- inner_join(
      dx %>% dplyr::select(gene, lfc_x = logFC, padj_x = adj.P.Val),
      dy %>% dplyr::select(gene, lfc_y = logFC, padj_y = adj.P.Val),
      by = "gene"
    ) %>%
      filter(is.finite(lfc_x), is.finite(lfc_y)) %>%
      group_by(gene) %>%
      slice_max(abs(lfc_x) + abs(lfc_y), n = 1, with_ties = FALSE) %>%
      ungroup()

    if (nrow(merged) < 10) next

    merged <- merged %>%
      annotate_sig() %>%
      mutate(
        cluster_raw = clean_cluster_names(ct),
        gene_label = gene
      )

    all_rows[[ct]] <- merged
  }

  if (length(all_rows) == 0) return(NULL)
  bind_rows(all_rows)
}

# FIX: Added contrast_label argument so direction labels and axis labels
# are dynamic per contrast rather than hardcoded to "Severe vs Mild"
load_deg_single <- function(celltypes, contrast) {
  all_rows <- list()

  for (ct in celltypes) {
    dx <- read_safe(deg_path(ct, contrast))

    if (is.null(dx)) next
    if (!all(c("gene", "logFC", "adj.P.Val") %in% colnames(dx))) next

    # Parse a human-readable label from the contrast string
    contrast_label <- contrast %>%
      str_replace_all("_vs_", " vs ") %>%
      str_replace_all("MildUntreated", "Mild Untreated") %>%
      str_replace_all("HC", "HC")

    df <- dx %>%
      dplyr::select(gene, logFC, adj.P.Val) %>%
      filter(is.finite(logFC)) %>%
      mutate(
        cluster_raw = clean_cluster_names(ct),
        # FIX: direction labels now use the dynamic contrast_label
        direction = case_when(
          logFC > 0 ~ paste0("Up in ", contrast_label),
          logFC < 0 ~ paste0("Down in ", contrast_label),
          TRUE ~ "Neutral"
        ),
        sig_tier = case_when(
          adj.P.Val <= FDR_STRICT ~ "FDR \u2264 0.05",
          adj.P.Val <= FDR_LOOSE  ~ "FDR 0.05\u20130.10",
          TRUE                    ~ "Not significant"
        ),
        neg_log10_padj = -log10(pmax(adj.P.Val, 1e-10)),
        label_strength = abs(logFC),
        # Store contrast_label so the plot function can use it
        contrast_label = contrast_label
      )

    if (nrow(df) < 10) next
    all_rows[[ct]] <- df
  }

  if (length(all_rows) == 0) return(NULL)
  bind_rows(all_rows)
}

add_celltype_labels <- function(df, celltype_order = NULL, celltype_labels = NULL) {
  if (is.null(df) || nrow(df) == 0) return(df)

  if (!is.null(celltype_order)) {
    present <- intersect(celltype_order, unique(df$cluster_raw))
    df <- df %>%
      filter(cluster_raw %in% present) %>%
      mutate(cluster_raw = factor(cluster_raw, levels = present))
  }

  if (!is.null(celltype_labels)) {
    df <- df %>%
      mutate(cluster_label = dplyr::recode(as.character(cluster_raw), !!!celltype_labels))

    if (!is.null(celltype_order)) {
      label_levels <- celltype_labels[intersect(celltype_order, names(celltype_labels))]
      df$cluster_label <- factor(df$cluster_label, levels = unname(label_levels))
    }
  } else {
    df$cluster_label <- df$cluster_raw
  }

  df
}

make_pub_scatter <- function(df,
                             x_label,
                             y_label,
                             figure_title,
                             figure_subtitle,
                             celltype_order = NULL,
                             celltype_labels = NULL,
                             show_nonsig = TRUE,
                             ncol = NCOL_FACET) {

  if (is.null(df) || nrow(df) == 0) return(NULL)

  df <- add_celltype_labels(df, celltype_order, celltype_labels)

  sig_df <- df %>% filter(sig_status != "Neither significant")
  nonsig_df <- df %>% filter(sig_status == "Neither significant")

  auto_labels <- sig_df %>%
    group_by(cluster_label, quadrant) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup()

  grep_labels <- sig_df %>%
    filter(str_detect(Description, regex(FORCE_LABEL_REGEX, ignore_case = TRUE))) %>%
    group_by(cluster_label, quadrant) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup()

  top_labels <- bind_rows(auto_labels, grep_labels) %>%
    distinct(cluster_label, Description, .keep_all = TRUE)

  cor_annots <- df %>%
    group_by(cluster_label) %>%
    summarise(
      pearson_r = suppressWarnings(cor(lfc_x, lfc_y, method = "pearson")),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(!is.na(pearson_r))

  ggplot() +
    {
      if (show_nonsig && nrow(nonsig_df) > 0)
        geom_point(
          data = nonsig_df,
          aes(x = lfc_x, y = lfc_y),
          color = "grey82",
          size = 0.6,
          alpha = 0.4,
          shape = 16
        )
    } +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "grey30", linewidth = 0.5) +
    geom_point(
      data = sig_df,
      aes(
        x = lfc_x,
        y = lfc_y,
        color = quadrant,
        shape = sig_status,
        alpha = sig_tier,
        size = neg_log10_padj
      )
    ) +
    geom_text_repel(
      data = top_labels,
      aes(x = lfc_x, y = lfc_y, label = pathway_clean, color = quadrant),
      size = 2.5,
      lineheight = 0.85,
      max.overlaps = 80,
      box.padding = 0.45,
      point.padding = 0.3,
      segment.size = 0.25,
      segment.alpha = 0.6,
      force = 2.5,
      min.segment.length = 0,
      show.legend = FALSE
    ) +
    geom_text(
      data = cor_annots,
      aes(x = -Inf, y = Inf, label = sprintf("r = %.2f  n = %d", pearson_r, n)),
      hjust = -0.08,
      vjust = 1.5,
      size = 3.0,
      color = "grey30",
      fontface = "italic",
      inherit.aes = FALSE
    ) +
    facet_wrap(~ cluster_label, scales = "free", ncol = ncol) +
    scale_color_manual(values = quadrant_colors, name = "Direction") +
    scale_shape_manual(values = sig_shapes, name = "Significance", drop = FALSE) +
    scale_alpha_manual(
      values = sig_alpha,
      breaks = c(
        "Both <= 0.05",
        "Both / mixed 0.05-0.1",
        "X only <= 0.05",
        "X only 0.05-0.1",
        "Y only <= 0.05",
        "Y only 0.05-0.1"
      ),
      name = "FDR tier"
    ) +
    scale_size_continuous(
      name = expression(-log[10](mean~padj)),
      range = c(0.8, 4.5),
      breaks = c(1, 2, 3, 5)
    ) +
    labs(
      title = figure_title,
      subtitle = figure_subtitle,
      x = x_label,
      y = y_label
    ) +
    theme_pub(base_size = 13)
}

# FIX: direction color scale and x-axis label now derived from data (contrast_label column)
# instead of being hardcoded to "Severe vs Mild Untreated"
make_single_gsea_plot <- function(df,
                                  figure_title,
                                  figure_subtitle,
                                  celltype_order = NULL,
                                  celltype_labels = NULL,
                                  ncol = NCOL_FACET) {

  if (is.null(df) || nrow(df) == 0) return(NULL)

  df <- add_celltype_labels(df, celltype_order, celltype_labels)

  # Derive axis label and color scale from the data's contrast_label column
  contrast_label <- unique(df$contrast_label)[1]
  up_label   <- paste0("Up in ", contrast_label)
  down_label <- paste0("Down in ", contrast_label)
  x_axis_label <- paste0(contrast_label, " \u2014 NES")

  direction_colors <- c(
    setNames("#C0392B", up_label),
    setNames("#2980B9", down_label),
    "Neutral" = "grey75"
  )

  sig_df    <- df %>% filter(p.adjust <= FDR_LOOSE)
  nonsig_df <- df %>% filter(p.adjust > FDR_LOOSE)

  auto_labels <- sig_df %>%
    group_by(cluster_label, direction) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup()

  grep_labels <- sig_df %>%
    filter(str_detect(Description, regex(FORCE_LABEL_REGEX, ignore_case = TRUE))) %>%
    group_by(cluster_label, direction) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup()

  top_labels <- bind_rows(auto_labels, grep_labels) %>%
    distinct(cluster_label, Description, .keep_all = TRUE)

  # FIX: sig_tier levels use the corrected Unicode labels (≤ and –)
  shape_values <- c(
    "FDR \u2264 0.05"    = 16,
    "FDR 0.05\u20130.10" = 17,
    "Not significant"    = 1
  )

  ggplot() +
    geom_point(
      data = nonsig_df,
      aes(x = NES, y = neg_log10_padj),
      color = "grey82", size = 0.6, alpha = 0.35
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_hline(
      yintercept = -log10(FDR_LOOSE),
      linetype = "dashed", color = "grey55", linewidth = 0.4
    ) +
    geom_point(
      data = sig_df,
      aes(
        x = NES, y = neg_log10_padj,
        color = direction, shape = sig_tier, size = neg_log10_padj
      ),
      alpha = 0.9
    ) +
    geom_text_repel(
      data = top_labels,
      aes(x = NES, y = neg_log10_padj, label = pathway_clean, color = direction),
      size = 2.5,
      lineheight = 0.85,
      max.overlaps = 80,
      box.padding = 0.45,
      point.padding = 0.3,
      segment.size = 0.25,
      segment.alpha = 0.6,
      force = 2.5,
      min.segment.length = 0,
      show.legend = FALSE
    ) +
    facet_wrap(~ cluster_label, scales = "free_y", ncol = ncol) +
    # FIX: dynamic color scale matching actual direction label values
    scale_color_manual(values = direction_colors, name = "Direction") +
    # FIX: shape legend uses correct Unicode label values matching sig_tier
    scale_shape_manual(values = shape_values, name = "FDR tier") +
    scale_size_continuous(
      name = expression(-log[10](adj.~p)),
      range = c(0.8, 4.5)
    ) +
    labs(
      title = figure_title,
      subtitle = figure_subtitle,
      # FIX: x-axis now reflects the actual contrast, not "Severe vs Mild Untreated"
      x = x_axis_label,
      y = expression(-log[10](adjusted~p))
    ) +
    theme_pub(base_size = 13)
}

make_deg_pair_scatter <- function(df,
                                  x_label,
                                  y_label,
                                  figure_title,
                                  figure_subtitle,
                                  celltype_order = NULL,
                                  celltype_labels = NULL,
                                  show_nonsig = TRUE,
                                  ncol = NCOL_FACET) {

  if (is.null(df) || nrow(df) == 0) return(NULL)

  df <- add_celltype_labels(df, celltype_order, celltype_labels)

  sig_df <- df %>% filter(sig_status != "Neither significant")
  nonsig_df <- df %>% filter(sig_status == "Neither significant")

  top_labels <- sig_df %>%
    group_by(cluster_label, quadrant) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup() %>%
    distinct(cluster_label, gene, .keep_all = TRUE)

  cor_annots <- df %>%
    group_by(cluster_label) %>%
    summarise(
      pearson_r = suppressWarnings(cor(lfc_x, lfc_y, method = "pearson")),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(!is.na(pearson_r))

  ggplot() +
    {
      if (show_nonsig && nrow(nonsig_df) > 0)
        geom_point(
          data = nonsig_df,
          aes(x = lfc_x, y = lfc_y),
          color = "grey82",
          size = 0.6,
          alpha = 0.35,
          shape = 16
        )
    } +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "grey30", linewidth = 0.5) +
    geom_point(
      data = sig_df,
      aes(
        x = lfc_x,
        y = lfc_y,
        color = quadrant,
        shape = sig_status,
        alpha = sig_tier,
        size = neg_log10_padj
      )
    ) +
    geom_text_repel(
      data = top_labels,
      aes(x = lfc_x, y = lfc_y, label = gene_label, color = quadrant),
      size = 2.5,
      max.overlaps = 80,
      box.padding = 0.45,
      point.padding = 0.3,
      segment.size = 0.25,
      segment.alpha = 0.6,
      force = 2.5,
      show.legend = FALSE
    ) +
    geom_text(
      data = cor_annots,
      aes(x = -Inf, y = Inf, label = sprintf("r = %.2f  n = %d", pearson_r, n)),
      hjust = -0.08,
      vjust = 1.5,
      size = 3.0,
      color = "grey30",
      fontface = "italic",
      inherit.aes = FALSE
    ) +
    facet_wrap(~ cluster_label, scales = "free", ncol = ncol) +
    scale_color_manual(values = quadrant_colors, name = "Direction") +
    scale_shape_manual(values = sig_shapes, name = "Significance", drop = FALSE) +
    scale_alpha_manual(values = sig_alpha, name = "FDR tier") +
    scale_size_continuous(
      name = expression(-log[10](mean~adj.~p)),
      range = c(0.8, 4.5),
      breaks = c(1, 2, 3, 5)
    ) +
    labs(
      title = figure_title,
      subtitle = figure_subtitle,
      x = x_label,
      y = y_label
    ) +
    theme_pub(base_size = 13)
}

# FIX: direction color scale and x-axis label now derived from data (contrast_label column)
# instead of being hardcoded to "Severe vs Mild Untreated"
make_deg_volcano <- function(df,
                             figure_title,
                             figure_subtitle,
                             celltype_order = NULL,
                             celltype_labels = NULL,
                             ncol = NCOL_FACET) {

  if (is.null(df) || nrow(df) == 0) return(NULL)

  df <- add_celltype_labels(df, celltype_order, celltype_labels)

  # Derive axis label and color scale from the data's contrast_label column
  contrast_label <- unique(df$contrast_label)[1]
  up_label   <- paste0("Up in ", contrast_label)
  down_label <- paste0("Down in ", contrast_label)
  x_axis_label <- paste0(contrast_label, " \u2014 logFC")

  direction_colors <- c(
    setNames("#C0392B", up_label),
    setNames("#2980B9", down_label),
    "Neutral" = "grey75"
  )

  # FIX: sig_tier levels use the corrected Unicode labels (≤ and –)
  shape_values <- c(
    "FDR \u2264 0.05"    = 16,
    "FDR 0.05\u20130.10" = 17,
    "Not significant"    = 1
  )

  sig_df    <- df %>% filter(adj.P.Val <= FDR_LOOSE)
  nonsig_df <- df %>% filter(adj.P.Val > FDR_LOOSE)

  top_labels <- sig_df %>%
    group_by(cluster_label, direction) %>%
    slice_max(label_strength, n = TOP_N, with_ties = FALSE) %>%
    ungroup()

  ggplot() +
    geom_point(
      data = nonsig_df,
      aes(x = logFC, y = neg_log10_padj),
      color = "grey82", size = 0.6, alpha = 0.35
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.4) +
    geom_hline(
      yintercept = -log10(FDR_LOOSE),
      linetype = "dashed", color = "grey55", linewidth = 0.4
    ) +
    geom_point(
      data = sig_df,
      aes(
        x = logFC, y = neg_log10_padj,
        color = direction, shape = sig_tier, size = neg_log10_padj
      ),
      alpha = 0.9
    ) +
    geom_text_repel(
      data = top_labels,
      aes(x = logFC, y = neg_log10_padj, label = gene, color = direction),
      size = 2.5,
      max.overlaps = 80,
      box.padding = 0.45,
      point.padding = 0.3,
      segment.size = 0.25,
      segment.alpha = 0.6,
      force = 2.5,
      show.legend = FALSE
    ) +
    facet_wrap(~ cluster_label, scales = "free_y", ncol = ncol) +
    # FIX: dynamic color scale matching actual direction label values
    scale_color_manual(values = direction_colors, name = "Direction") +
    # FIX: shape legend uses correct Unicode label values matching sig_tier
    scale_shape_manual(values = shape_values, name = "FDR tier") +
    scale_size_continuous(
      name = expression(-log[10](adj.~p)),
      range = c(0.8, 4.5)
    ) +
    labs(
      title = figure_title,
      subtitle = figure_subtitle,
      # FIX: x-axis now reflects the actual contrast, not "Severe vs Mild Untreated"
      x = x_axis_label,
      y = expression(-log[10](adjusted~p))
    ) +
    theme_pub(base_size = 13)
}

# ── NEW: Summary count functions ──────────────────────────────────────────────

#' Summarise significant pathway counts per cell type × contrast × database
#' Returns a wide table: rows = cell types, columns = contrast_db combinations
summarise_pathway_counts <- function(celltypes, contrasts, dbs,
                                     fdr_threshold = FDR_LOOSE) {

  rows <- list()

  for (ct in celltypes) {
    ct_clean <- clean_cluster_names(ct)
    row <- list(cell_type = CELLTYPE_LABELS[ct_clean] %||% ct_clean)

    for (contrast in contrasts) {
      for (db in dbs) {
        dx <- read_safe(gsea_path(ct, contrast, db))
        col_name <- paste0(
          contrast %>%
            str_replace_all("_vs_", " vs ") %>%
            str_replace_all("MildUntreated", "Mild"),
          " | ", db
        )
        if (!is.null(dx) && "p.adjust" %in% colnames(dx)) {
          row[[col_name]] <- sum(dx$p.adjust <= fdr_threshold, na.rm = TRUE)
        } else {
          row[[col_name]] <- NA_integer_
        }
      }
    }

    rows[[ct]] <- as.data.frame(row, check.names = FALSE)
  }

  if (length(rows) == 0) return(NULL)
  bind_rows(rows)
}

#' Summarise significant DEG counts per cell type × contrast
#' Returns a wide table: rows = cell types, columns = contrasts
summarise_deg_counts <- function(celltypes, contrasts,
                                  fdr_threshold = FDR_LOOSE) {

  rows <- list()

  for (ct in celltypes) {
    ct_clean <- clean_cluster_names(ct)
    row <- list(cell_type = CELLTYPE_LABELS[ct_clean] %||% ct_clean)

    for (contrast in contrasts) {
      dx <- read_safe(deg_path(ct, contrast))
      col_name <- contrast %>%
        str_replace_all("_vs_", " vs ") %>%
        str_replace_all("MildUntreated", "Mild Untreated")

      if (!is.null(dx) && "adj.P.Val" %in% colnames(dx)) {
        n_up   <- sum(dx$adj.P.Val <= fdr_threshold & dx$logFC > 0, na.rm = TRUE)
        n_down <- sum(dx$adj.P.Val <= fdr_threshold & dx$logFC < 0, na.rm = TRUE)
        row[[paste0(col_name, " (Up)")]]   <- n_up
        row[[paste0(col_name, " (Down)")]] <- n_down
        row[[paste0(col_name, " (Total)")]] <- n_up + n_down
      } else {
        row[[paste0(col_name, " (Up)")]]    <- NA_integer_
        row[[paste0(col_name, " (Down)")]]  <- NA_integer_
        row[[paste0(col_name, " (Total)")]] <- NA_integer_
      }
    }

    rows[[ct]] <- as.data.frame(row, check.names = FALSE)
  }

  if (length(rows) == 0) return(NULL)
  bind_rows(rows)
}

# Operator for default values (like Python's 'or')
`%||%` <- function(a, b) if (!is.null(a) && !is.na(a) && nchar(a) > 0) a else b
```

# Run Analysis and Embed Plots

```{r}
#| label: run-analysis
#| fig.show: hold
#| fig.width: 22
#| fig.height: 18
#| results: markup

cat("Discovering cell types...\n")

celltypes <- tryCatch(
  discover_celltypes(),
  error = function(e) {
    message(e)
    character(0)
  }
)

cat("Found:", length(celltypes), "\n\n")

if (length(celltypes) == 0) {
  stop("No cell types found — check BASE_DIR.")
}

PATHWAY_FIGURES <- list(
  list(
    fig = "Figure1",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_HC",
    x_label = "Mild Untreated vs HC \u2014 NES",
    y_label = "Severe vs HC \u2014 NES",
    title = "Pathway enrichment: Mild Untreated vs HC \u00d7 Severe vs HC",
    subtitle = paste0(
      "Each point = one pathway in one cell type \u2022 FDR \u2264 0.10 \u2022 ",
      "Point size = \u2212log\u2081\u2080(mean adj. p-value)\n",
      "Shows whether disease-associated pathway changes are shared between Mild and Severe relative to HC"
    )
  ),
  list(
    fig = "Figure2",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Mild Untreated vs HC \u2014 NES",
    y_label = "Severe vs Mild Untreated \u2014 NES",
    title = "Pathway enrichment: Mild Untreated vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = paste0(
      "Compares early disease changes with progression changes\n",
      "Q2/Q4 highlight pathways that reverse or newly emerge during severe disease progression"
    )
  ),
  list(
    fig = "Figure3",
    contrast_x = "Severe_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Severe vs HC \u2014 NES",
    y_label = "Severe vs Mild Untreated \u2014 NES",
    title = "Pathway enrichment: Severe vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = paste0(
      "Compares severe disease relative to healthy controls and mild untreated patients\n",
      "High concordance supports robust severe disease-associated dysregulation"
    )
  ),
  list(
    fig = "Figure4",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Mild Untreated vs HC \u2014 NES",
    y_label = "Severe vs Mild Untreated \u2014 NES",
    title = "Disease progression: Mild vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = paste0(
      "Compares early disease changes (HC\u2192Mild) with progression changes (Mild\u2192Severe)\n",
      "Q1/Q3 = same direction across onset and progression; Q2/Q4 = reversed or newly emerging pathways"
    )
  )
)

for (db in PATHWAY_DBS) {
  for (fig in PATHWAY_FIGURES) {

    cat("===", fig$fig, db, "===\n")

    fig_db <- load_gsea_pair(
      celltypes,
      contrast_x = fig$contrast_x,
      contrast_y = fig$contrast_y,
      db = db
    )

    if (!is.null(fig_db)) {
      p <- make_pub_scatter(
        df = fig_db,
        x_label = fig$x_label,
        y_label = fig$y_label,
        figure_title = paste(fig$fig, "|", db, fig$title),
        figure_subtitle = fig$subtitle,
        celltype_order = CELLTYPE_ORDER,
        celltype_labels = CELLTYPE_LABELS,
        ncol = NCOL_FACET
      )

      print(p)

      out_name <- paste0(
        fig$fig, "_", db, "_",
        fig$contrast_x, "_x_", fig$contrast_y
      )

      save_png(p, out_name, height = fig_height(fig_db))

      data.table::fwrite(
        fig_db,
        file.path(OUT_DIR, paste0(out_name, "_data.csv"))
      )
    }
  }
}

DIRECT_CONTRASTS <- list(
  list(
    fig = "Figure5A",
    contrast = "MildUntreated_vs_HC",
    title = "Direct Mild Untreated vs HC — Pathway Enrichment",
    subtitle = "Single-contrast view of pathways enriched or depleted in Mild Untreated patients compared with Healthy Controls"
  ),
  list(
    fig = "Figure5B",
    contrast = "Severe_vs_HC",
    title = "Direct Severe vs HC — Pathway Enrichment",
    subtitle = "Single-contrast view of pathways enriched or depleted in Severe patients compared with Healthy Controls"
  ),
  list(
    fig = "Figure5C",
    contrast = "Severe_vs_MildUntreated",
    title = "Direct Severe vs Mild Untreated — Pathway Enrichment",
    subtitle = "Single-contrast view of pathways enriched or depleted in Severe patients compared with Mild Untreated patients"
  )
)

for (db in PATHWAY_DBS) {
  for (direct_fig in DIRECT_CONTRASTS) {

    cat("===", direct_fig$fig, direct_fig$contrast, db, "===\n")

    direct_db <- load_gsea_single(
      celltypes,
      contrast = direct_fig$contrast,
      db = db
    )

    if (!is.null(direct_db)) {
      p_direct <- make_single_gsea_plot(
        df = direct_db,
        figure_title = paste(direct_fig$fig, "|", direct_fig$title, "|", db),
        figure_subtitle = paste0(
          direct_fig$subtitle, "\n",
          "Red = positive NES (up in condition); blue = negative NES (down in condition); ",
          "dashed horizontal line = FDR 0.10"
        ),
        celltype_order = CELLTYPE_ORDER,
        celltype_labels = CELLTYPE_LABELS,
        ncol = NCOL_FACET
      )

      print(p_direct)

      out_name <- paste0(direct_fig$fig, "_", db, "_Direct_", direct_fig$contrast)

      save_png(p_direct, out_name, height = fig_height(direct_db))

      data.table::fwrite(
        direct_db,
        file.path(OUT_DIR, paste0(out_name, "_data.csv"))
      )
    }
  }
}

DEG_PAIR_FIGURES <- list(
  list(
    fig = "DEG_Figure1",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_HC",
    x_label = "Mild Untreated vs HC \u2014 logFC",
    y_label = "Severe vs HC \u2014 logFC",
    title = "DEG comparison: Mild Untreated vs HC \u00d7 Severe vs HC",
    subtitle = "Each point = one gene in one cell type. Shows whether gene-level changes are shared between Mild and Severe relative to HC."
  ),
  list(
    fig = "DEG_Figure2",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Mild Untreated vs HC \u2014 logFC",
    y_label = "Severe vs Mild Untreated \u2014 logFC",
    title = "DEG comparison: Mild Untreated vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = "Compares early disease gene changes with progression gene changes from Mild to Severe."
  ),
  list(
    fig = "DEG_Figure3",
    contrast_x = "Severe_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Severe vs HC \u2014 logFC",
    y_label = "Severe vs Mild Untreated \u2014 logFC",
    title = "DEG comparison: Severe vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = "Compares severe-associated gene changes relative to HC and Mild Untreated patients."
  ),
  list(
    fig = "DEG_Figure4",
    contrast_x = "MildUntreated_vs_HC",
    contrast_y = "Severe_vs_MildUntreated",
    x_label = "Mild Untreated vs HC \u2014 logFC",
    y_label = "Severe vs Mild Untreated \u2014 logFC",
    title = "DEG progression model: Mild vs HC \u00d7 Severe vs Mild Untreated",
    subtitle = "Q1/Q3 = same gene direction across onset and progression; Q2/Q4 = reversed or newly emerging gene changes."
  )
)

for (fig in DEG_PAIR_FIGURES) {

  cat("===", fig$fig, "===\n")

  deg_pair_db <- load_deg_pair(
    celltypes,
    contrast_x = fig$contrast_x,
    contrast_y = fig$contrast_y
  )

  if (!is.null(deg_pair_db)) {
    p_deg_pair <- make_deg_pair_scatter(
      df = deg_pair_db,
      x_label = fig$x_label,
      y_label = fig$y_label,
      figure_title = fig$title,
      figure_subtitle = fig$subtitle,
      celltype_order = CELLTYPE_ORDER,
      celltype_labels = CELLTYPE_LABELS,
      ncol = NCOL_FACET
    )

    print(p_deg_pair)

    out_name <- paste0(
      fig$fig, "_",
      fig$contrast_x, "_x_", fig$contrast_y
    )

    save_png(p_deg_pair, out_name, height = fig_height(deg_pair_db))

    data.table::fwrite(
      deg_pair_db,
      file.path(OUT_DIR, paste0(out_name, "_data.csv"))
    )
  }
}

DEG_DIRECT_FIGURES <- list(
  list(
    fig = "DEG_Direct_MildvsHC",
    contrast = "MildUntreated_vs_HC",
    title = "DEG | Direct Mild Untreated vs HC gene-level comparison",
    subtitle = "Each point = one gene \u2022 FDR \u2264 0.10\nRed = upregulated in Mild Untreated; blue = downregulated in Mild Untreated"
  ),
  list(
    fig = "DEG_Direct_SeverevsHC",
    contrast = "Severe_vs_HC",
    title = "DEG | Direct Severe vs HC gene-level comparison",
    subtitle = "Each point = one gene \u2022 FDR \u2264 0.10\nRed = upregulated in Severe; blue = downregulated in Severe"
  ),
  list(
    fig = "DEG_Direct_SeverevsMild",
    contrast = "Severe_vs_MildUntreated",
    title = "DEG | Direct Severe vs Mild Untreated gene-level comparison",
    subtitle = "Each point = one gene \u2022 FDR \u2264 0.10\nRed = upregulated in Severe; blue = downregulated in Severe"
  )
)

for (direct_deg in DEG_DIRECT_FIGURES) {

  cat("===", direct_deg$fig, direct_deg$contrast, "===\n")

  deg_direct_db <- load_deg_single(
    celltypes,
    contrast = direct_deg$contrast
  )

  if (!is.null(deg_direct_db)) {
    p_deg_direct <- make_deg_volcano(
      df = deg_direct_db,
      figure_title = direct_deg$title,
      figure_subtitle = direct_deg$subtitle,
      celltype_order = CELLTYPE_ORDER,
      celltype_labels = CELLTYPE_LABELS
    )

    print(p_deg_direct)

    save_png(
      p_deg_direct,
      direct_deg$fig,
      height = fig_height(deg_direct_db)
    )

    data.table::fwrite(
      deg_direct_db,
      file.path(OUT_DIR, paste0(direct_deg$fig, "_data.csv"))
    )
  }
}

cat("\n===================================================\n")
cat("All figures saved to:", OUT_DIR, "\n")
cat("Files produced:\n")
cat("  Figure1\u2013Figure4 pathway plots for Hallmark, Reactome, and GO_BP\n")
cat("  Figure5A\u20135C direct pathway plots for Mild vs HC, Severe vs HC, and Severe vs Mild across Hallmark, Reactome, and GO_BP\n")
cat("  DEG_Figure1\u2013DEG_Figure4 pairwise comparison plots\n")
cat("  DEG direct volcano plots for Mild vs HC, Severe vs HC, and Severe vs Mild\n")
cat("  *_data.csv underlying data files\n")
cat("===================================================\n")
```

# Significant Count Summaries

## DEG Counts per Cell Type and Contrast

Number of differentially expressed genes (FDR ≤ 0.10) per cell type across all three contrasts, split by direction (Up = higher in the numerator group; Down = lower in the numerator group).

```{r}
#| label: deg-count-table
#| results: asis

ALL_CONTRASTS <- c("MildUntreated_vs_HC", "Severe_vs_HC", "Severe_vs_MildUntreated")

deg_counts <- summarise_deg_counts(celltypes, ALL_CONTRASTS, fdr_threshold = FDR_LOOSE)

if (!is.null(deg_counts) && nrow(deg_counts) > 0) {

  # Save to CSV
  data.table::fwrite(
    deg_counts,
    file.path(OUT_DIR, "Summary_DEG_counts_by_celltype_contrast.csv")
  )

  # Display "Total" columns only in the rendered table for conciseness;
  # the CSV has Up/Down/Total columns
  total_cols <- grep("\\(Total\\)$", colnames(deg_counts), value = TRUE)
  display_deg <- deg_counts %>%
    dplyr::select(cell_type, all_of(total_cols)) %>%
    rename_with(
      ~ str_remove(.x, " \\(Total\\)") %>%
          str_replace_all("MildUntreated", "Mild Untreated"),
      -cell_type
    )

  cat(
    kable(
      display_deg,
      caption = paste0(
        "Significant DEGs per cell type and contrast (FDR \u2264 ", FDR_LOOSE, "). ",
        "Values = total significant genes (Up + Down). ",
        "Full Up/Down breakdown saved to Summary_DEG_counts_by_celltype_contrast.csv."
      ),
      format = "html",
      align = c("l", rep("r", ncol(display_deg) - 1))
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed", "responsive"),
        full_width = TRUE,
        font_size = 12
      ) %>%
      column_spec(1, bold = TRUE, width = "18em")
  )

} else {
  cat("_No DEG data found — check BASE_DIR and contrast file paths._\n")
}
```

### DEG Direction Breakdown (Up / Down / Total)

```{r}
#| label: deg-direction-table
#| results: asis

if (!is.null(deg_counts) && nrow(deg_counts) > 0) {

  cat(
    kable(
      deg_counts,
      caption = paste0(
        "Significant DEGs per cell type and contrast (FDR \u2264 ", FDR_LOOSE, ") — full Up/Down breakdown."
      ),
      format = "html",
      align = c("l", rep("r", ncol(deg_counts) - 1))
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed", "responsive"),
        full_width = TRUE,
        font_size = 11
      ) %>%
      column_spec(1, bold = TRUE, width = "18em") %>%
      scroll_box(width = "100%", height = "500px")
  )
}
```

## Pathway Counts per Cell Type, Contrast, and Database

Number of significantly enriched pathways (FDR ≤ 0.10) per cell type across all contrasts and pathway databases.

```{r}
#| label: pathway-count-table
#| results: asis

pathway_counts <- summarise_pathway_counts(
  celltypes, ALL_CONTRASTS, PATHWAY_DBS, fdr_threshold = FDR_LOOSE
)

if (!is.null(pathway_counts) && nrow(pathway_counts) > 0) {

  # Save to CSV
  data.table::fwrite(
    pathway_counts,
    file.path(OUT_DIR, "Summary_pathway_counts_by_celltype_contrast_db.csv")
  )

  cat(
    kable(
      pathway_counts,
      caption = paste0(
        "Significant pathways per cell type, contrast, and database (FDR \u2264 ", FDR_LOOSE, "). ",
        "Column format: Contrast | Database. ",
        "Full table also saved to Summary_pathway_counts_by_celltype_contrast_db.csv."
      ),
      format = "html",
      align = c("l", rep("r", ncol(pathway_counts) - 1))
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed", "responsive"),
        full_width = TRUE,
        font_size = 11
      ) %>%
      column_spec(1, bold = TRUE, width = "18em") %>%
      scroll_box(width = "100%", height = "500px")
  )

} else {
  cat("_No pathway data found — check BASE_DIR and gsea file paths._\n")
}
```

## Cross-contrast Summary (Totals Across All Cell Types)

```{r}
#| label: grand-total-table
#| results: asis

# Grand totals across all cell types per contrast
if (!is.null(deg_counts) && nrow(deg_counts) > 0) {

  up_cols    <- grep("\\(Up\\)$",    colnames(deg_counts), value = TRUE)
  down_cols  <- grep("\\(Down\\)$",  colnames(deg_counts), value = TRUE)
  total_cols <- grep("\\(Total\\)$", colnames(deg_counts), value = TRUE)

  deg_grand <- tibble(
    Contrast = ALL_CONTRASTS %>%
      str_replace_all("_vs_", " vs ") %>%
      str_replace_all("MildUntreated", "Mild Untreated"),
    `Total Up`   = colSums(deg_counts[up_cols],    na.rm = TRUE),
    `Total Down` = colSums(deg_counts[down_cols],  na.rm = TRUE),
    `Total DEGs` = colSums(deg_counts[total_cols], na.rm = TRUE),
    `Cell Types with ≥1 DEG` = sapply(total_cols, function(col)
      sum(deg_counts[[col]] > 0, na.rm = TRUE))
  )

  cat(
    kable(
      deg_grand,
      caption = paste0(
        "Grand total significant DEGs across all cell types per contrast (FDR \u2264 ", FDR_LOOSE, ")."
      ),
      format = "html",
      align = c("l", "r", "r", "r", "r")
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE,
        font_size = 13
      )
  )
}

if (!is.null(pathway_counts) && nrow(pathway_counts) > 0) {

  pw_cols <- colnames(pathway_counts)[-1]  # exclude cell_type

  pw_grand <- tibble(
    `Contrast | Database` = pw_cols,
    `Total Significant Pathways` = colSums(pathway_counts[pw_cols], na.rm = TRUE),
    `Cell Types with ≥1 Pathway` = sapply(pw_cols, function(col)
      sum(pathway_counts[[col]] > 0, na.rm = TRUE))
  )

  cat("\n\n")

  cat(
    kable(
      pw_grand,
      caption = paste0(
        "Grand total significant pathways across all cell types per contrast and database (FDR \u2264 ", FDR_LOOSE, ")."
      ),
      format = "html",
      align = c("l", "r", "r")
    ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE,
        font_size = 13
      )
  )
}
```

# Key Interpretation

The pathway-level analysis compares disease-associated programs across mild and severe disease. Concordant quadrants indicate pathways that are consistently altered across disease states, while discordant quadrants highlight pathways that reverse or newly emerge during progression.

The DEG-level analysis provides gene-level support for the pathway findings by identifying specific genes that are upregulated or downregulated across the same contrasts.

# Output Files

```{r}
#| label: output-files

list.files(OUT_DIR, pattern = "\\.(png|csv)$", full.names = FALSE)
```

# Session Info

```{r}
sessionInfo()
```
