---
title: "Paired Treatment Pathway Response Report"
subtitle: "Treated vs Untreated pathways, interpreted against the Yu2021 healthy reference"
date: today
format:
  html:
    embed-resources: true
    theme: flatly
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    fig-align: center
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
params:
  mapped_csv: "deg_gsea_YuHealthy_Internal_paired_limma_fgsea/06_PAIRED_TREATMENT_PATHWAY_SUMMARY/paired_significant_pathways_mapped_to_external_references.csv"
  all_gsea_csv: "deg_gsea_YuHealthy_Internal_paired_limma_fgsea/04_SUMMARY/all_GSEA_results.csv"
 ## fdr_cutoff: 0.10
  fdr_cutoff: 0.05
  top_n_per_celltype: 15
  top_n_overall_heatmap: 40
  jakstat_pattern: "INTERFERON|IFN_|IFNG|IFNA|JAK|STAT[0-9]|CYTOKINE|IL2_STAT5|IL6_JAK|INFLAMMATORY_RESPONSE|TNFA_SIGNALING|TNF|NFKB|NF_KB|OXIDATIVE_PHOSPHORYLATION|RESPIRATORY_ELECTRON|AEROBIC_RESPIRATION|ADAPTIVE_IMMUNE|T_CELL_RECEPTOR|LYMPHOCYTE|LEUKOCYTE|CELL_KILLING"
  top_n_recurring: 15
  featured_celltype: "Macrophage"
  n_per_curated_category: 5
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DT)
  library(plotly)
  library(forcats)
  library(scales)
  library(htmltools)
  library(ggrepel)
})

direction_levels <- c(
  "Disease-high, suppressed by treatment",
  "Disease-low, restored by treatment",
  "Disease-high, further increased",
  "Disease-low, further decreased",
  "Treatment-associated; no external disease direction"
)

direction_colors <- c(
  "Disease-high, suppressed by treatment" = "#009E73",
  "Disease-low, restored by treatment"    = "#56B4E9",
  "Disease-high, further increased"       = "#D55E00",
  "Disease-low, further decreased"        = "#CC79A7",
  "Treatment-associated; no external disease direction" = "grey55"
)

theme_report <- function(base_size = 11) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = 9, color = "grey30"),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )
}

# reorder_within / scale_y_reordered: lightweight local reimplementation
# (avoids a tidytext dependency) so categorical axes can be sorted
# independently within each facet panel.
reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new_x <- paste(x, within, sep = sep)
  stats::reorder(new_x, by, FUN = fun, ...)
}

scale_y_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
}
```

```{r load-data}
if (!file.exists(params$mapped_csv)) {
  stop("Mapped pathway CSV not found: ", params$mapped_csv)
}

mapped <- read_csv(params$mapped_csv, show_col_types = FALSE) %>%
  mutate(
    directional_response = factor(directional_response, levels = direction_levels),
    celltype = as.character(celltype),
    database = as.character(database)
  )

required_columns <- c(
  "celltype", "database", "pathway", "pathway_label",
  "NES_UH", "padj_UH", "NES_TH", "padj_TH", "NES_TU", "padj_TU",
  "directional_response", "broad_direction",
  "paired_fdr_strength", "direction_strength"
)

missing_columns <- setdiff(required_columns, colnames(mapped))
if (length(missing_columns) > 0) {
  stop("Mapped CSV is missing expected columns: ", paste(missing_columns, collapse = ", "))
}

n_celltypes <- n_distinct(mapped$celltype)
n_databases <- n_distinct(mapped$database)
n_pathways_total <- nrow(mapped)
```

```{r load-raw-universe}
# Full, unfiltered Treated_vs_Untreated results (every pathway tested, not just
# the paired-significant ones). This is required for the enrichment and
# recurrence tests below — testing "is this cluster of hits more than chance"
# needs the full background of what was tested, not just the hits themselves.
has_raw_universe <- file.exists(params$all_gsea_csv)

if (has_raw_universe) {
  raw_all <- read_csv(params$all_gsea_csv, show_col_types = FALSE)

  clean_pathway_label_local <- function(x) {
    x %>%
      as.character() %>%
      str_remove("^HALLMARK_") %>%
      str_remove("^REACTOME_") %>%
      str_remove("^GOBP_") %>%
      str_remove("^GO_") %>%
      str_replace_all("_", " ") %>%
      str_to_title()
  }

  raw_tu <- raw_all %>%
    filter(contrast == "Treated_vs_Untreated") %>%
    transmute(
      celltype = as.character(celltype),
      database = as.character(database),
      pathway = as.character(pathway),
      NES_TU = as.numeric(NES),
      padj_TU = as.numeric(padj),
      pval_TU = if ("pval" %in% names(.)) as.numeric(pval) else NA_real_
    ) %>%
    filter(!is.na(celltype), !is.na(database), !is.na(pathway)) %>%
    distinct(celltype, database, pathway, .keep_all = TRUE)

  has_raw_pvals <- any(!is.na(raw_tu$pval_TU))

  # Full Untreated_vs_Healthy universe — needed for the disease-associated
  # pathway counts and recovery-after-treatment section below, which (unlike
  # the rest of this report) is anchored on UH significance to mirror an
  # external reference figure's design.
  raw_uh <- raw_all %>%
    filter(contrast == "Untreated_vs_Healthy") %>%
    transmute(
      celltype = as.character(celltype),
      database = as.character(database),
      pathway = as.character(pathway),
      pathway_label = clean_pathway_label_local(pathway),
      NES_UH = as.numeric(NES),
      padj_UH = as.numeric(padj)
    ) %>%
    filter(!is.na(celltype), !is.na(database), !is.na(pathway), is.finite(NES_UH)) %>%
    distinct(celltype, database, pathway, .keep_all = TRUE)

  # NOTE: despite the variable name (kept for minimal downstream changes),
  # this is NOT filtered by padj_UH -- Untreated vs Healthy uses NES sign
  # only, with no significance filter, since it is an unpaired comparison
  # against only 2 external healthy donors. The FDR filter is applied on the
  # Treated vs Untreated side instead (see recovery-join chunk), which is
  # the paired, well-powered comparison.
  uh_significant <- raw_uh %>%
    mutate(uh_direction = if_else(NES_UH > 0, "Up", "Down"))

} else {
  has_raw_pvals <- FALSE
}
```

# How to read this report {.unnumbered}

Every pathway shown here passed one filter only: **statistically significant in the
paired `Treated vs Untreated` comparison** (FDR ≤ `r params$fdr_cutoff`). That
comparison is the sole basis for inclusion and for statistical inference.

The Yu2021 healthy cohort is not part of that statistical test. It's used purely as an
**external directional reference** — an independent dataset that tells us roughly what
"looks healthy" for a given pathway, so we can ask whether the treatment effect we
already found to be significant moves a pathway *toward* or *away from* that reference.

| Untreated vs Healthy | Treated vs Untreated | Interpretation |
|---|---|---|
| NES > 0 (elevated in disease) | NES < 0 (treatment decreases it) | **Suppressed by treatment** — disease-high pathway pulled down |
| NES < 0 (depleted in disease) | NES > 0 (treatment increases it) | **Restored by treatment** — disease-low pathway brought back up |
| NES > 0 | NES > 0 | Disease-high pathway pushed **further up** |
| NES < 0 | NES < 0 | Disease-low pathway pushed **further down** |
| not available | — | No external reference; treatment effect only |

::: {.callout-note}
## Direction, not destination

The interpretation labels above (e.g. "restored by treatment") are based **only on
sign**: did the paired treatment effect move a pathway toward healthy? They say
nothing about how close it got. `Treated vs Healthy` is also an independent, unpaired
comparison against a small reference cohort (2 donors), so short arrows can still
reflect a real, statistically robust paired effect — they just haven't been benchmarked
against an absolute "healthy" cutoff, because the data don't support drawing one
reliably. The paired `Treated vs Untreated` NES and FDR (visible in the table and
quadrant plot below) are the rigorously tested quantities; treat the trajectory plot as
showing direction of movement, not proximity to normal.
:::

```{r overview-summary}
overview_table <- mapped %>%
  count(celltype, database, name = "n_significant_pathways") %>%
  arrange(celltype, database)

tibble(
  Metric = c(
    "Paired-significant pathway x celltype x database rows",
    "Distinct cell types",
    "Distinct pathway databases",
    "Paired FDR cutoff used for inclusion"
  ),
  Value = c(
    comma(n_pathways_total),
    n_celltypes,
    n_databases,
    params$fdr_cutoff
  )
) %>%
  knitr::kable(caption = "Report scope")
```

# Overall pattern: toward vs. away from healthy

For each cell type, what fraction of its paired-significant pathways move toward the
external healthy reference versus further away from it?

```{r toward-away-summary, fig.width=11, fig.height=max(5, 0.35 * n_celltypes + 3)}
broad_summary <- mapped %>%
  count(celltype, database, broad_direction, name = "n_pathways") %>%
  group_by(celltype, database) %>%
  mutate(percentage = 100 * n_pathways / sum(n_pathways)) %>%
  ungroup()

p_broad <- ggplot(
  broad_summary,
  aes(x = celltype, y = percentage, fill = broad_direction,
      text = paste0(
        celltype, " — ", broad_direction, "<br>",
        n_pathways, " pathways (", round(percentage, 1), "%)"
      ))
) +
  geom_col() +
  coord_flip() +
  facet_wrap(~database, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "Moves toward external healthy direction" = "#1B9E77",
    "Moves farther from external healthy direction" = "#D95F02",
    "No external direction available" = "grey65"
  )) +
  labs(
    title = "Direction of paired-significant treatment effects, relative to Yu2021 healthy",
    x = NULL, y = "% of paired-significant pathways", fill = NULL
  ) +
  theme_report()

ggplotly(p_broad, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```

# Pathway trajectories: disease state → treatment effect {#sec-trajectory}

This is the most direct view of the question this analysis is asking. For each
paired-significant pathway, the **open circle** marks its position in
`Untreated vs Healthy` (how abnormal it was before treatment) and the **filled,
colored point** marks `Treated vs Healthy` (where it ended up after treatment). The
arrow shows the movement treatment caused. The dashed line at zero is the healthy
reference itself.

Only the top `r params$top_n_per_celltype` pathways per cell type (ranked by paired
`Treated vs Untreated` FDR) are shown, to keep each panel readable.

```{r trajectory-plot, fig.width=13, fig.height=max(7, 3.6 * ceiling(n_celltypes / 2))}
trajectory_data <- mapped %>%
  filter(!is.na(NES_UH), !is.na(NES_TH)) %>%
  group_by(celltype) %>%
  slice_min(padj_TU, n = params$top_n_per_celltype, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(pathway_ord = reorder_within(pathway_label, NES_TH, celltype))

p_trajectory <- ggplot(trajectory_data, aes(y = pathway_ord)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  geom_segment(
    aes(x = NES_UH, xend = NES_TH, yend = pathway_ord, color = directional_response),
    arrow = arrow(length = unit(0.10, "cm"), type = "closed"),
    linewidth = 0.8, alpha = 0.85
  ) +
  geom_point(aes(x = NES_UH), shape = 21, fill = "white", color = "grey30", size = 1.9) +
  geom_point(aes(x = NES_TH, color = directional_response), size = 2.6) +
  scale_y_reordered() +
  scale_color_manual(values = direction_colors, drop = FALSE, name = "Interpretation") +
  facet_wrap(~celltype, scales = "free_y", ncol = 2) +
  labs(
    title = "Pathway movement relative to the Yu2021 healthy reference",
    subtitle = "Open circle = Untreated vs Healthy · Filled circle = Treated vs Healthy · 0 = healthy-equivalent",
    x = "NES relative to Yu2021 healthy", y = NULL
  ) +
  theme_report() +
  theme(axis.text.y = element_text(size = 7))

print(p_trajectory)
```

# Quadrant view: treatment effect vs. baseline disease direction

The same paired-significant pathways, plotted by their raw disease direction
(`Untreated vs Healthy`, x-axis) against the paired treatment effect
(`Treated vs Untreated`, y-axis). Point size reflects paired-treatment significance
strength. The 5 strongest pathways in each quadrant (by paired FDR) are labeled.

```{r quadrant-scatter, fig.width=13, fig.height=max(7, 3.6 * ceiling(n_celltypes / 2))}
quadrant_data <- mapped %>%
  filter(!is.na(NES_UH), !is.na(NES_TU)) %>%
  mutate(
    quadrant = case_when(
      NES_UH > 0 & NES_TU > 0 ~ "UH+ / TU+",
      NES_UH > 0 & NES_TU < 0 ~ "UH+ / TU-",
      NES_UH < 0 & NES_TU > 0 ~ "UH- / TU+",
      NES_UH < 0 & NES_TU < 0 ~ "UH- / TU-",
      TRUE ~ NA_character_
    )
  )

label_data <- quadrant_data %>%
  filter(!is.na(quadrant)) %>%
  group_by(celltype, quadrant) %>%
  slice_min(padj_TU, n = 5, with_ties = FALSE) %>%
  ungroup()

p_quadrant <- ggplot(
  quadrant_data,
  aes(x = NES_UH, y = NES_TU, color = directional_response, size = paired_fdr_strength)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = label_data,
    aes(label = str_wrap(pathway_label, width = 28)),
    size = 2.4,
    max.overlaps = 100,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0,
    segment.size = 0.3,
    show.legend = FALSE,
    color = "grey15"
  ) +
  scale_color_manual(values = direction_colors, drop = FALSE, name = "Interpretation") +
  scale_size_continuous(name = expression(-log[10](paired~FDR)), range = c(0.6, 3.5)) +
  facet_wrap(~celltype, ncol = 2) +
  labs(
    title = "Paired treatment effect vs. baseline disease direction",
    subtitle = "Top 5 pathways per quadrant (by paired FDR) labeled",
    x = "Untreated vs Healthy NES", y = "Treated vs Untreated NES"
  ) +
  theme_report() +
  theme(legend.box = "vertical")

print(p_quadrant)
```

# Strongest hits across all cell types

The top `r params$top_n_overall_heatmap` paired-significant pathways overall, ranked by
paired FDR, shown across all three contrasts for direct comparison.

```{r top-heatmap, fig.width=11, fig.height=max(8, 0.28 * params$top_n_overall_heatmap + 3)}
top_overall <- mapped %>%
  arrange(padj_TU, desc(abs(NES_TU))) %>%
  slice_head(n = params$top_n_overall_heatmap) %>%
  mutate(pathway_celltype = paste0(pathway_label, "  (", celltype, ")"))

heatmap_long <- top_overall %>%
  select(pathway_celltype, NES_UH, NES_TH, NES_TU) %>%
  pivot_longer(-pathway_celltype, names_to = "comparison", values_to = "NES") %>%
  mutate(
    comparison = factor(
      comparison,
      levels = c("NES_UH", "NES_TH", "NES_TU"),
      labels = c("Untreated vs Healthy", "Treated vs Healthy", "Paired Treated vs Untreated")
    ),
    pathway_celltype = fct_reorder(
      pathway_celltype,
      top_overall$NES_TU[match(pathway_celltype, top_overall$pathway_celltype)]
    )
  )

p_top_heatmap <- ggplot(heatmap_long, aes(x = comparison, y = pathway_celltype, fill = NES)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_text(aes(label = ifelse(is.na(NES), "", sprintf("%.2f", NES))), size = 2.4) +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC", midpoint = 0, na.value = "grey90") +
  labs(
    title = "Top paired-significant pathways across all cell types",
    x = NULL, y = NULL, fill = "NES"
  ) +
  theme_report() +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    axis.text.y = element_text(size = 7)
  )

print(p_top_heatmap)
```

# Statistical Tests: Could This Pattern Be Random?

Interferon/cytokine pathways showing up repeatedly, across many independent cell
types, in the direction expected for a JAK inhibitor is a striking pattern — but
"striking" isn't a statistic. Two independent tests below formalize the question:
is this cluster of hits more concentrated in JAK-STAT-adjacent pathways, and more
recurrent across cell types, than would be expected if paired-significant hits were
scattered randomly across the tested pathway universe?

::: {.callout-note}
## The expected signature is broader than "interferon" alone

The cross-cell-type recurrence results (Test 2) surfaced two related but distinct
signals recurring across nearly every cell type: the anticipated
**JAK-STAT/interferon axis** (`Interferon Gamma/Alpha Response`,
`Interferon Signaling`, `Adaptive Immune Response`, `T Cell Receptor Signaling`), and
a second, mechanistically plausible **downstream/crosstalk signature** —
`Tnfa Signaling Via Nfkb` (the single most recurrent pathway overall, in 17/18 cell
types), plus `Oxidative Phosphorylation` and `Aerobic Respiration`. NF-\u03baB
activity is amplified by IL-6/JAK-STAT signaling in inflamed tissue, so its
suppression alongside JAK-STAT targets is expected pathway crosstalk rather than a
separate mechanism; the metabolic pathways are consistent with reduced immune-cell
activation state (activated immune cells shift toward glycolysis; quiescent cells
lean more oxidative). `params$jakstat_pattern` has been broadened accordingly so
Test 1's enrichment reflects this fuller expected-pharmacology signature rather than
interferon pathways alone.
:::

```{r stat-tests-guard, results='asis'}
if (!has_raw_universe) {
  cat(paste0(
    "<div class='callout callout-warning'>",
    "Full, unfiltered GSEA results not found at `", params$all_gsea_csv, "`. ",
    "The tests in this section require the complete tested pathway universe ",
    "(not just the paired-significant subset) to define a proper background ",
    "rate, so they are skipped. Point `params$all_gsea_csv` at your ",
    "`all_GSEA_results.csv` to enable them.",
    "</div>"
  ))
}
```

## Test 1: Enrichment of expected-pharmacology pathways among significant hits

**Question:** among every pathway actually tested in the paired
`Treated vs Untreated` comparison, are pathways related to JAK-STAT/interferon
signaling, downstream NF-\u03baB/TNF crosstalk, and immune-activation-linked metabolism
disproportionately more likely to be paired-significant than other pathways?

A pathway is flagged as belonging to this expected-pharmacology group by simple
keyword matching against its name (pattern below — adjust `params$jakstat_pattern`
to broaden or narrow this). This is a coarse label, not a curated canonical gene
list, so treat it as a first-pass test rather than a definitive mechanistic claim.

```{r enrichment-pattern-display}
cat("Pattern used:", params$jakstat_pattern)
```

```{r enrichment-test}
if (has_raw_universe) {

  universe <- raw_tu %>%
    mutate(
      is_jakstat_ifn = str_detect(pathway, regex(params$jakstat_pattern, ignore_case = TRUE)),
      is_significant = !is.na(padj_TU) & padj_TU <= params$fdr_cutoff
    )

  run_fisher <- function(df) {
    tab <- table(
      factor(df$is_jakstat_ifn, levels = c(FALSE, TRUE)),
      factor(df$is_significant, levels = c(FALSE, TRUE))
    )
    ft <- fisher.test(tab)
    tibble(
      n_tested = nrow(df),
      n_jakstat_ifn_tested = sum(df$is_jakstat_ifn),
      n_significant = sum(df$is_significant),
      n_jakstat_ifn_significant = sum(df$is_jakstat_ifn & df$is_significant),
      pct_significant_in_jakstat_ifn = round(
        100 * sum(df$is_jakstat_ifn & df$is_significant) / max(sum(df$is_jakstat_ifn), 1), 2
      ),
      pct_significant_elsewhere = round(
        100 * sum(!df$is_jakstat_ifn & df$is_significant) / max(sum(!df$is_jakstat_ifn), 1), 2
      ),
      odds_ratio = unname(ft$estimate),
      fold_enrichment = pct_significant_in_jakstat_ifn / pmax(pct_significant_elsewhere, 1e-6),
      fisher_p_value = ft$p.value,
      fisher_neg_log10_p = -log10(pmax(ft$p.value, 1e-300))
    )
  }

  overall_enrichment <- run_fisher(universe) %>% mutate(database = "All databases")

  per_database_enrichment <- universe %>%
    group_split(database) %>%
    map_dfr(~ run_fisher(.x) %>% mutate(database = unique(.x$database)))

  enrichment_results <- bind_rows(overall_enrichment, per_database_enrichment) %>%
    select(database, everything()) %>%
    mutate(
      fisher_p_display = if_else(
        fisher_p_value < 1e-300,
        "< 1e-300",
        formatC(fisher_p_value, format = "e", digits = 2)
      ),
      fisher_neg_log10_p = round(fisher_neg_log10_p, 1)
    ) %>%
    select(-fisher_p_value)

  knitr::kable(
    enrichment_results,
    caption = paste0(
      "Fisher's exact test: expected-pharmacology-pattern pathways ",
      "(JAK-STAT/interferon/NF-\u03baB-TNF/immune-linked metabolism) vs. all others, ",
      "significant (FDR \u2264 ", params$fdr_cutoff, ") vs. not. ",
      "P-values below machine precision are floored and shown as '< 1e-300' rather ",
      "than 0."
    ),
    col.names = c(
      "Database", "Pathways tested", "Pattern pathways tested", "Total significant",
      "Pattern significant", "% sig. within pattern", "% sig. elsewhere",
      "Odds ratio", "Fold enrichment", "-log10(p)", "Fisher p-value"
    )
  )
}
```

```{r enrichment-plot, fig.width=9, fig.height=5}
if (has_raw_universe) {
  enrichment_plot_data <- universe %>%
    mutate(group = if_else(is_jakstat_ifn, "Expected-pharmacology pattern", "All other pathways")) %>%
    count(group, is_significant) %>%
    group_by(group) %>%
    mutate(percentage = 100 * n / sum(n)) %>%
    ungroup() %>%
    filter(is_significant)

  ggplot(enrichment_plot_data, aes(x = group, y = percentage, fill = group)) +
    geom_col(width = 0.55, show.legend = FALSE) +
    geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5, size = 3.5) +
    scale_fill_manual(values = c(
      "Expected-pharmacology pattern" = "#009E73",
      "All other pathways" = "grey65"
    )) +
    labs(
      title = "% of tested pathways reaching paired significance",
      subtitle = "Comparing expected-pharmacology-pattern pathways (JAK-STAT/IFN/NF-\u03baB/metabolism) to everything else tested",
      x = NULL, y = "% significant (paired FDR \u2264 cutoff)"
    ) +
    theme_report() +
    theme(legend.position = "none")
}
```

**Interpretation:** a fold enrichment noticeably above 1, with a small Fisher
p-value, indicates expected-pharmacology-pattern pathways are hit at a higher rate
than the rest of the tested universe — consistent with an on-target and downstream
pharmacological effect rather than pathways lighting up at the same background rate
as everything else. Note this test pools rows across cell types, so a pathway tested
in 18 cell types contributes up to 18 rows; treat the overall p-value as indicative
rather than a formally independent-samples test (see Test 2 for a complementary,
per-pathway recurrence check that respects cell-type structure more directly).

## Test 2: Cross-cell-type recurrence via Fisher's method

**Question:** for pathways that turned up as paired-significant in several
independent cell types, is that recurrence itself stronger than chance? Each cell
type's paired test is run on a biologically distinct, largely independent population,
so seeing the same pathway significant across many of them is a separate piece of
evidence from any single cell type's FDR.

For the top `r params$top_n_recurring` most-recurring paired-significant pathways
(ranked by number of cell types in which they reached significance), this combines
that pathway's p-value **from every cell type it was tested in** (whether or not it
was individually significant there) using Fisher's method, then applies a
Benjamini-Hochberg correction across the resulting list.

```{r recurrence-test}
if (has_raw_universe) {

  if (!has_raw_pvals) {
    cat(paste0(
      "Raw p-values (`pval`) were not found in `all_GSEA_results.csv` — only ",
      "adjusted p-values (`padj`) are available. Falling back to `padj` in place ",
      "of `pval` for Fisher's method below. This is more conservative/approximate ",
      "than combining true p-values and should be treated accordingly."
    ))
  }

  recurring_pathways <- mapped %>%
    distinct(celltype, database, pathway, pathway_label) %>%
    count(database, pathway, pathway_label, name = "n_celltypes_significant") %>%
    arrange(desc(n_celltypes_significant)) %>%
    slice_head(n = params$top_n_recurring)

  fisher_combine <- function(p) {
    p <- pmax(pmin(p, 1 - 1e-12), 1e-300)  # guard against exact 0/1
    k <- length(p)
    stat <- -2 * sum(log(p))
    pchisq(stat, df = 2 * k, lower.tail = FALSE)
  }

  recurrence_results <- recurring_pathways %>%
    rowwise() %>%
    mutate(
      celltype_pvals = list(
        raw_tu %>%
          filter(database == .env$database, pathway == .env$pathway) %>%
          mutate(p_use = if (has_raw_pvals) coalesce(pval_TU, padj_TU) else padj_TU) %>%
          pull(p_use)
      ),
      n_celltypes_tested = length(celltype_pvals),
      combined_p = fisher_combine(celltype_pvals)
    ) %>%
    ungroup() %>%
    select(-celltype_pvals) %>%
    mutate(
      combined_fdr = p.adjust(combined_p, method = "BH"),
      neg_log10_p = -log10(pmax(combined_p, 1e-300)),
      combined_p_display = if_else(
        combined_p < 1e-300, "< 1e-300", formatC(combined_p, format = "e", digits = 2)
      ),
      combined_fdr_display = if_else(
        combined_fdr < 1e-300, "< 1e-300", formatC(combined_fdr, format = "e", digits = 2)
      )
    ) %>%
    arrange(combined_p)

  recurrence_results %>%
    transmute(
      Database = database,
      Pathway = pathway_label,
      `Significant in N cell types` = n_celltypes_significant,
      `Tested in N cell types` = n_celltypes_tested,
      `-log10(p)` = round(neg_log10_p, 1),
      `Fisher combined p-value` = combined_p_display,
      `BH-adjusted q-value` = combined_fdr_display
    ) %>%
    knitr::kable(
      caption = paste0(
        "Cross-cell-type recurrence: Fisher's method combining per-cell-type p-values. ",
        "P-values below machine precision are floored and shown as '< 1e-300' rather than 0."
      )
    )
}
```

**Interpretation:** a very small combined p-value for a pathway that recurs across
many cell types means that pattern is extremely unlikely to arise if each cell type's
result were an independent random draw — i.e. the recurrence itself, not just any
single test, is statistically meaningful. This corroborates Test 1 from a different
angle: Test 1 asks whether expected-pharmacology pathways as a category are
overrepresented among hits; Test 2 asks whether specific pathways' repeated
appearance across independent cell populations could plausibly be coincidence.

::: {.callout-note}
## What these tests do and don't establish

Both tests support the claim that this cluster of results is **statistically unlikely
to reflect random scatter** across the tested pathway universe. Neither test proves
the biological mechanism — that would require leading-edge gene overlap with canonical
JAK-STAT target lists, an orthogonal assay (e.g. phospho-STAT flow cytometry), or a
permutation test that reruns the full limma+fgsea pipeline under label-shuffled null
data. Test 1's p-value should also be read cautiously since it pools
celltype-database-pathway rows rather than treating cell types as fully independent
samples.
:::

# External Validation Comparison

An independently published analysis of JAK inhibitor treatment (PBMC and duodenum,
using healthy PBMC as the control) reported the same five pathway categories moving
in the same direction — elevated in disease, suppressed by treatment: STAT/IFN
signaling, antigen presentation, cytokine/inflammatory signaling,
chemotaxis/migration, and mitochondrial/OXPHOS. The heatmap below reproduces that
figure's layout using this dataset, to allow direct visual comparison.

Within each category, only pathways where the paired treatment effect genuinely
**opposes** the disease-associated direction (`sign(NES_UH) != sign(NES_TU)`) are
shown — the same "improving" criterion used in @sec-primary — ranked by paired FDR.
This matches the external figure's selection of pathways that improve with
treatment, rather than showing top hits regardless of direction.

::: {.callout-note}
## This section is curated, not an unbiased ranking

Unlike the other sections in this report, the pathways shown here were **selected
because they match five categories chosen for comparison to the external figure**,
not purely by statistical rank. Within each category, the pathways shown are the
top hits by paired FDR that also matched the category's keyword pattern — but the
categories themselves were picked to mirror the external reference, so this section
should be read as "does our data reproduce that published pattern," not as a
neutral summary of the strongest hits overall (see the Strongest Hits and Explore
sections above for that).
:::

::: {.callout-important}
## Why only one featured cell type

An initial version of this section also featured Granulocyte as a secondary
myeloid-lineage comparator. It was dropped after review: cell-count diagnostics
showed one of four paired subjects contributed fewer than 20 granulocyte cells
across both treatment conditions combined, meaning that subject's contribution to
the paired comparison was effectively noise. A follow-up gene-set overlap check
also found that the curated pathway list for that population collapsed to roughly
half as many *effectively independent* signals as its nominal count, since several
of the "agreeing" pathways were near-synonyms sharing most of their genes. Rather
than present a comparison with that many caveats attached, this section relies
solely on **Macrophage** — a well-powered population across all subjects, and a
closer lineage match to the external reference's myeloid dendritic cells (both are
professional antigen-presenting myeloid cells).
:::

```{r curated-category-setup}
curated_categories <- tribble(
  ~category, ~pattern,
  "STAT / IFN signaling",        "INTERFERON|IFN_|IFNG|IFNA|JAK|STAT[0-9]",
  "Antigen presentation",        "ANTIGEN_PRESENT|MHC_CLASS|HLA_|CROSS_PRESENT|PEPTIDE_LOADING|TAP|ANTIGEN_PROCESSING",
  "Cytokine / inflammatory",     "CYTOKINE|INTERLEUKIN|IL[0-9]|TNF|NFKB|NF_KB|INFLAMMATORY",
  "Chemotaxis / migration",      "CHEMOTAXIS|MIGRATION|CELL_MOTILITY|LEUKOCYTE_MIGRATION|ENDOTHELIAL",
  "Mitochondrial / OXPHOS",      "OXIDATIVE_PHOSPHORYLATION|MITOCHONDRI|RESPIRATORY_ELECTRON|AEROBIC_RESPIRATION|ATP_SYNTHESIS|ATP_METABOLIC"
)

curated_categories <- curated_categories %>%
  mutate(category = factor(category, levels = category))

featured_celltype <- params$featured_celltype

celltype_data <- mapped %>% filter(celltype == featured_celltype)

curated_selection <- map_dfr(seq_len(nrow(curated_categories)), function(i) {
  celltype_data %>%
    filter(
      str_detect(pathway, regex(curated_categories$pattern[i], ignore_case = TRUE)),
      # This is the "opposes the disease-associated direction" filter described in
      # the prose above, made explicit: disease-elevated (Untreated vs Healthy,
      # NES sign only, no FDR filter -- unpaired, 2-donor reference) that the
      # paired, significant treatment effect pulls back down.
      NES_UH > 0,                        # disease-high (sign only, no FDR filter)
      NES_TU < 0,                        # treatment pushes it down
      padj_TU <= params$fdr_cutoff       # AND that treatment effect is significant
    ) %>%
    arrange(padj_TU) %>%
    slice_head(n = params$n_per_curated_category) %>%
    mutate(
      category = curated_categories$category[i],
      # Disambiguate pathways with identical labels across databases (e.g. GO_BP
      # and Hallmark can both have a pathway called "Oxidative Phosphorylation")
      # so they don't collide onto the same heatmap row.
      pathway_label_unique = if_else(
        pathway_label %in% pathway_label[duplicated(pathway_label)],
        paste0(pathway_label, " (", database, ")"),
        pathway_label
      )
    )
})
```

```{r curated-heatmap-guard, results='asis'}
if (nrow(curated_selection) == 0) {
  cat(paste0(
    "<div class='callout callout-warning'>No paired-significant pathways in `",
    featured_celltype, "` matched any of the curated category patterns. Try a ",
    "different `params$featured_celltype`, or broaden the patterns in ",
    "`curated_categories`.</div>"
  ))
}
```

```{r curated-heatmap, fig.width=11, fig.height=max(6, 0.28 * nrow(curated_selection) + 3)}
if (nrow(curated_selection) > 0) {
  # Flat, single-panel style (same layout as the "Strongest hits" heatmap
  # above): no facet strips/panels, category folded into the row label as
  # "Category | Pathway", and all three columns rendered identically (no
  # fading/border distinction between the paired and unpaired comparisons --
  # see the callout immediately below for how to read that difference here).
  curated_flat <- curated_selection %>%
    arrange(category, padj_TU) %>%
    mutate(
      pathway_category = paste0(category, " | ", pathway_label_unique),
      pathway_category = fct_rev(fct_inorder(pathway_category))
    )

  curated_long <- curated_flat %>%
    select(pathway_category, NES_UH, NES_TH, NES_TU) %>%
    pivot_longer(
      cols = c(NES_UH, NES_TH, NES_TU),
      names_to = "comparison",
      values_to = "NES"
    ) %>%
    mutate(
      comparison = factor(
        comparison,
        levels = c("NES_UH", "NES_TH", "NES_TU"),
        labels = c(
          paste0(featured_celltype, "\nUntreated vs Healthy"),
          paste0(featured_celltype, "\nTreated vs Healthy"),
          paste0(featured_celltype, "\nTreated vs Untreated")
        )
      )
    )

  ggplot(curated_long, aes(x = comparison, y = pathway_category, fill = NES)) +
    geom_tile(color = "white", linewidth = 0.4) +
    geom_text(
      aes(label = ifelse(is.na(NES), "", sprintf("%.2f", NES))),
      size = 2.9, color = "grey10"
    ) +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0, na.value = "grey90"
    ) +
    labs(
      title = paste0("Example pathways altered in disease in ", featured_celltype, " that improve with treatment"),
      subtitle = paste0(
        "Untreated vs Healthy: unpaired, n=2 healthy donors (directional context only) | ",
        "Treated vs Untreated: paired, matched subjects (the statistically tested result)"
      ),
      x = NULL, y = NULL, fill = "NES"
    ) +
    theme_report() +
    theme(
      axis.text.y = element_text(size = 7.5),
      axis.text.x = element_text(size = 8, lineheight = 0.9)
    )
}
```

::: {.callout-important}
## How to read the three columns

The three columns here are **not equally reliable**, even though they're drawn with
equal visual weight. `Treated vs Untreated` (right) compares matched subjects within
the same dataset and lab workflow — this is the statistically robust result that
determined which pathways appear in this section at all (paired-significant, FDR ≤
`r params$fdr_cutoff`). `Untreated vs Healthy` and `Treated vs Healthy` (left, middle)
compare against only 2 external healthy donors from a different study, unpaired — a
much weaker comparison included purely to suggest a direction, not to make a
confident claim on its own. Inconsistent or near-zero values in those two columns do
not undermine the paired result on the right; they simply reflect the limits of a
2-donor unpaired reference.
:::

**Featured cell type:** `r featured_celltype`. Set `params$featured_celltype` to
feature a different one if a more granular myeloid/dendritic-cell annotation becomes
available.

# Disease-Associated Pathways and Recovery With Treatment

This section reproduces the two-panel style of an external reference figure, using
this dataset's actual, well-powered starting point:

- **`Treated vs Untreated`** (defines the pathway set to look at): **must be
  significant** (FDR ≤ `r params$fdr_cutoff`). This is the paired, well-powered
  comparison, so it is the primary filter — the same logic as @sec-primary and the
  original pathway-selection script this report is built from.
- **`Untreated vs Healthy`** (checks disease relevance): **NES sign only, no
  significance filter.** This comparison is unpaired and rests on only 2 external
  healthy donors, so it is used here strictly to check direction — was this
  significant treatment effect also pointing back against a disease-associated
  signal — not as a statistically filtered result on its own.

::: {.callout-note}
## Why Treated vs Untreated comes first

The paired `Treated vs Untreated` comparison is the well-powered, reliable one in
this dataset, so it defines which pathways are considered at all. `Untreated vs
Healthy` is then used only to ask a follow-up question about those already-
significant pathways — does the treatment effect line up with a disease-associated
direction — rather than being used to define a separate pathway set of its own.
:::

## Pathways significantly changed by treatment (Treated vs Untreated)

```{r tu-significant-guard, results='asis'}
if (!has_raw_universe) {
  cat(paste0(
    "<div class='callout callout-warning'>Full GSEA results not found at `",
    params$all_gsea_csv, "` — this section is skipped.</div>"
  ))
}
```

```{r tu-significant-plot, fig.width=13, fig.height=max(7, 0.32 * n_distinct(raw_tu$celltype) + 3)}
if (has_raw_universe) {

  tu_significant <- raw_tu %>%
    filter(!is.na(padj_TU), padj_TU <= params$fdr_cutoff) %>%
    mutate(tu_direction = if_else(NES_TU > 0, "Up", "Down"))

  celltype_order <- tu_significant %>%
    count(celltype, name = "n") %>%
    arrange(n) %>%
    pull(celltype)

  tu_dir_counts <- tu_significant %>%
    count(celltype, database, tu_direction, name = "n_pathways") %>%
    mutate(celltype = factor(celltype, levels = celltype_order))

  ggplot(tu_dir_counts, aes(x = celltype, y = n_pathways, fill = tu_direction)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~database, nrow = 1, scales = "free_x") +
    scale_fill_manual(
      values = c(Up = "#D7191C", Down = "#2C7FB8"),
      name = "Direction in\nTreated vs Untreated"
    ) +
    labs(
      title = "Treated vs Untreated identifies significantly changed pathways",
      subtitle = paste0(
        "FDR \u2264 ", params$fdr_cutoff,
        "; paired comparison, matched subjects within the Internal dataset"
      ),
      x = NULL, y = "Number of significant pathways"
    ) +
    theme_report()
}
```

## Are those pathways also disease-associated? (checked against Untreated vs Healthy)

For every pathway significant in `Treated vs Untreated` above, this checks whether
`Untreated vs Healthy`'s NES sign (direction only, no significance filter) points
in the **opposite** direction from the treatment effect — i.e. the treatment is
pushing back against a disease-associated signal, not just moving a pathway for
unrelated reasons.

```{r disease-relevance-join}
if (has_raw_universe) {
  disease_relevance_data <- tu_significant %>%
    left_join(raw_uh, by = c("celltype", "database", "pathway")) %>%
    mutate(
      tested_in_uh = !is.na(NES_UH),
      opposite_direction = tested_in_uh & (sign(NES_UH) != sign(NES_TU)),
      relevance_status = case_when(
        !tested_in_uh ~ "Not tested in Untreated vs Healthy",
        opposite_direction ~ "Opposes disease direction (moves toward healthy)",
        TRUE ~ "Same direction as disease (does not oppose)"
      ),
      relevance_status = factor(relevance_status, levels = c(
        "Opposes disease direction (moves toward healthy)",
        "Same direction as disease (does not oppose)",
        "Not tested in Untreated vs Healthy"
      ))
    )
}
```

```{r disease-relevance-plot, fig.width=13, fig.height=max(7, 0.32 * n_distinct(tu_significant$celltype) + 3)}
if (has_raw_universe) {

  relevance_counts <- disease_relevance_data %>%
    count(celltype, database, relevance_status, name = "n_pathways") %>%
    mutate(celltype = factor(celltype, levels = celltype_order))

  ggplot(relevance_counts, aes(x = celltype, y = n_pathways, fill = relevance_status)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~database, nrow = 1, scales = "free_x") +
    scale_fill_manual(
      values = c(
        "Opposes disease direction (moves toward healthy)" = "#1B9E77",
        "Same direction as disease (does not oppose)" = "#D7191C",
        "Not tested in Untreated vs Healthy" = "grey70"
      ),
      name = NULL
    ) +
    labs(
      title = "Treatment-significant pathways that oppose the disease-associated direction",
      subtitle = "Among pathways significant in Treated vs Untreated, direction relative to Untreated vs Healthy (NES sign only)",
      x = NULL, y = "Number of pathways"
    ) +
    theme_report()
}
```

```{r disease-relevance-summary-table}
if (has_raw_universe) {
  relevance_summary_table <- disease_relevance_data %>%
    count(celltype, relevance_status, name = "n_pathways") %>%
    pivot_wider(names_from = relevance_status, values_from = n_pathways, values_fill = 0) %>%
    mutate(
      total_tu_significant = rowSums(across(where(is.numeric))),
      pct_opposing_disease = round(
        100 * `Opposes disease direction (moves toward healthy)` / pmax(total_tu_significant, 1), 1
      )
    ) %>%
    arrange(desc(pct_opposing_disease))

  datatable(
    relevance_summary_table,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 15, scrollX = TRUE, dom = "ftip")
  ) %>%
    formatStyle(
      "pct_opposing_disease",
      background = styleColorBar(c(0, 100), "#c6e8c6"),
      backgroundPosition = "center"
    )
}
```

# Explore all significant pathways

```{r explore-table}
explore_table <- mapped %>%
  transmute(
    `Cell type` = celltype,
    Database = database,
    Pathway = pathway_label,
    Interpretation = directional_response,
    `NES (Untreated vs Healthy)` = round(NES_UH, 2),
    `NES (Treated vs Healthy)` = round(NES_TH, 2),
    `NES (Treated vs Untreated)` = round(NES_TU, 2),
    `Paired FDR` = signif(padj_TU, 3)
  ) %>%
  arrange(`Paired FDR`)

datatable(
  explore_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
) %>%
  formatStyle(
    "Interpretation",
    backgroundColor = styleEqual(
      names(direction_colors),
      scales::alpha(unname(direction_colors), 0.25)
    )
  ) %>%
  formatStyle(
    c("NES (Untreated vs Healthy)", "NES (Treated vs Healthy)", "NES (Treated vs Untreated)"),
    background = styleColorBar(range(explore_table[, 5:7], na.rm = TRUE), "#eee"),
    backgroundPosition = "center"
  )
```

# Focused biological programs

A curated set of pathway keyword groups, showing the single strongest
paired-significant pathway per cell type × database × program.

```{r focused-programs}
program_definitions <- tribble(
  ~program, ~pattern,
  "IFN / IFNG signaling",     "INTERFERON|IFNG|IFN_GAMMA|TYPE_II_INTERFERON",
  "Antigen presentation",     "ANTIGEN_PRESENT|MHC_CLASS|HLA_|CROSS_PRESENT|PEPTIDE_LOADING|TAP",
  "Cytotoxicity",             "CYTOTOX|GRANZYME|PERFORIN|NATURAL_KILLER|NK_CELL_MEDIATED|T_CELL_MEDIATED",
  "Stem / regeneration",      "STEM|WNT|NOTCH|REGENERATION|TRANSIT_AMPLIFY|EPITHELIAL_REPAIR|LGR5"
)

focused_programs <- map_dfr(seq_len(nrow(program_definitions)), function(i) {
  mapped %>%
    filter(str_detect(pathway, regex(program_definitions$pattern[i], ignore_case = TRUE))) %>%
    mutate(biological_program = program_definitions$program[i])
})

if (nrow(focused_programs) > 0) {
  focused_programs <- focused_programs %>%
    group_by(celltype, database, biological_program) %>%
    slice_min(padj_TU, n = 1, with_ties = FALSE) %>%
    ungroup()

  p_bubble <- ggplot(focused_programs, aes(x = biological_program, y = celltype)) +
    geom_point(
      aes(size = paired_fdr_strength, color = directional_response, shape = database),
      alpha = 0.85
    ) +
    scale_color_manual(values = direction_colors, drop = FALSE, name = "Interpretation") +
    labs(
      title = "Focused paired-significant pathway responses",
      subtitle = "One strongest pathway per cell type, database, and program",
      x = "Biological program", y = NULL, size = NULL, shape = "Database"
    ) +
    theme_report() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))

  ggplotly(p_bubble) %>% layout(legend = list(orientation = "h", y = -0.2))
} else {
  cat("No pathways matched the defined biological programs.")
}
```

# Notes

- Pathway inclusion and all statistical inference are based **exclusively** on the
  paired `Treated vs Untreated` comparison at FDR ≤ `r params$fdr_cutoff`.
- The Yu2021 healthy cohort (`Untreated vs Healthy`, `Treated vs Healthy`) is an
  **external, unpaired reference** used only to give directional context — it does not
  contribute to significance testing here.
- @sec-primary is the headline, most defensible figure for "did treatment reverse
  the disease-associated direction" — it relies only on the paired, well-powered
  `Treated vs Untreated` comparison and the sign of the disease direction, never on
  the noisy `Treated vs Healthy` endpoint.
- @sec-trajectory shows approximately where pathways landed relative to healthy, but
  layers a robust metric (color) on top of a much noisier one (arrow position/length,
  based on the 2-donor `Treated vs Healthy` comparison) — read the color, treat arrow
  length as context only. The quadrant view is useful for spotting pathways where a
  strong baseline disease signal and a strong treatment effect disagree in direction.
- The enrichment and recurrence tests test whether the JAK-STAT/interferon-heavy
  pattern in the results is more concentrated than chance would produce — they support
  but do not on their own prove a specific mechanism of action.
- The External Validation Comparison section is intentionally curated to mirror a
  specific published figure, and features only Macrophage after Granulocyte was
  excluded on cell-count and pathway-redundancy grounds (see the callout in that
  section for details). Its heatmap now uses the flat, single-panel layout (no
  facet strips, no fading/border) rather than category-faceted panels; see the
  "How to read the three columns" callout for the reliability distinction the
  fading previously conveyed.
