---
title: "Interferon Pathway UpSet Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    code-fold: true
    embed-resources: true
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(ggbeeswarm)
library(pheatmap)
library(viridis)
library(scales)
library(DT)

# ─────────────────────────────────────────────────────────────────────────────
# USER SETTINGS
# ─────────────────────────────────────────────────────────────────────────────
GSEA_PATH  <- "./interferon_leading_edge_genes.xlsx"
MIN_CD45NEG_FREQ <- 0.50
MAX_CD45POS_FREQ <- 0.20

# ─────────────────────────────────────────────────────────────────────────────
# CD45 compartment definitions
# ─────────────────────────────────────────────────────────────────────────────
cd45_pos <- c("B_plasma", "Conventional_CD4", "Conventional_CD8",
              "Cycling_T/NK", "DC", "Granulocyte", "Macrophage",
              "Mature_B", "Monocyte", "NK", "Treg", "Unconventional_T/ILC")
cd45_neg <- c("Absorptive", "Cycling_epithelia", "Enteroendocrine",
              "Fibroblast", "Microfold", "Secretory", "Stem",
              "Transit_amplifying", "Vascular_endothelia",
              "Lymphatic_endothelia", "Myofibroblast", "Pericyte")

RED  <- "#E63946"
BLUE <- "#457B9D"
PURP <- "#6A0572"
ct_palette <- c(
  "#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#F4A261",
  "#264653", "#A8DADC", "#F77F00", "#8338EC", "#3A86FF",
  "#06D6A0", "#FFB703", "#FB8500", "#8ECAE6", "#219EBC",
  "#023047", "#FFAFCC", "#BDE0FE", "#CDB4DB", "#A2D2FF",
  "#CAFFBF", "#FDFFB6", "#FFD6A5", "#FFADAD")

# ─────────────────────────────────────────────────────────────────────────────
# Load GSEA data
# ─────────────────────────────────────────────────────────────────────────────
ip <- read_excel(GSEA_PATH)
ip_filt <- ip %>%
  filter(p.adjust < 0.05, !is.na(leading_edge_symbols)) %>%
  mutate(
    genes = str_split(leading_edge_symbols, "/"),
    cd45  = case_when(
      cluster %in% cd45_pos ~ "CD45+",
      cluster %in% cd45_neg ~ "CD45-",
      TRUE ~ "Unknown"
    )
  ) %>%
  filter(cd45 != "Unknown")

pathways <- unique(ip_filt$Description)

# ─────────────────────────────────────────────────────────────────────────────
# Pooled gene summary (union across all pathways per cluster)
# ─────────────────────────────────────────────────────────────────────────────
pooled_all <- ip_filt %>%
  group_by(cluster, cd45) %>%
  summarise(genes = list(sort(unique(unlist(genes)))), .groups = "drop")

n_cd45pos <- sum(pooled_all$cd45 == "CD45+")
n_cd45neg <- sum(pooled_all$cd45 == "CD45-")

cd45pos_counts <- pooled_all %>%
  filter(cd45 == "CD45+") %>% unnest(genes) %>%
  distinct(cluster, genes) %>% count(genes, name = "n_cd45pos_celltypes")
cd45neg_counts <- pooled_all %>%
  filter(cd45 == "CD45-") %>% unnest(genes) %>%
  distinct(cluster, genes) %>% count(genes, name = "n_cd45neg_celltypes")

pooled_summary <- full_join(cd45pos_counts, cd45neg_counts, by = "genes") %>%
  replace_na(list(n_cd45pos_celltypes = 0, n_cd45neg_celltypes = 0)) %>%
  mutate(
    freq_cd45pos   = n_cd45pos_celltypes / n_cd45pos,
    freq_cd45neg   = n_cd45neg_celltypes / n_cd45neg,
    cd45neg_specific = freq_cd45neg >= MIN_CD45NEG_FREQ & freq_cd45pos <= MAX_CD45POS_FREQ,
    cd45pos_specific = freq_cd45pos >= MIN_CD45NEG_FREQ & freq_cd45neg <= MAX_CD45POS_FREQ,
    label = case_when(
      cd45neg_specific ~ "CD45- specific",
      cd45pos_specific ~ "CD45+ specific",
      freq_cd45neg >= 0.3 & freq_cd45pos >= 0.3 ~ "Shared",
      TRUE ~ "Low frequency"
    )
  )

# ─────────────────────────────────────────────────────────────────────────────
# Helper: gene intersection sets per pathway
# ─────────────────────────────────────────────────────────────────────────────
get_sets <- function(pw) {
  d   <- ip_filt %>% filter(Description == pw)
  cp  <- unique(unlist(d %>% filter(cd45 == "CD45+") %>% pull(genes)))
  cn  <- unique(unlist(d %>% filter(cd45 == "CD45-") %>% pull(genes)))
  list(
    cd45pos_genes = cp, cd45neg_genes = cn,
    shared        = intersect(cp, cn),
    cd45pos_only  = setdiff(cp, cn),
    cd45neg_only  = setdiff(cn, cp),
    all_genes     = union(cp, cn)
  )
}

# ─────────────────────────────────────────────────────────────────────────────
# Helper: UpSet plot builder
# ─────────────────────────────────────────────────────────────────────────────
build_upset <- function(set_list, title, bar_color, set_colors = NULL) {
  if (length(set_list) < 2) return(NULL)
  m_full <- make_comb_mat(set_list, mode = "distinct")
  m      <- m_full[comb_size(m_full) >= 1]
  if (length(comb_size(m)) == 0) return(NULL)
  tss  <- set_size(m_full)[set_name(m)]
  sno  <- set_name(m)
  cols <- if (!is.null(set_colors)) {
    v <- set_colors[sno]; v[is.na(v)] <- "grey60"; v
  } else rep("grey60", length(sno))
  right_ann <- rowAnnotation(
    "Set size" = anno_barplot(tss, baseline = 0,
      axis_param = list(direction = "reverse", gp = gpar(fontsize = 8)),
      gp = gpar(fill = cols, col = NA), border = FALSE, width = unit(3, "cm")),
    "n" = anno_text(as.character(tss), gp = gpar(fontsize = 8),
                    just = "left", width = unit(1.2, "cm")),
    annotation_name_gp = gpar(fontsize = 9), annotation_name_rot = 0,
    show_annotation_name = TRUE)
  UpSet(m,
    set_order = order(tss, decreasing = TRUE),
    comb_order = order(comb_size(m), decreasing = TRUE),
    top_annotation = upset_top_annotation(m,
      add_numbers = TRUE, numbers_rot = 0,
      numbers_gp = gpar(fontsize = 8),
      annotation_name_gp = gpar(fontsize = 9), height = unit(5, "cm")),
    right_annotation = right_ann,
    row_names_gp = gpar(fontsize = 8),
    column_title = title, column_title_gp = gpar(fontsize = 11, fontface = "bold"),
    pt_size = unit(4, "mm"), lwd = 2,
    bg_col = c("#F8F8F8", "#EFEFEF"), comb_col = bar_color)
}

draw_safe <- function(ht) {
  if (!is.null(ht)) draw(ht, padding = unit(c(10, 10, 10, 10), "mm"))
}
```

---

## Overview

Paired 10X scRNA-seq before and after JAK-STAT inhibitor treatment in ACEPD responders. Differential expression was computed using **limma** (post minus pre); GSEA was run on limma t-statistics. All NES values are negative — IFN pathways are uniformly **suppressed after treatment**. Leading-edge genes are the primary drivers of this suppression: genes most strongly downregulated post-treatment.

**Key questions:**

1. Is the magnitude of IFN suppression different between CD45<sup>+</sup> and CD45<sup>-</sup> compartments?
2. Are the same IFN genes switched off in both compartments?
3. Are there genes consistently suppressed across CD45<sup>-</sup> cell types but not CD45<sup>+</sup>?

---

## Summary table

```{r summary_table}
summary_tbl <- ip_filt %>%
  mutate(n_leading_edge_genes = lengths(genes)) %>%
  select(Description, cluster, cd45, NES, p.adjust, n_leading_edge_genes) %>%
  arrange(Description, cd45, cluster)

datatable(
  summary_tbl, rownames = FALSE,
  caption    = "GSEA results — interferon pathways, FDR < 0.05",
  extensions = "Buttons",
  options    = list(pageLength = 20, scrollX = TRUE,
                    dom = "Bfrtip", buttons = c("copy", "csv", "excel"))) %>%
  DT::formatRound(c("NES", "p.adjust"), digits = 4) %>%
  DT::formatStyle("cd45",
    backgroundColor = DT::styleEqual(c("CD45+", "CD45-"), c("#fde8e8", "#e8f0fb"))) %>%
  DT::formatStyle("NES",
    color = DT::styleInterval(0, c("#457B9D", "#E63946")))
```

Each row = one pathway × cell type at FDR < 0.05. All NES values are **negative**, confirming uniform IFN pathway suppression after treatment.

---

## NES Distribution: CD45<sup>+</sup> vs CD45<sup>-</sup>

Each point = one cell type. More negative NES = stronger suppression. Black bar = group mean. Faceted by pathway.

```{r nes_per_pathway, fig.width=12, fig.height=7}
mean_ann <- ip_filt %>%
  group_by(Description, cd45) %>%
  summarise(mean_NES = round(mean(NES), 2), .groups = "drop")

p_nes <- ggplot(ip_filt, aes(x = cd45, y = NES, colour = cd45, fill = cd45)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.5) +
  geom_boxplot(width = 0.4, alpha = 0.15, outlier.shape = NA, linewidth = 0.7) +
  geom_beeswarm(size = 3, alpha = 0.85, cex = 2.8) +
  ggrepel::geom_text_repel(
    aes(label = cluster),
    size = 2.6, max.overlaps = 10,
    segment.size = 0.25, segment.alpha = 0.4,
    show.legend = FALSE
  ) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.35,
               colour = "black", linewidth = 0.6, fatten = 2) +
  facet_wrap(~Description, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = c("CD45+" = RED, "CD45-" = BLUE)) +
  scale_fill_manual(values   = c("CD45+" = RED, "CD45-" = BLUE)) +
  labs(
    title    = "IFN Pathway Suppression: CD45- vs CD45+ per pathway (post - pre JAK-STAT inhibition)",
    subtitle = "Each point = one cell type. Black bar = group mean. More negative NES = stronger suppression.",
    x = NULL, y = "Normalised Enrichment Score (NES)",
    colour = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    panel.spacing    = unit(1.2, "lines")
  )
print(p_nes)
```

```{r nes_summary_table}
nes_summary <- ip_filt %>%
  group_by(Pathway = Description, Compartment = cd45) %>%
  summarise(
    `n cell types` = n(),
    `Mean NES`     = round(mean(NES), 3),
    `Min NES`      = round(min(NES), 3),
    `Max NES`      = round(max(NES), 3),
    .groups = "drop"
  ) %>%
  arrange(Pathway, Compartment)

datatable(
  nes_summary, rownames = FALSE,
  caption = "Mean NES per pathway and compartment.",
  options = list(pageLength = 20, scrollX = TRUE)) %>%
  DT::formatStyle("Mean NES",
    background = DT::styleColorBar(range(nes_summary$`Mean NES`), "#d4e8f7"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  ) %>%
  DT::formatStyle("Compartment",
    backgroundColor = DT::styleEqual(
      c("CD45+", "CD45-"), c("#fde8e8", "#e8f0fb")
    )
  )
```

More negative mean NES in CD45<sup>-</sup> rows indicates stronger suppression in the epithelial/stromal compartment for that pathway; compare direction across all four pathways for a consistent signal.

---

## Leading-Edge Gene t-Statistics Across Cell Types

The dot plot below focuses on **CD45<sup>-</sup> specific genes** (≥50% of CD45<sup>-</sup> cell types). Each row = gene, each column = cell type. Dot size = |t-statistic|; colour = direction. Blue dots in CD45<sup>-</sup> columns = consistent downregulation (expected); red dots = upregulation (unexpected).

```{r setup_limma}
LIMMA_PATH <- "./PanGI_L2__ALL_clusters_limma.csv"
limma_raw <- read.csv(LIMMA_PATH)

limma_raw <- limma_raw %>%
  mutate(cd45 = case_when(
    cluster %in% cd45_pos ~ "CD45+",
    cluster %in% cd45_neg ~ "CD45-",
    TRUE ~ "Unknown"
  ))

le_genes_long <- ip_filt %>%
  select(cluster, cd45, Description) %>%
  left_join(
    ip_filt %>% select(cluster, Description, genes) %>% unnest(genes),
    by = c("cluster", "Description")
  ) %>%
  distinct(cluster, cd45, genes) %>%
  rename(gene = genes)

le_tstat <- le_genes_long %>%
  left_join(limma_raw %>% select(gene, cluster, t, logFC, adj.P.Val),
            by = c("gene", "cluster")) %>%
  filter(!is.na(t))

le_tstat <- le_tstat %>%
  left_join(pooled_summary %>% select(gene = genes, label, freq_cd45neg, freq_cd45pos),
            by = "gene")
```

```{r tstat_dotplot, fig.width=13, fig.height=9, out.width="100%"}
cd45neg_specific_genes <- pooled_summary %>%
  filter(cd45neg_specific) %>%
  arrange(desc(freq_cd45neg)) %>%
  pull(genes)

dot_data <- le_tstat %>%
  filter(gene %in% cd45neg_specific_genes) %>%
  mutate(
    gene      = factor(gene, levels = rev(cd45neg_specific_genes)),
    direction = ifelse(t < 0, "Down (expected)", "Up (unexpected)"),
    abs_t     = abs(t)
  )

all_clusters <- unique(dot_data$cluster)
cluster_order <- c(
  intersect(cd45_neg, all_clusters),
  intersect(cd45_pos, all_clusters))

dot_data <- dot_data %>%
  mutate(
    cluster = factor(cluster, levels = cluster_order),
    cd45    = factor(cd45, levels = c("CD45-", "CD45+"))
  )

ggplot(dot_data, aes(x = cluster, y = gene,
                     colour = direction, size = abs_t)) +
  geom_point(alpha = 0.85) +
  geom_vline(
    xintercept = length(intersect(cd45_neg, all_clusters)) + 0.5,
    linetype = "dashed", colour = "grey40", linewidth = 0.7
  ) +
  annotate("text",
    x = length(intersect(cd45_neg, all_clusters)) / 2,
    y = length(cd45neg_specific_genes) + 0.7,
    label = "CD45- cell types", colour = BLUE,
    size = 3.5, fontface = "bold"
  ) +
  annotate("text",
    x = length(intersect(cd45_neg, all_clusters)) +
        length(intersect(cd45_pos, all_clusters)) / 2,
    y = length(cd45neg_specific_genes) + 0.7,
    label = "CD45+ cell types", colour = RED,
    size = 3.5, fontface = "bold"
  ) +
  scale_colour_manual(values = c(
    "Down (expected)"   = BLUE,
    "Up (unexpected)"   = RED
  )) +
  scale_size_continuous(range = c(1.5, 8), name = "|t-statistic|") +
  labs(
    title    = "t-statistic consistency: CD45- specific leading-edge genes",
    subtitle = "Blue = downregulated post-treatment. Red = upregulated. Size = magnitude. Dashed line separates compartments.",
    x = NULL, y = NULL, colour = "Direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 12),
    plot.subtitle   = element_text(size = 9, colour = "grey40"),
    axis.text.x     = element_text(angle = 40, hjust = 1, size = 9),
    axis.text.y     = element_text(size = 9),
    legend.position = "right",
    panel.grid      = element_line(colour = "grey92")
  )
```

---

## Gene Specificity Across All Pathways (Pooled)

Leading-edge genes are pooled (union) across all IFN pathways per cell type. Axes show the fraction of CD45<sup>+</sup> (x) and CD45<sup>-</sup> (y) cell types in which each gene appears. Dashed lines mark specificity thresholds (≥50% in one compartment, ≤20% in the other).

- **Top-right** — canonical IFN genes suppressed in both compartments (*STAT1*, *IRF1*, *MX1*, *HLA-A/B/C*)
- **Top-left** — CD45<sup>-</sup> specific: epithelial/stromal suppression (*CIITA*, *NUP210/98/153*, *GCH1*, *HSPA1A/B*, *TNF*)
- **Bottom-right** — CD45<sup>+</sup> specific: immune-restricted suppression

```{r scatter_plot, fig.width=10, fig.height=8}
label_genes <- pooled_summary %>%
  filter(freq_cd45neg > 0.4 | freq_cd45pos > 0.4 | cd45neg_specific | cd45pos_specific)

ggplot(pooled_summary, aes(x = freq_cd45pos, y = freq_cd45neg, colour = label)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_text_repel(data = label_genes, aes(label = genes),
                  size = 3, max.overlaps = 35, segment.size = 0.3,
                  segment.alpha = 0.5) +
  geom_hline(yintercept = MIN_CD45NEG_FREQ, linetype = "dashed",
             colour = BLUE, linewidth = 0.7) +
  geom_vline(xintercept = MAX_CD45POS_FREQ, linetype = "dashed",
             colour = RED, linewidth = 0.7) +
  annotate("rect", xmin = 0, xmax = MAX_CD45POS_FREQ,
           ymin = MIN_CD45NEG_FREQ, ymax = 1,
           fill = BLUE, alpha = 0.06) +
  annotate("rect", xmin = MIN_CD45NEG_FREQ, xmax = 1,
           ymin = 0, ymax = MAX_CD45POS_FREQ,
           fill = RED, alpha = 0.06) +
  annotate("text", x = 0.01, y = 0.98,
           label = "CD45- specific zone", hjust = 0,
           colour = BLUE, size = 3.5, fontface = "italic") +
  annotate("text", x = 0.99, y = 0.01,
           label = "CD45+ specific zone", hjust = 1,
           colour = RED, size = 3.5, fontface = "italic") +
  scale_colour_manual(values = c(
    "CD45- specific" = BLUE,
    "CD45+ specific" = RED,
    "Shared"         = PURP,
    "Low frequency"  = "grey70"
  )) +
  scale_x_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    title    = "IFN Leading-Edge Gene Specificity: CD45- vs CD45+ (All Pathways Pooled)",
    subtitle = paste0(
      "CD45- specific: >=50% of CD45- cell types & <=20% of CD45+ cell types  |  ",
      "n(CD45+) = ", n_cd45pos, " cell types, n(CD45-) = ", n_cd45neg, " cell types"
    ),
    x      = "Frequency in CD45+ cell types",
    y      = "Frequency in CD45- cell types",
    colour = "Gene category"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

The shared zone (purple) is populated as expected with canonical IFN targets, serving as a positive control. The CD45<sup>-</sup> specific zone contains a biologically coherent cluster: nucleoporins (*NUP210*, *NUP98*, *NUP153*, *NUP160*) implicated in nuclear STAT transport, *CIITA* (epithelial MHC-II), *HSPA1A/B* (stress response), *GCH1* (tetrahydrobiopterin/NO pathway), and *TNF*. Note that n(CD45<sup>-</sup>) = 6 cell types means 50% = 3/6 cells — genes at exactly this threshold (*PML*, *EGR1*, *VIM*) should be treated as suggestive. The four CD45<sup>+</sup> specific genes include *GAPDH* and *TUBA4A*, likely technical artefacts.

### CD45<sup>-</sup> specific genes

```{r cd45neg_specific_table}
cd45neg_tbl <- pooled_summary %>%
  filter(cd45neg_specific) %>%
  arrange(desc(freq_cd45neg), freq_cd45pos) %>%
  mutate(
    `CD45- freq` = sprintf("%.0f%%", freq_cd45neg * 100),
    `CD45+ freq` = sprintf("%.0f%%", freq_cd45pos * 100)
  ) %>%
  select(
    Gene = genes,
    `CD45- cell types` = n_cd45neg_celltypes,
    `CD45+ cell types` = n_cd45pos_celltypes,
    `CD45- freq`, `CD45+ freq`
  )

if (nrow(cd45neg_tbl) > 0) {
  datatable(cd45neg_tbl, rownames = FALSE,
            caption = paste0("Genes enriched in ≥50% of CD45- and ≤20% of CD45+ cell types (n = ", nrow(cd45neg_tbl), ")"),
            options = list(pageLength = 15, scrollX = TRUE)) %>%
    DT::formatStyle("CD45- freq", color = BLUE, fontWeight = "bold") %>%
    DT::formatStyle("CD45+ freq", color = "grey50")
} else {
  cat("No genes met the CD45- specific criteria.")
}
```

### CD45<sup>+</sup> specific genes

```{r cd45pos_specific_table}
cd45pos_tbl <- pooled_summary %>%
  filter(cd45pos_specific) %>%
  arrange(desc(freq_cd45pos), freq_cd45neg) %>%
  mutate(
    `CD45- freq` = sprintf("%.0f%%", freq_cd45neg * 100),
    `CD45+ freq` = sprintf("%.0f%%", freq_cd45pos * 100)
  ) %>%
  select(
    Gene = genes,
    `CD45- cell types` = n_cd45neg_celltypes,
    `CD45+ cell types` = n_cd45pos_celltypes,
    `CD45- freq`, `CD45+ freq`
  )

if (nrow(cd45pos_tbl) > 0) {
  datatable(cd45pos_tbl, rownames = FALSE,
            caption = paste0("Genes enriched in ≥50% of CD45+ and ≤20% of CD45- cell types (n = ", nrow(cd45pos_tbl), ")"),
            options = list(pageLength = 15, scrollX = TRUE)) %>%
    DT::formatStyle("CD45+ freq", color = RED, fontWeight = "bold") %>%
    DT::formatStyle("CD45- freq", color = "grey50")
} else {
  cat("No genes met the CD45+ specific criteria.")
}
```

---

## Statistical Testing: CD45<sup>-</sup> Specificity

Three complementary tests assess whether leading-edge IFN genes are significantly enriched in CD45<sup>-</sup> vs CD45<sup>+</sup> cell types.

**Test 1 — Fisher's exact test (per gene):** One-sided test for enrichment in CD45<sup>-</sup>. BH FDR correction applied over CD45<sup>-</sup> specific genes only.

**Test 2 — Hypergeometric test (gene-set level):** Tests whether high-frequency genes (≥50% of CD45<sup>-</sup> cell types) are over-represented among CD45<sup>-</sup> specific genes vs the CD45<sup>-</sup> universe.

**Test 3 — Permutation test:** CD45<sup>+</sup>/CD45<sup>-</sup> labels are shuffled 10,000 times across all cell types; the observed frequency difference is compared to the null distribution.

```{r cd45neg_stats, fig.width=9, fig.height=7}
# =============================================================================
# Statistical testing: CD45-neg specific leading-edge IFN genes
# Tests:
#   1. Fisher's exact test per gene (with BH correction)
#   2. Hypergeometric test (gene set level)
#   3. Permutation test per gene
# =============================================================================

N_PERMUTATIONS  <- 10000
SHARED_FREQ_MIN <- 0.30
OUT_STATS_CSV   <- "cd45neg_specificity_stats.csv"

# ─────────────────────────────────────────────────────────────────────────────
# Pooled gene counts per compartment (uses ip_filt from setup chunk)
# ─────────────────────────────────────────────────────────────────────────────
gene_ct_long <- pooled_all %>%
  unnest(genes) %>%
  rename(gene = genes) %>%
  distinct(gene, cluster, cd45)

cd45pos_counts_s <- gene_ct_long %>%
  filter(cd45 == "CD45+") %>%
  count(gene, name = "n_pos")

cd45neg_counts_s <- gene_ct_long %>%
  filter(cd45 == "CD45-") %>%
  count(gene, name = "n_neg")

pooled_summary_s <- full_join(cd45pos_counts_s, cd45neg_counts_s, by = "gene") %>%
  replace_na(list(n_pos = 0, n_neg = 0)) %>%
  mutate(
    freq_pos         = n_pos / n_cd45pos,
    freq_neg         = n_neg / n_cd45neg,
    cd45neg_specific = freq_neg >= MIN_CD45NEG_FREQ & freq_pos <= MAX_CD45POS_FREQ,
    shared           = freq_neg >= SHARED_FREQ_MIN  & freq_pos >= SHARED_FREQ_MIN,
    label = case_when(
      cd45neg_specific ~ "CD45-neg specific",
      freq_pos >= MIN_CD45NEG_FREQ & freq_neg <= MAX_CD45POS_FREQ ~ "CD45-pos specific",
      shared           ~ "Shared",
      TRUE             ~ "Low frequency"
    )
  )

# ─────────────────────────────────────────────────────────────────────────────
# TEST 1: Fisher's exact test per gene
# ─────────────────────────────────────────────────────────────────────────────
fisher_results <- pooled_summary_s %>%
  rowwise() %>%
  mutate(
    fisher_p = fisher.test(
      matrix(c(n_neg, n_pos, n_cd45neg - n_neg, n_cd45pos - n_pos), nrow = 2),
      alternative = "greater"
    )$p.value,
    odds_ratio = fisher.test(
      matrix(c(n_neg, n_pos, n_cd45neg - n_neg, n_cd45pos - n_pos), nrow = 2)
    )$estimate
  ) %>%
  ungroup() %>%
  mutate(fisher_fdr = NA_real_, fisher_sig = FALSE)

fisher_results$fisher_fdr[fisher_results$cd45neg_specific] <-
  p.adjust(fisher_results$fisher_p[fisher_results$cd45neg_specific], method = "BH")
fisher_results$fisher_sig <- !is.na(fisher_results$fisher_fdr) &
  fisher_results$fisher_fdr < 0.05

n_tested <- sum(fisher_results$cd45neg_specific)

cat("=== TEST 1: Fisher's exact test ===\n")
cat("CD45-neg specific genes:", n_tested, "\n")
cat("FDR significant (FDR < 0.05):",
    sum(fisher_results$fisher_sig, na.rm = TRUE), "\n\n")

fisher_results %>%
  filter(cd45neg_specific) %>%
  select(gene, n_neg, n_pos, freq_neg, freq_pos, odds_ratio, fisher_p, fisher_fdr, fisher_sig) %>%
  arrange(fisher_fdr) %>%
  mutate(across(c(freq_neg, freq_pos), ~round(.x, 2)),
         across(c(odds_ratio, fisher_p, fisher_fdr), ~signif(.x, 3))) %>%
  datatable(rownames = FALSE,
            caption = paste0("Fisher's exact test — CD45-neg specific genes (n = ", n_tested, ")"),
            options = list(pageLength = 15, scrollX = TRUE)) %>%
  DT::formatStyle("fisher_sig",
    backgroundColor = DT::styleEqual(c(TRUE, FALSE), c("#e8f0fb", "white")))

# ─────────────────────────────────────────────────────────────────────────────
# TEST 2: Hypergeometric test (gene-set level)
# ─────────────────────────────────────────────────────────────────────────────
cd45neg_universe  <- pooled_summary_s %>% filter(n_neg > 0)
n_universe        <- nrow(cd45neg_universe)
k_threshold       <- ceiling(MIN_CD45NEG_FREQ * n_cd45neg)
high_freq_in_univ <- sum(cd45neg_universe$n_neg >= k_threshold)
n_specific        <- sum(cd45neg_universe$cd45neg_specific)
specific_high     <- sum(cd45neg_universe$cd45neg_specific &
                           cd45neg_universe$n_neg >= k_threshold)

hyper_p <- phyper(
  q          = specific_high - 1,
  m          = high_freq_in_univ,
  n          = n_universe - high_freq_in_univ,
  k          = n_specific,
  lower.tail = FALSE
)

cat("=== TEST 2: Hypergeometric (CD45-neg universe, n =", n_universe, "genes) ===\n")
cat("High-frequency threshold (>=", k_threshold, "of", n_cd45neg, "CTs):", high_freq_in_univ, "genes\n")
cat("CD45-neg specific genes in universe:", n_specific, "\n")
cat("Of those, high-frequency:", specific_high, "\n")
cat("Hypergeometric p-value:", signif(hyper_p, 3), "\n")

# ─────────────────────────────────────────────────────────────────────────────
# TEST 3: Permutation test
# ─────────────────────────────────────────────────────────────────────────────
gene_ct_mat <- gene_ct_long %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols     = gene,
    names_from  = cluster,
    values_from = present,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("gene") %>%
  as.matrix()

true_labels <- ifelse(colnames(gene_ct_mat) %in% cd45_neg, "CD45-", "CD45+")

obs_diff <- rowMeans(gene_ct_mat[, true_labels == "CD45-", drop = FALSE]) -
  rowMeans(gene_ct_mat[, true_labels == "CD45+", drop = FALSE])

set.seed(42)
perm_diffs <- replicate(N_PERMUTATIONS, {
  perm_labels <- sample(true_labels)
  rowMeans(gene_ct_mat[, perm_labels == "CD45-", drop = FALSE]) -
    rowMeans(gene_ct_mat[, perm_labels == "CD45+", drop = FALSE])
})

perm_p <- rowMeans(perm_diffs >= obs_diff)

perm_results <- tibble(
  gene          = names(obs_diff),
  obs_freq_diff = obs_diff,
  perm_p        = perm_p
)

specific_genes <- pooled_summary_s %>% filter(cd45neg_specific) %>% pull(gene)
perm_results <- perm_results %>%
  mutate(perm_fdr = NA_real_)
perm_results$perm_fdr[perm_results$gene %in% specific_genes] <-
  p.adjust(perm_results$perm_p[perm_results$gene %in% specific_genes], method = "BH")

cat("\n=== TEST 3: Permutation test (", N_PERMUTATIONS, "permutations) ===\n")
perm_results %>%
  inner_join(pooled_summary_s %>% filter(cd45neg_specific) %>%
               select(gene, freq_neg, freq_pos), by = "gene") %>%
  arrange(perm_fdr) %>%
  mutate(across(c(freq_neg, freq_pos, obs_freq_diff), ~round(.x, 2)),
         across(c(perm_p, perm_fdr), ~signif(.x, 3))) %>%
  datatable(rownames = FALSE,
            caption = paste0("Permutation test — CD45-neg specific genes (", N_PERMUTATIONS, " permutations)"),
            options = list(pageLength = 15, scrollX = TRUE))

# ─────────────────────────────────────────────────────────────────────────────
# Combine all results & save
# ─────────────────────────────────────────────────────────────────────────────
final_stats <- pooled_summary_s %>%
  left_join(fisher_results %>% select(gene, odds_ratio, fisher_p, fisher_fdr, fisher_sig),
            by = "gene") %>%
  left_join(perm_results   %>% select(gene, obs_freq_diff, perm_p, perm_fdr),
            by = "gene") %>%
  arrange(desc(cd45neg_specific), fisher_fdr)

write.csv(final_stats, OUT_STATS_CSV, row.names = FALSE)

# ─────────────────────────────────────────────────────────────────────────────
# Volcano plot
# ─────────────────────────────────────────────────────────────────────────────
plot_data <- final_stats %>%
  filter(n_neg > 0) %>%
  mutate(
    log10_fdr = -log10(fisher_fdr + 1e-10),
    sig_label = case_when(
      cd45neg_specific & fisher_fdr < 0.05 ~ "CD45-neg specific (FDR<0.05)",
      cd45neg_specific                     ~ "CD45-neg specific (ns)",
      shared                               ~ "Shared",
      TRUE                                 ~ "Other"
    )
  )

label_data <- plot_data %>%
  filter(cd45neg_specific | fisher_fdr < 0.05 | (shared & freq_neg > 0.6))

p_volcano <- ggplot(plot_data,
                    aes(x = obs_freq_diff, y = log10_fdr, colour = sig_label)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             colour = "grey50", linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "solid", colour = "grey85") +
  geom_text_repel(
    data = label_data, aes(label = gene),
    size = 3, max.overlaps = 30,
    segment.size = 0.3, segment.alpha = 0.5
  ) +
  scale_colour_manual(values = c(
    "CD45-neg specific (FDR<0.05)" = BLUE,
    "CD45-neg specific (ns)"       = "#A8DADC",
    "Shared"                       = PURP,
    "Other"                        = "grey70"
  )) +
  annotate("text",
           x = max(plot_data$obs_freq_diff, na.rm = TRUE) * 0.98,
           y = -log10(0.05) + 0.15,
           label = "FDR = 0.05", hjust = 1, size = 3, colour = "grey50") +
  labs(
    title    = "CD45-neg specificity of IFN leading-edge genes",
    subtitle = paste0(
      "x-axis: freq(CD45-neg) - freq(CD45-pos)  |  ",
      "y-axis: -log10(Fisher FDR)  |  ",
      "FDR correction over CD45-neg specific genes only (n = ", n_tested, ")"
    ),
    x      = "Frequency difference (CD45-neg minus CD45-pos)",
    y      = "-log10(Fisher FDR)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 8, colour = "grey40"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(clip = "off")

print(p_volcano)
```

---

## Gene × Cell Type Heatmap (Pooled, All Pathways)

Binary presence/absence of all leading-edge IFN genes across cell types (pooled union). Rows = cell types (red = CD45<sup>+</sup>, blue = CD45<sup>-</sup>); columns = genes coloured by category. Dark blue = gene is a leading-edge suppressed gene in that cell type.

```{r heatmap, fig.width=20, fig.height=8, out.width="100%"}
all_heat_genes <- pooled_all %>%
  unnest(genes) %>%
  distinct(cluster, genes) %>%
  pull(genes) %>%
  unique()

heat_mat <- pooled_all %>%
  unnest(genes) %>%
  distinct(cluster, genes) %>%
  mutate(present = 1) %>%
  pivot_wider(names_from = genes, values_from = present, values_fill = 0) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix()

gene_cat <- pooled_summary %>%
  select(genes, label) %>%
  filter(genes %in% colnames(heat_mat)) %>%
  tibble::column_to_rownames("genes")

col_ann <- data.frame(
  Category = gene_cat[colnames(heat_mat), "label"],
  row.names = colnames(heat_mat))

row_ann <- data.frame(
  Compartment = ifelse(rownames(heat_mat) %in% cd45_pos, "CD45+", "CD45-"),
  row.names = rownames(heat_mat))

ann_colours <- list(
  Compartment = c("CD45+" = RED,    "CD45-"        = BLUE),
  Category    = c(
    "CD45- specific" = BLUE,
    "CD45+ specific" = RED,
    "Shared"         = PURP,
    "Low frequency"  = "grey80"
  ))

n_genes    <- ncol(heat_mat)
n_celltypes <- nrow(heat_mat)
fig_w <- max(16, n_genes * 0.18 + 4)
fig_h <- max(6,  n_celltypes * 0.45 + 3)
options(repr.plot.width = fig_w, repr.plot.height = fig_h)

pheatmap(heat_mat,
         color              = c("white", "#2c7bb6"),
         annotation_row     = row_ann,
         annotation_col     = col_ann,
         annotation_colors  = ann_colours,
         cluster_rows       = TRUE,
         cluster_cols       = TRUE,
         fontsize_row       = 9,
         fontsize_col       = max(4, min(7, 500 / n_genes)),
         main               = sprintf(
           "ALL IFN Leading-Edge Genes × Cell Types (Pooled, %d genes × %d cell types)",
           n_genes, n_celltypes
         ),
         border_color       = "grey90",
         legend_breaks      = c(0, 1),
         legend_labels      = c("Absent", "Present"),
         angle_col          = 45,
         treeheight_row     = 35,
         treeheight_col     = 35,
         annotation_names_col = TRUE,
         annotation_names_row = TRUE)
```

---

```{r pathway_sections, results='asis', fig.width=18, fig.height=11}
for (pw in pathways) {
  sets     <- get_sets(pw)
  n_pos_ct <- nrow(ip_filt %>% filter(Description == pw, cd45 == "CD45+"))
  n_neg_ct <- nrow(ip_filt %>% filter(Description == pw, cd45 == "CD45-"))

  cat(sprintf("\n## %s {.tabset}\n\n", pw))
  cat(sprintf(
    paste0("Enriched (FDR < 0.05) in **%d CD45+** and **%d CD45-** ",
           "cell types. Leading-edge genes: **%d shared**, ",
           "**%d CD45+ only**, **%d CD45- only**.\n\n"),
    n_pos_ct, n_neg_ct,
    length(sets$shared), length(sets$cd45pos_only), length(sets$cd45neg_only)
  ))

  # ── UpSet: CD45+ vs CD45- pooled ────────────────────────────────────────
  cat("### Pooled: CD45+ vs CD45-\n\n")
  if (length(sets$cd45pos_genes) > 0 && length(sets$cd45neg_genes) > 0) {
    draw_safe(build_upset(
      list("CD45+" = sets$cd45pos_genes, "CD45-" = sets$cd45neg_genes),
      paste0(pw, "\nCD45+ vs CD45- — pooled leading-edge genes"),
      PURP, c("CD45+" = RED, "CD45-" = BLUE)))
  } else {
    cat("*Insufficient data.*\n\n")
  }

  # ── UpSet: CD45+ cell types ─────────────────────────────────────────────
  cat("### CD45+ cell types\n\n")
  sub_pos <- ip_filt %>% filter(Description == pw, cd45 == "CD45+")
  if (nrow(sub_pos) >= 2) {
    gl  <- setNames(sub_pos$genes, sub_pos$cluster)
    cls <- setNames(ct_palette[seq_along(gl)], names(gl))
    draw_safe(build_upset(gl,
      paste0(pw, "\nCD45+ cell types — leading-edge gene overlap"),
      RED, cls))
  } else {
    cat("*Fewer than 2 CD45+ cell types pass FDR < 0.05.*\n\n")
  }

  # ── UpSet: CD45- cell types ─────────────────────────────────────────────
  cat("### CD45- cell types\n\n")
  sub_neg <- ip_filt %>% filter(Description == pw, cd45 == "CD45-")
  if (nrow(sub_neg) >= 2) {
    gl  <- setNames(sub_neg$genes, sub_neg$cluster)
    cls <- setNames(ct_palette[seq_along(gl)], names(gl))
    draw_safe(build_upset(gl,
      paste0(pw, "\nCD45- cell types — leading-edge gene overlap"),
      BLUE, cls))
  } else {
    cat("*Fewer than 2 CD45- cell types pass FDR < 0.05.*\n\n")
  }

  cat("\n---\n\n")
}
```

## Render command

```{r render_hint, echo=TRUE, eval=FALSE}
# quarto render interferon_upset_plots.qmd
```
