---
title: "Leading-edge IFNγ and TNF Module Scores in Enterocyte Populations Pre- and Post-Ruxolitinib"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(stringr)
  library(readr)
  library(DT)
  library(plotly)
  library(tibble)
  library(purrr)
  library(forcats)
  library(glue)
  library(htmltools)
})

theme_set(theme_minimal(base_size = 11))
set.seed(42)
```

# Overview

This report addresses the question:

<span style="background-color: yellow;"> Can we use the leading-edge genes from significant IFNγ- and TNF-related GSEA pathways in enterocyte populations to generate IFNγ and TNF module scores before and after ruxolitinib treatment as an additional output of the dataset?</span>

This report is designed to provide a donor-level summary that complements the GSEA results.

::: {.callout-note collapse="true"}
Two key points are important for interpretation:

1.  **GSEA and module scores are not the same statistic.** GSEA detects coordinated rank shifts across pathway genes, while module scores summarize the average behavior of a selected gene set in cells.

2.  **IFNγ and TNF may behave differently.** IFNγ-related genes often move coherently and are therefore well captured by module scores. TNF-related pathways can contain mixed regulatory and feedback genes, so significant GSEA results may not always translate into equally strong module-score separation.
:::

# Inputs

```{r}
SEURAT_PATH   <- "./object_final_annotated_panGI_withDay.rds"
GSEA_PATH     <- "./PanGI_L2__ALL_clusters_GSEA.csv"

COL_CONDITION <- "Treatment"
PRE_LABEL     <- "Untreated"
POST_LABEL    <- "Treated"
COL_CELLTYPE  <- "PanGI_L2"
COL_DONOR     <- "subject"

ENTEROCYTE_CLUSTERS <- c(
  "Absorptive",
  "Transit_amplifying",
  "Stem",
  "Cycling_epithelia"
)

MIN_CELLS_PSEUDOBULK <- 10

IFNG_PATTERN <- "INTERFERON.GAMMA|IFN.GAMMA|IFNG"
TNF_PATTERN  <- "^TNF|TUMOR_NECROSIS|TNFR|TNF_SIGNAL|TNFA"
```

# Load data

```{r}
seu <- readRDS(SEURAT_PATH)
DefaultAssay(seu) <- "RNA"

gsea <- readr::read_csv(GSEA_PATH, show_col_types = FALSE)

stopifnot(
  COL_CONDITION %in% colnames(seu@meta.data),
  COL_CELLTYPE  %in% colnames(seu@meta.data),
  COL_DONOR     %in% colnames(seu@meta.data)
)
```

::: {.callout-note collapse="true"}
# Dataset summary

```{r}
dataset_summary <- tibble(
  Metric = c(
    "Total cells",
    "Unique donors",
    "Unique cell types",
    "Conditions"
  ),
  Value = c(
    format(ncol(seu), big.mark = ","),
    dplyr::n_distinct(seu@meta.data[[COL_DONOR]]),
    dplyr::n_distinct(seu@meta.data[[COL_CELLTYPE]]),
    paste(sort(unique(seu@meta.data[[COL_CONDITION]])), collapse = ", ")
  )
)

DT::datatable(
  dataset_summary,
  rownames = FALSE,
  options = list(dom = "t", paging = FALSE)
)
```
:::

# Enterocyte focus

The analysis is restricted to these enterocyte populations:

```{r}
enterocyte_tbl <- tibble(Enterocyte_cluster = ENTEROCYTE_CLUSTERS)

DT::datatable(
  enterocyte_tbl,
  rownames = FALSE,
  options = list(dom = "t", paging = FALSE)
)
```

# Significant GSEA pathways with leading-edge genes

```{r}
gsea_sig <- gsea %>%
  filter(
    p.adjust < 0.05,
    !is.na(leading_edge_symbols),
    leading_edge_symbols != ""
  ) %>%
  mutate(
    genes = str_split(leading_edge_symbols, "/"),
    Direction = ifelse(NES < 0, "Negative NES", "Positive NES")
  )
```

::: {.callout-note collapse="true"}
## IFNγ-related significant pathways

```{r}
gsea_ifng <- gsea_sig %>%
  filter(str_detect(toupper(Description), IFNG_PATTERN)) %>%
  arrange(cluster, p.adjust)

DT::datatable(
  gsea_ifng %>%
    select(cluster, Description, NES, p.adjust, setSize, Direction, leading_edge_symbols),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 10,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```
:::

::: {.callout-note collapse="true"}
## TNF-related significant pathways

```{r}
gsea_tnf <- gsea_sig %>%
  filter(str_detect(toupper(Description), TNF_PATTERN)) %>%
  arrange(cluster, p.adjust)

DT::datatable(
  gsea_tnf %>%
    select(cluster, Description, NES, p.adjust, setSize, Direction, leading_edge_symbols),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 10,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```
:::

# Enterocyte-restricted pathway sets

```{r}
gsea_ifng_ent <- gsea_ifng %>%
  filter(cluster %in% ENTEROCYTE_CLUSTERS)

gsea_tnf_ent <- gsea_tnf %>%
  filter(cluster %in% ENTEROCYTE_CLUSTERS)

ifng_down <- gsea_ifng_ent %>% filter(NES < 0)
tnf_down  <- gsea_tnf_ent  %>% filter(NES < 0)
```

::: {.callout-note collapse="true"}
## Enterocyte IFNγ pathways with negative NES

```{r}
DT::datatable(
  ifng_down %>%
    select(cluster, Description, NES, p.adjust, setSize),
  rownames = FALSE,
  filter = "top",
  options = list(scrollX = TRUE, pageLength = 10)
)
```
:::


::: {.callout-note collapse="true"}
## Enterocyte TNF pathways with negative NES

```{r}
DT::datatable(
  tnf_down %>%
    select(cluster, Description, NES, p.adjust, setSize),
  rownames = FALSE,
  filter = "top",
  options = list(scrollX = TRUE, pageLength = 10)
)
```
:::

::: {.callout-note collapse="true"}
# Leading-edge gene extraction

```{r}
extract_pathway_genes <- function(df, seu) {
  if (nrow(df) == 0) return(tibble())
  df %>%
    group_by(Description) %>%
    summarise(
      n_clusters      = n(),
      clusters        = paste(sort(unique(cluster)), collapse = ", "),
      mean_NES        = mean(NES, na.rm = TRUE),
      min_padj        = min(p.adjust, na.rm = TRUE),
      genes_union     = list(unique(unlist(genes))),
      n_genes_union   = length(unique(unlist(genes))),
      .groups         = "drop"
    ) %>%
    mutate(
      genes_present   = lapply(genes_union, function(g) intersect(g, rownames(seu))),
      n_genes_present = lengths(genes_present)
    ) %>%
    filter(n_genes_present >= 3) %>%
    arrange(mean_NES)
}

ifng_down_pathways <- extract_pathway_genes(ifng_down, seu)
tnf_down_pathways  <- extract_pathway_genes(tnf_down, seu)
```
:::

::: {.callout-note collapse="true"}
## IFNγ leading-edge pathway gene sets

```{r}
DT::datatable(
  ifng_down_pathways %>%
    mutate(genes_present = vapply(genes_present, paste, collapse = ", ", FUN.VALUE = character(1))) %>%
    select(Description, n_clusters, clusters, mean_NES, min_padj, n_genes_present, genes_present),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 10,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```

## TNF leading-edge pathway gene sets

```{r}
DT::datatable(
  tnf_down_pathways %>%
    mutate(genes_present = vapply(genes_present, paste, collapse = ", ", FUN.VALUE = character(1))) %>%
    select(Description, n_clusters, clusters, mean_NES, min_padj, n_genes_present, genes_present),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 10,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```

# Combined leading-edge gene sets

To provide a concise donor-level summary, the report also generates one combined IFNγ leading-edge set and one combined TNF leading-edge set.

```{r}
get_combined <- function(pathway_df, seu) {
  if (nrow(pathway_df) == 0) return(character(0))
  intersect(unique(unlist(pathway_df$genes_present)), rownames(seu))
}

combined_ifng <- get_combined(ifng_down_pathways, seu)
combined_tnf  <- get_combined(tnf_down_pathways, seu)

combined_gene_tbl <- tibble(
  Group = c("Combined IFNγ", "Combined TNF"),
  n_genes = c(length(combined_ifng), length(combined_tnf)),
  genes = c(
    paste(combined_ifng, collapse = ", "),
    paste(combined_tnf, collapse = ", ")
  )
)

DT::datatable(
  combined_gene_tbl,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 5,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```

# Subset to enterocyte cells before module scoring

This reduces memory use and ensures scoring is restricted to the biologically relevant subset.

```{r}
cells_keep <- rownames(seu@meta.data)[
  seu@meta.data[[COL_CELLTYPE]] %in% ENTEROCYTE_CLUSTERS &
    seu@meta.data[[COL_CONDITION]] %in% c(PRE_LABEL, POST_LABEL)
]

if (length(cells_keep) == 0) {
  stop("No enterocyte cells found in the selected conditions.")
}

seu_ent <- subset(seu, cells = cells_keep)
rm(seu)
gc()
```

## Enterocyte subset summary

```{r}
enterocyte_summary <- seu_ent@meta.data %>%
  count(
    cell_type = .data[[COL_CELLTYPE]],
    condition = .data[[COL_CONDITION]]
  ) %>%
  arrange(cell_type, condition)

DT::datatable(
  enterocyte_summary,
  rownames = FALSE,
  filter = "top",
  options = list(scrollX = TRUE, pageLength = 10)
)
```
:::

# Build module-score gene sets

This report computes:

-   pathway-specific IFNγ leading-edge scores
-   pathway-specific TNF leading-edge scores
-   one combined IFNγ score
-   one combined TNF score

```{r}
clean_name <- function(x) {
  x %>%
    str_replace_all("[^A-Za-z0-9]", "_") %>%
    str_replace_all("_+", "_") %>%
    str_trunc(45, ellipsis = "")
}

feature_list <- list()

if (nrow(ifng_down_pathways) > 0) {
  tmp_ifng <- setNames(
    ifng_down_pathways$genes_present,
    paste0(clean_name(ifng_down_pathways$Description), "_IFNg")
  )
  feature_list <- c(feature_list, tmp_ifng)
}

if (nrow(tnf_down_pathways) > 0) {
  tmp_tnf <- setNames(
    tnf_down_pathways$genes_present,
    paste0(clean_name(tnf_down_pathways$Description), "_TNF")
  )
  feature_list <- c(feature_list, tmp_tnf)
}

if (length(combined_ifng) >= 3) {
  feature_list <- c(feature_list, list(COMBINED_IFNg = combined_ifng))
}

if (length(combined_tnf) >= 3) {
  feature_list <- c(feature_list, list(COMBINED_TNF = combined_tnf))
}

if (length(feature_list) == 0) {
  stop("No valid leading-edge feature lists were generated.")
}

feature_list_summary <- tibble(
  score_name = names(feature_list),
  n_genes = lengths(feature_list),
  genes = vapply(feature_list, paste, collapse = ", ", FUN.VALUE = character(1))
)

DT::datatable(
  feature_list_summary,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 10,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```

::: {.callout-note collapse="true"}
# Compute module scores

```{r}
seu_ent <- AddModuleScore(
  seu_ent,
  features = feature_list,
  name = "LEMOD_",
  seed = 42
)

raw_cols <- paste0("LEMOD_", seq_along(feature_list))
new_cols <- paste0("score_", names(feature_list))

for (i in seq_along(raw_cols)) {
  seu_ent@meta.data[[new_cols[i]]] <- seu_ent@meta.data[[raw_cols[i]]]
}

score_lookup <- tibble(
  raw_col = raw_cols,
  score_col = new_cols
)

DT::datatable(
  score_lookup,
  rownames = FALSE,
  options = list(dom = "t", paging = FALSE, scrollX = TRUE)
)
```
:::

::: {.callout-note collapse="true"}
# Build donor-level paired summaries

We average module scores per donor, condition, and enterocyte subtype, then perform paired donor comparisons.

```{r}
cell_score_df <- seu_ent@meta.data %>%
  rownames_to_column("cell") %>%
  select(
    cell,
    donor = all_of(COL_DONOR),
    condition = all_of(COL_CONDITION),
    cell_type = all_of(COL_CELLTYPE),
    all_of(new_cols)
  ) %>%
  mutate(condition = factor(condition, levels = c(PRE_LABEL, POST_LABEL))) %>%
  pivot_longer(
    cols = all_of(new_cols),
    names_to = "score_col",
    values_to = "score"
  ) %>%
  mutate(
    group = case_when(
      str_detect(score_col, "IFNg") ~ "IFNg",
      str_detect(score_col, "TNF")  ~ "TNF",
      TRUE ~ "Other"
    ),
    pathway_label = score_col %>%
      str_replace("^score_", "") %>%
      str_replace_all("_", " ")
  )

donor_means <- cell_score_df %>%
  group_by(score_col, group, pathway_label, cell_type, donor, condition) %>%
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    n_cells = n(),
    .groups = "drop"
  ) %>%
  filter(n_cells >= MIN_CELLS_PSEUDOBULK) %>%
  group_by(score_col, cell_type, donor) %>%
  filter(n_distinct(condition) == 2) %>%
  ungroup()
```
:::

::: {.callout-note collapse="true"}
# Paired donor statistics

```{r}
paired_stats <- donor_means %>%
  group_by(score_col, group, pathway_label, cell_type) %>%
  group_modify(function(data, keys) {
    wide <- data %>%
      select(donor, condition, mean_score) %>%
      pivot_wider(names_from = condition, values_from = mean_score) %>%
      drop_na()

    if (nrow(wide) < 2) {
      return(tibble(
        n_pairs = nrow(wide),
        mean_pre = NA_real_,
        mean_post = NA_real_,
        delta = NA_real_,
        sd_delta = NA_real_,
        t_stat = NA_real_,
        p_value = NA_real_
      ))
    }

    diffs <- wide[[POST_LABEL]] - wide[[PRE_LABEL]]
    tt <- t.test(wide[[POST_LABEL]], wide[[PRE_LABEL]], paired = TRUE)

    tibble(
      n_pairs = nrow(wide),
      mean_pre = mean(wide[[PRE_LABEL]], na.rm = TRUE),
      mean_post = mean(wide[[POST_LABEL]], na.rm = TRUE),
      delta = mean(diffs, na.rm = TRUE),
      sd_delta = sd(diffs, na.rm = TRUE),
      t_stat = as.numeric(tt$statistic),
      p_value = tt$p.value
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    sig = case_when(
      is.na(p_adj) ~ "nd",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      p_adj < 0.1 ~ "trend",
      TRUE ~ "ns"
    )
  ) %>%
  arrange(group, pathway_label, cell_type, p_adj)
```
:::

::: {.callout-note collapse="true"}
## Searchable paired statistics table

```{r}
DT::datatable(
  paired_stats,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 20,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```

# Compact summary of statistically strongest paired results

```{r}
top_results <- paired_stats %>%
  filter(!is.na(p_adj)) %>%
  arrange(p_adj, desc(abs(delta))) %>%
  mutate(
    interpretation = case_when(
      delta < 0 ~ "Lower after treatment",
      delta > 0 ~ "Higher after treatment",
      TRUE ~ "No change"
    )
  )

DT::datatable(
  top_results %>%
    select(group, pathway_label, cell_type, n_pairs, mean_pre, mean_post, delta, p_value, p_adj, sig, interpretation),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = 15,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)
```
:::

# Donor trajectory plots

Each line represents one donor before and after treatment. Black pointranges show the mean ± SE across donors.

## IFNγ pathway trajectories

```{r}
ifng_cols <- new_cols[str_detect(new_cols, "IFNg")]

traj_ifng <- donor_means %>%
  filter(score_col %in% ifng_cols) %>%
  mutate(
    pathway = score_col %>%
      str_replace("^score_", "") %>%
      str_replace_all("_", " ") %>%
      str_wrap(32),
    hover_text = paste0(
      "<b>Donor:</b> ", donor,
      "<br><b>Cell type:</b> ", cell_type,
      "<br><b>Pathway:</b> ", pathway,
      "<br><b>Condition:</b> ", condition,
      "<br><b>Mean score:</b> ", round(mean_score, 4),
      "<br><b>Cells contributing:</b> ", n_cells
    )
  )

if (nrow(traj_ifng) > 0) {
  p_ifng <- ggplot(
    traj_ifng,
    aes(x = condition, y = mean_score, group = donor, color = donor, text = hover_text)
  ) +
    geom_line(linewidth = 0.9, alpha = 0.85) +
    geom_point(size = 3) +
    stat_summary(
      aes(group = 1),
      fun.data = mean_se,
      geom = "pointrange",
      color = "black",
      linewidth = 1
    ) +
    facet_grid(pathway ~ cell_type, scales = "free_y") +
    labs(
      title = "IFNγ leading-edge pathway trajectories",
      subtitle = "Donor-level mean module scores before and after ruxolitinib",
      x = NULL,
      y = "Mean module score"
    ) +
    theme(
      strip.text = element_text(face = "bold", size = 8),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  plotly::ggplotly(p_ifng, tooltip = "text")
} else {
  HTML("<p><em>No IFNγ trajectory data available after filtering.</em></p>")
}
```

## TNF pathway trajectories

```{r}
tnf_cols <- new_cols[str_detect(new_cols, "TNF")]

traj_tnf <- donor_means %>%
  filter(score_col %in% tnf_cols) %>%
  mutate(
    pathway = score_col %>%
      str_replace("^score_", "") %>%
      str_replace_all("_", " ") %>%
      str_wrap(32),
    hover_text = paste0(
      "<b>Donor:</b> ", donor,
      "<br><b>Cell type:</b> ", cell_type,
      "<br><b>Pathway:</b> ", pathway,
      "<br><b>Condition:</b> ", condition,
      "<br><b>Mean score:</b> ", round(mean_score, 4),
      "<br><b>Cells contributing:</b> ", n_cells
    )
  )

if (nrow(traj_tnf) > 0) {
  p_tnf <- ggplot(
    traj_tnf,
    aes(x = condition, y = mean_score, group = donor, color = donor, text = hover_text)
  ) +
    geom_line(linewidth = 0.9, alpha = 0.85) +
    geom_point(size = 3) +
    stat_summary(
      aes(group = 1),
      fun.data = mean_se,
      geom = "pointrange",
      color = "black",
      linewidth = 1
    ) +
    facet_grid(pathway ~ cell_type, scales = "free_y") +
    labs(
      title = "TNF leading-edge pathway trajectories",
      subtitle = "Donor-level mean module scores before and after ruxolitinib",
      x = NULL,
      y = "Mean module score"
    ) +
    theme(
      strip.text = element_text(face = "bold", size = 8),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  plotly::ggplotly(p_tnf, tooltip = "text")
} else {
  HTML("<p><em>No TNF trajectory data available after filtering.</em></p>")
}
```

# Combined IFNγ and TNF donor trajectories

These provide the clearest single-score donor-level summaries for each signaling program.

```{r}
combined_cols <- new_cols[str_detect(new_cols, "COMBINED")]

combined_df <- donor_means %>%
  filter(score_col %in% combined_cols) %>%
  mutate(
    pathway = score_col %>%
      str_replace("^score_", "") %>%
      str_replace_all("_", " "),
    hover_text = paste0(
      "<b>Donor:</b> ", donor,
      "<br><b>Cell type:</b> ", cell_type,
      "<br><b>Score:</b> ", pathway,
      "<br><b>Condition:</b> ", condition,
      "<br><b>Mean score:</b> ", round(mean_score, 4)
    )
  )

if (nrow(combined_df) > 0) {
  p_combined <- ggplot(
    combined_df,
    aes(x = condition, y = mean_score, group = donor, color = donor, text = hover_text)
  ) +
    geom_line(alpha = 0.85, linewidth = 0.9) +
    geom_point(size = 3) +
    stat_summary(
      aes(group = 1),
      fun.data = mean_se,
      geom = "pointrange",
      color = "black",
      linewidth = 1.1
    ) +
    facet_grid(pathway ~ cell_type, scales = "free_y") +
    labs(
      title = "Combined leading-edge donor trajectories",
      subtitle = "Each line is one donor; black pointrange shows mean ± SE across donors",
      x = NULL,
      y = "Mean module score"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  plotly::ggplotly(p_combined, tooltip = "text")
} else {
  HTML("<p><em>No combined trajectory data available.</em></p>")
}
```

::: {.callout-note collapse="true"}
# Delta summary plots

These plots summarize the donor-level change:

\[ \Delta = \text{Treated} - \text{Untreated}\]

Negative values indicate lower scores after treatment.

## All pathway-specific delta scores

```{r}
delta_all_df <- paired_stats %>%
  filter(!is.na(delta)) %>%
  mutate(
    pathway = pathway_label %>% str_wrap(32),
    cell_type = factor(cell_type, levels = ENTEROCYTE_CLUSTERS),
    sig_class = case_when(
      p_adj < 0.05 & delta < 0 ~ "Significant decrease",
      p_adj < 0.05 & delta > 0 ~ "Significant increase",
      p_adj < 0.1 & delta < 0 ~ "Trend decrease",
      p_adj < 0.1 & delta > 0 ~ "Trend increase",
      delta < 0 ~ "Decrease (ns)",
      TRUE ~ "Increase (ns)"
    ),
    hover_text = paste0(
      "<b>Group:</b> ", group,
      "<br><b>Pathway:</b> ", pathway,
      "<br><b>Cell type:</b> ", cell_type,
      "<br><b>Delta:</b> ", round(delta, 4),
      "<br><b>Mean pre:</b> ", round(mean_pre, 4),
      "<br><b>Mean post:</b> ", round(mean_post, 4),
      "<br><b>Adjusted p:</b> ", signif(p_adj, 3),
      "<br><b>Pairs:</b> ", n_pairs
    )
  )

if (nrow(delta_all_df) > 0) {
  p_delta_all <- ggplot(
    delta_all_df,
    aes(x = cell_type, y = delta, fill = sig_class, text = hover_text)
  ) +
    geom_col() +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    facet_wrap(~ pathway, scales = "free_y", ncol = 2) +
    labs(
      title = "Pathway-specific donor-level delta scores",
      subtitle = "Delta = Treated - Untreated",
      x = NULL,
      y = "Delta module score",
      fill = NULL
    ) +
    theme(
      axis.text.x = element_text(angle = 40, hjust = 1),
      strip.text = element_text(face = "bold", size = 8),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  plotly::ggplotly(p_delta_all, tooltip = "text")
} else {
  HTML("<p><em>No pathway-specific delta data available.</em></p>")
}
```

## Combined IFNγ and TNF delta scores

```{r}
delta_combined_df <- paired_stats %>%
  filter(score_col %in% combined_cols, !is.na(delta)) %>%
  mutate(
    pathway = score_col %>% str_replace("^score_", "") %>% str_replace_all("_", " "),
    cell_type = factor(cell_type, levels = ENTEROCYTE_CLUSTERS),
    hover_text = paste0(
      "<b>Score:</b> ", pathway,
      "<br><b>Cell type:</b> ", cell_type,
      "<br><b>Delta:</b> ", round(delta, 4),
      "<br><b>Mean pre:</b> ", round(mean_pre, 4),
      "<br><b>Mean post:</b> ", round(mean_post, 4),
      "<br><b>Adjusted p:</b> ", signif(p_adj, 3),
      "<br><b>Pairs:</b> ", n_pairs
    )
  )

if (nrow(delta_combined_df) > 0) {
  p_delta_combined <- ggplot(
    delta_combined_df,
    aes(x = cell_type, y = delta, fill = sig, text = hover_text)
  ) +
    geom_col() +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    facet_wrap(~ pathway, scales = "free_y") +
    labs(
      title = "Combined leading-edge delta scores",
      subtitle = "Delta = Treated - Untreated donor mean module score",
      x = NULL,
      y = "Delta module score",
      fill = "Adjusted significance"
    ) +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  plotly::ggplotly(p_delta_combined, tooltip = "text")
} else {
  HTML("<p><em>No combined delta data available.</em></p>")
}
```
:::

# Interpretation

## IFNγ

If IFNγ leading-edge module scores decrease after treatment in enterocyte populations, that provides a useful donor-level summary consistent with IFNγ GSEA findings.

Because IFNγ pathways are often transcriptionally coherent, these module scores are usually easier to interpret directly.

## TNF

TNF leading-edge module scores should be interpreted more cautiously.

TNF-related GSEA signals can remain biologically meaningful even if donor-level module scores are weaker or not formally significant. This can happen because TNF pathways often include receptor signaling, feedback regulators, and mixed-direction genes, so average module scores may partially cancel coordinated rank-based enrichment that is still detected by GSEA.

Therefore, TNF module scores are best treated as an **additional descriptive output**, not as a replacement for the original GSEA result.
