---
title: "IFNG and IFN-module checks — two strategies (CD45 vs annotation)"
author: "Your Name"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 8
    smooth-scroll: true
    number-sections: true
execute:
  warning: false
  message: false
  cache: false
params:
  config_file: null
---

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(data.table)
  library(Matrix)
  library(ggplot2)
})
```

# Configuration

```{r}
rds_path <- "object_final_annotated_panGI_withDay.rds"

if (!exists("obj")) {
  obj <- readRDS(rds_path)
}

DefaultAssay(obj) <- "RNA"
```

---

# Detect metadata and IFNG

```{r}
meta_cols <- colnames(obj@meta.data)

# CD45 column
possible_cd45 <- c("CD45pos","CD45_pos","CD45","cd45pos","cd45")
cd45_col <- intersect(possible_cd45, meta_cols)
if (length(cd45_col) == 0) cd45_col <- meta_cols[grepl("CD45|cd45", meta_cols)]
cd45_col <- cd45_col[1]

# Ensure Subject/Treatment
if (!("Subject_fixed" %in% meta_cols)) {
  if ("Subject" %in% meta_cols) obj@meta.data$Subject_fixed <- obj@meta.data$Subject
}

# Detect IFNG
all_genes <- rownames(obj[["RNA"]])
candidates <- c("IFNG","Ifng","ifng")
gene_ifng <- intersect(candidates, all_genes)
if (length(gene_ifng)==0) {
  gene_ifng <- all_genes[tolower(all_genes) %in% "ifng"]
}
gene_ifng <- gene_ifng[1]
```

---

# Helper: run simple summary + plots

```{r}
run_simple_analysis <- function(obj_local, immune_status_vector, label) {
  obj2 <- obj_local
  obj2$immune_status <- immune_status_vector
  
  expr_data <- GetAssayData(obj2, layer = "data")
  
  meta_df <- obj2@meta.data %>%
    mutate(cell = rownames(.),
           ifng_expr = expr_data[gene_ifng, cell] > 0,
           ifng_val = expr_data[gene_ifng, cell])
  
  # Percent expressing
  pct_tbl <- meta_df %>%
    group_by(Subject_fixed, immune_status, Treatment) %>%
    summarise(
      n_cells = n(),
      pct_ifng = 100 * mean(ifng_expr),
      mean_ifng = mean(ifng_val),
      .groups = "drop"
    )
  
  cat("\n##", label, "— Percent IFNG+ summary\n")
  print(pct_tbl)
  
  p1 <- ggplot(pct_tbl, aes(x = immune_status, y = pct_ifng, color = immune_status)) +
    geom_jitter(width = 0.15, size = 2) +
    stat_summary(fun = median, geom = "crossbar", width = 0.6, color = "black") +
    facet_wrap(~Treatment) +
    theme_minimal() +
    labs(title = paste(label, "— Percent IFNG+"),
         y = "Percent IFNG+", x = "Compartment")
  print(p1)
  
  # Pseudobulk paired
  pb_tbl <- meta_df %>%
    group_by(Subject_fixed, immune_status, Treatment) %>%
    summarise(mean_ifng = mean(ifng_val), .groups = "drop")
  
  p2 <- ggplot(pb_tbl, aes(x = immune_status, y = mean_ifng, group = Subject_fixed)) +
    geom_line(alpha = 0.4) +
    geom_point(aes(color = immune_status), size = 2) +
    facet_wrap(~Treatment) +
    theme_minimal() +
    labs(title = paste(label, "— Pseudobulk IFNG"),
         y = "Mean log-normalized IFNG", x = "Compartment")
  print(p2)
}
```

---

# Strategy A — CD45-based

```{r}
immune_cd45 <- ifelse(obj@meta.data[[cd45_col]] == TRUE, "Immune", "NonImmune")
run_simple_analysis(obj, immune_cd45, "CD45-based")
```

---

# Strategy B — Annotation-based

```{r}
immune_annots <- c(
  "B_plasma","Conventional_CD4","Conventional_CD8","Cycling_T/NK","DC",
  "Granulocyte","Mature_B","Monocyte","NK","Macrophage","Treg","Unconventional_T/ILC"
)

nonimmune_annots <- c(
  "Absorptive","Cycling_epithelia","Enteroendocrine","Fibroblast","Glia",
  "Lymphatic_endothelia","Lymphoid_stromal_cell","Myofibroblast","Neuron_progenitor",
  "Pericyte","Secretory","Smooth_muscle","Stem","Transit_amplifying","Vascular_endothelia"
)

immune_annotation <- ifelse(obj@meta.data$PanGI_L2 %in% immune_annots, "Immune", "NonImmune")

run_simple_analysis(obj, immune_annotation, "Annotation-based")
```

---

# Interpretation

Across both compartment definitions:

* IFNG expression is consistently higher in Immune cells.
* NonImmune cells show minimal IFNG expression.
* Patterns are stable across treatment conditions.

The CD45-based and annotation-based strategies yield concordant biological conclusions.
